--- 
title: "INLA and inlabru with spatial patterns"
author: "Thierry Onkelinx"
site: bookdown::bookdown_site
output:
  bookdown::pdf_book:
    base_format: INBOmd::inbo_slides
    slide_level: 3
    theme: inboenglish
    flandersfont: TRUE
    cover: aerial-aerial-shot-aerial-view-753868.jpg # Photo by Stephan Müller from Pexels https://www.pexels.com/photo/herd-of-brown-and-white-cows-on-green-grass-field-753868/
    cover_horizontal: FALSE
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  dev = "cairo_pdf",
  dpi = 300
)
library(raster)
library(sf)
library(tidyverse)
library(inlabru)
library(INLA)
library(gstat)
# devtools::install_github("AckerDWM/gg3D")
library(gg3D)
library(INBOtheme)
if (interactive()) {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 12, 
      transparent = "plot"
    )
  )
} else {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 8, 
      transparent = "plot"
    )
  )
  update_geom_defaults("text", list(size = 3))
  update_geom_defaults("point", list(size = 0.5))
  update_geom_defaults("line", list(size = 0.2))
  update_geom_defaults("hline", list(size = 0.2))
  update_geom_defaults("vline", list(size = 0.2))
  update_geom_defaults("segment", list(size = 0.2))
}
set.seed(200190307)
```

# Checking spatial autocorrelation

## Pearson residuals

### Definition

> - components?
> - observed value ($y_i$), fitted value ($\hat{y}_i$), mean squared error ($MSE$)
> - formula
> - $$pr_i = \frac{y_i - \hat{y}_i}{\sqrt{MSE}}$$
> - $MSE$ (variance) depends on distribution! Check it using `inla.doc("name_of_your_distribution")`.

### Example data: rainfall in Parana state, Brazil

```{r rain-raw, echo = FALSE}
data(PRprec)
PRprec %>%
  transmute(Longitude, Latitude, Altitude, Rain = d0701) %>%
  filter(!is.na(Rain), Rain > 0) -> dataset
ggplot(dataset, aes(x = Longitude, y = Latitude, colour = Rain, size = Rain)) +
  geom_point() + coord_map()
```

```{r project, echo = FALSE}
st_as_sf(dataset, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = 5880) -> dataset
dataset %>%
  bind_cols(st_coordinates(dataset) %>%
  as.data.frame()) %>%
  mutate(Xc = X / 100e3 - 53, Yc = Y / 100e3 - 71) -> dataset
```

### Calculate Pearson residuals

```{r model-iid}
model_iid <- inla(Rain ~ Xc + Yc, family = "gamma", data = dataset, 
                  control.compute = list(waic = TRUE))
dataset %>%
  mutate(
    mu = model_iid$summary.fitted.values$mean,
    sigma2 = mu ^ 2 / model_iid$summary.hyperpar[1, "mean"],
    Pearson_iid = (Rain - mu) / sqrt(sigma2)
  ) -> dataset
```

### Challenge 1

- What is the mean for your model?
- What is the variance for your model? Hint: `inla.doc("your distribution")`
- Calculate the Pearson residuals for your model

## Variogram

### Definition

```{r vg-default}
vg_default <- variogram(Pearson_iid ~ 1, locations = ~X + Y, 
                        data = as.data.frame(dataset), cressie = TRUE)
```

```{r vg-default-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 170e3, nugget = 0.375) %>%
  fit.variogram(object = vg_default) -> vgm_default
vgl_default <- variogramLine(vgm_default, maxdist = max(vg_default$dist))
ggplot(vg_default, aes(x = dist / 1000, y = gamma)) +
  geom_line(data = vgl_default) +
  geom_point() +
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Important characteristics

```{r vg-default-char, echo = FALSE}
characteristics <- tribble(
  ~id, ~label, ~x, ~xend, ~y, ~yend,
  1, "range", 0, 3 * vgm_default$range[2] / 1000, 0.25, 0.25,
  3, "nugget", 0, 0, 0, vgm_default$psill[1],
  2, "sill", 10, 10, 0, sum(vgm_default$psill),
  4, "partial sill", 0, 0, vgm_default$psill[1], sum(vgm_default$psill)
)
ggplot(vg_default, aes(x = dist / 1000, y = gamma)) +
  geom_line(data = vgl_default) +
  geom_point() +
  geom_hline(yintercept = cumsum(vgm_default$psill), linetype = 2) +
  geom_vline(xintercept = 3 * vgm_default$range[2] / 1000, linetype = 3) +
  geom_segment(data = characteristics, arrow = arrow(ends = "both"),
               aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_label(data = characteristics,
             aes(x = (x + xend) / 2, y = (y + yend) / 2, label = id)) +
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Important characteristics

```{r vg-default-char-lab, echo = FALSE}
ggplot(vg_default, aes(x = dist / 1000, y = gamma)) +
  geom_line(data = vgl_default) +
  geom_point() +
  geom_hline(yintercept = cumsum(vgm_default$psill), linetype = 2) +
  geom_vline(xintercept = 3 * vgm_default$range[2] / 1000, linetype = 3) +
  geom_segment(data = characteristics, arrow = arrow(ends = "both"),
               aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_label(data = characteristics, hjust = 0,
             aes(x = (x + xend) / 2, y = (y + yend) / 2, label = label)) +
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Projected example data

```{r rainfall-projected, echo = FALSE}
ggplot(dataset) + geom_sf(aes(colour = Rain, size = Rain)) + 
  coord_sf(datum = st_crs(5880))
```

### Increased cutoff

```{r vg-large-cutoff}
vg_large <- variogram(Pearson_iid ~ 1, locations = ~X + Y, cressie = TRUE, 
                      data = as.data.frame(dataset), cutoff = 600e3)
```

```{r vg-large-cutoff-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 170e3, nugget = 0.375) %>%
  fit.variogram(object = vg_large) -> vgm_large
vgl_large <- variogramLine(vgm_large, maxdist = max(vg_large$dist))
ggplot(vg_large, aes(x = dist / 1000, y = gamma)) +
  geom_line(data = vgl_large) +
  geom_point() +
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Number of point pairs is important

```{r vg-large-np, echo = FALSE}
ggplot(vg_large, aes(x = dist / 1000, y = gamma)) +
  geom_line(data = vgl_large) + geom_point() + 
  geom_text(aes(label = np), hjust = -0.2, vjust = 1.2) + 
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Too small width leads to unstable variograms

```{r vg-small}
vg_small <- variogram(Pearson_iid ~ 1, locations = ~X + Y, cressie = TRUE, 
                      data = as.data.frame(dataset), width = 1e3)
```

```{r vg-small-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 170e3, nugget = 0.375) %>%
  fit.variogram(object = vg_small) -> vgm_small
vgl_small <- variogramLine(vgm_small, maxdist = max(vg_default$dist))
ggplot(vg_small, aes(x = dist / 1000, y = gamma)) +
  geom_line(data = vgl_small) +
  geom_point(aes(colour = cut(np, c(0, 10, 20, 50, 100, 200, 500, Inf)))) +
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA)) +
  labs(colour = "number of\npoint pairs")
```

### Sensible small width yields the most informative variogram

```{r vg-final}
vg_final <- variogram(Pearson_iid ~ 1, locations = ~X + Y, cressie = TRUE, 
                      data = as.data.frame(dataset), width = 10e3)
```

```{r vg-final-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 170e3, nugget = 0.375) %>%
  fit.variogram(object = vg_final) -> vgm_final
vgl_final <- variogramLine(vgm_final, maxdist = max(vg_default$dist))
ggplot(vg_final, aes(x = dist / 1000, y = gamma)) +
  geom_line(data = vgl_final) + geom_point() +
  geom_text(aes(label = np), hjust = -0.2, vjust = 1.2) + 
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Challenge 2

- What is the minimum binwidth for your data?
- Calculate the variogram for your model
- What is the approximate range of of the variogram?
- What is the nugget, sill and partial sill?

# Prepare the model

## Creating a mesh

### Size of a mesh {.allowframebreaks}

```{r volcano, echo = FALSE, warning = FALSE}
data(volcano)
vr <- raster(volcano)
aggregate(vr, 16) %>%
  as.data.frame(xy = TRUE) %>%
  ggplot(aes(x = x, y = y, z = layer)) +
  theme_void() +
  stat_wireframe()
aggregate(vr, 8) %>%
  as.data.frame(xy = TRUE) %>%
  ggplot(aes(x = x, y = y, z = layer)) +
  theme_void() +
  stat_wireframe()
aggregate(vr, 4) %>%
  as.data.frame(xy = TRUE) %>%
  ggplot(aes(x = x, y = y, z = layer)) +
  theme_void() +
  stat_wireframe()
aggregate(vr, 2) %>%
  as.data.frame(xy = TRUE) %>%
  ggplot(aes(x = x, y = y, z = layer)) +
  theme_void() +
  stat_wireframe()
vr %>%
  as.data.frame(xy = TRUE) %>%
  ggplot(aes(x = x, y = y, z = layer)) +
  theme_void() +
  stat_wireframe()
```

### Guidelines

- equilateral triangles work best
- edge length should be around a third to a tenth of the range
- avoid narrow triangles
- avoid small edges
- add extra, larger triangles around the border
- simplify the border

### Mesh only within the border

```{r dummy-border, echo = FALSE}
list(
  Polygons(
    list(
      Polygon(matrix(c(0, 0, 1, 1, 0, 1, 1, 0), ncol = 2)),
      Polygon(matrix(c(0.3, 0.3, 0.7, 0.7, 0.3, 0.7, 0.7, 0.3), ncol = 2), 
              hole = TRUE)
    ), 1
  ),
  Polygons(
    list(Polygon(matrix(c(2, 2, 3, 3, 0, 1, 1, 0), ncol = 2))), 2
  )
) %>%
  SpatialPolygons() -> border
```

```{r mesh-inside}
mesh <- inla.mesh.2d(boundary = border, max.edge = 0.15)
ggplot() + gg(mesh) + coord_fixed() + theme_map() + 
  ggtitle(paste("Vertices: ", mesh$n))
```

### Mesh going outside the border

```{r mesh-outside}
mesh <- inla.mesh.2d(boundary = border, max.edge = c(0.15, 0.3))
ggplot() + gg(mesh) + coord_fixed() + theme_map() + 
  ggtitle(paste("Vertices: ", mesh$n))
```

### Mesh for rainfall data

```{r project-boundary, echo = FALSE}
data(PRborder)
PRborder %>%
  Polygon() %>%
  list() %>%
  Polygons(ID = 1) %>%
  list() %>%
  SpatialPolygons(proj4string = CRS("+init=epsg:4326")) %>%
  spTransform(CRS("+init=epsg:5880")) -> boundary
```

```{r mesh-rainfall-1}
mesh <- inla.mesh.2d(boundary = boundary, max.edge = c(30e3, 100e3))
ggplot(dataset) + gg(mesh) + geom_sf() + ggtitle(paste("Vertices: ", mesh$n)) +
  coord_sf(datum = st_crs(5880))
```

### Use `cutoff` to simplify mesh

```{r mesh-rainfall-simple}
mesh1 <- inla.mesh.2d(boundary = boundary, max.edge = c(30e3, 100e3), 
                      cutoff = 10e3)
ggplot(dataset) + gg(mesh1) + geom_sf() + 
  ggtitle(paste("Vertices: ", mesh1$n)) + coord_sf(datum = st_crs(5880))
```

### Finer mesh for final model run

```{r mesh-rainfall-simple2}
mesh2 <- inla.mesh.2d(boundary = boundary, max.edge = c(10e3, 30e3), 
                      cutoff = 5e3)
ggplot(dataset) + gg(mesh2) + geom_sf() + 
  ggtitle(paste("Vertices: ", mesh2$n)) + coord_sf(datum = st_crs(5880))
```

### Challenge 3

- What are the relevant `max.edge` and `cutoff` for a course mesh?
- What are the relevant `max.edge` and `cutoff` for a smooth mesh?
- Create a course and a smooth mesh for your data

## Creating an SPDE model

### SPDE using penalised complexity priors

**S**tochastic **P**artial **D**ifferential **E**quations

- `prior.range = c(r, alpha_r)`: $P(\rho < r) < \alpha_r$
- `prior.sigma = c(s, alpha_s)`: $P(\sigma > s) < \alpha_s$

```{r spde1}
spde1 <- inla.spde2.pcmatern(mesh1, prior.range = c(100e3, 0.5), 
                             prior.sigma = c(0.9, 0.05))
spde2 <- inla.spde2.pcmatern(mesh2, prior.range = c(100e3, 0.5), 
                             prior.sigma = c(0.9, 0.05))
```

### Challenge 4

- What are relevant priors for the range and sigma for your data
    - Hint: see challenge 2
- Make the SPDE models for your data

# Fitting the model

## Only the data

### The stack for the observed data

```{r stacks}
A1 <- inla.spde.make.A(mesh = mesh1, loc = st_coordinates(dataset))
stack1 <- inla.stack(
  tag = "estimation", ## tag
  data = list(Rain = dataset$Rain), ## response
  A = list(A1, 1), ## projector matrices (SPDE and fixed effects)
  effects = list(
    list(site = seq_len(spde1$n.spde)), ## random field index
    dataset %>%
      as.data.frame() %>%
      transmute(Intercept = 1, Xc, Yc) ## fixed effect covariates
  )
)
```

### Model fit

#### INLA

```{r model-fit-INLA}
model_spde1 <- inla(Rain ~ 0 + Intercept + Xc + Yc + f(site, model = spde1), 
                    family = "gamma", data = inla.stack.data(stack1),
                    control.predictor = list(A = inla.stack.A(stack1)),
                    control.compute = list(waic = TRUE)
                  )
```

#### inlabru

```{r model-fit-inlabru0, eval = FALSE}
bru_spde1 <- bru(Rain ~ Xc + Yc + site(map = st_coordinates, model = spde1), 
                 family = "gamma", data = dataset)
```
```{r model-fit-inlabru}
bru_spde1 <- bru(Rain ~ Xc + Yc + site(map = coordinates, model = spde1), 
                 family = "gamma", data = as_Spatial(dataset))
```

### Comparison of fixed effect parameters

```{r comparison-fe, echo = FALSE}
model_spde1$summary.fixed %>%
  rownames_to_column("parameter") %>%
  mutate(model = "INLA") %>%
  bind_rows(
    bru_spde1$summary.fixed %>%
    rownames_to_column("parameter") %>%
    mutate(model = "inlabru")
  ) %>%
  ggplot(aes(x = model, y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_errorbar() + geom_point() + facet_wrap(~parameter, scales = "free_y")
```

### Comparing hyperparameters

```{r comparison-hp, echo = FALSE}
model_spde1$summary.hyperpar %>%
  rownames_to_column("parameter") %>%
  mutate(model = "INLA") %>%
  bind_rows(
    bru_spde1$summary.hyperpar %>%
    rownames_to_column("parameter") %>%
    mutate(model = "inlabru")
  ) %>%
  ggplot(aes(x = model, y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_errorbar() + geom_point() + facet_wrap(~parameter, scales = "free_y")
```

### Correlation structure

```{r corstruct}
spde.posterior(bru_spde1, "site", what = "matern.covariance") -> covplot
spde.posterior(bru_spde1, "site", what = "matern.correlation") -> corplot
multiplot(plot(covplot), plot(corplot))
```

### Calculate Pearson residuals

```{r vg-new, error = TRUE}
dataset %>%
  mutate(
    mu = model_spde1$summary.fitted.values$mean,
    sigma2 = mu ^ 2 / model_spde1$summary.hyperpar[1, "mean"],
    Pearson_iid = (Rain - mu) / sqrt(sigma2)
  ) -> dataset
```

### Using the stack index

```{r}
si <- inla.stack.index(stack1, "estimation")$data
dataset %>%
  mutate(
    mu = model_spde1$summary.fitted.values$mean[si],
    sigma2 = mu ^ 2 / model_spde1$summary.hyperpar[1, "mean"],
    Pearson_spde = (Rain - mu) / sqrt(sigma2)
  ) -> dataset
```

### Using inlabru

```{r}
fit <- predict(bru_spde1, as_Spatial(dataset), ~exp(Intercept + Xc + Yc + site))
dataset %>%
  mutate(
    mu = fit$mean, 
    sigma2 = mu ^ 2 / model_spde1$summary.hyperpar[1, "mean"],
    Pearson_spde = (Rain - mu) / sqrt(sigma2)
  ) -> dataset
```

### Variogram

```{r vg-fit}
vg_fit <- variogram(Pearson_spde ~ 1, cressie = TRUE, 
                    data = as_Spatial(dataset), width = 10e3)
```

```{r vg-fit-plot, echo = FALSE}
ggplot(vg_fit, aes(x = dist / 1000, y = gamma)) +
  geom_point() +
  scale_x_continuous("distance (km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Interpolate GMRF

```{r spde_eta}
A1.grid <- inla.mesh.projector(mesh1, dims = c(41, 41))
inla.mesh.project(A1.grid, model_spde1$summary.random$site) %>%
  as.matrix() %>%
  as.data.frame() %>%
  bind_cols(
    expand.grid(x = A1.grid$x, y = A1.grid$y)
  ) %>%
  filter(!is.na(ID)) -> eta_spde
```

### Plot GMRF

```{r spde-eta-plot}
ggplot(dataset) + geom_tile(data = eta_spde, aes(x = x, y = y, fill = mean)) +
  geom_sf() + scale_fill_gradient2()
```

## Predictions

### Prediction stack for SPDE grid + fixed effects

```{r stack_grid}
expand.grid(X = A1.grid$x, Y = A1.grid$y) %>%
  mutate(Intercept = 1, Xc = X / 1e5 - 53, Yc = Y / 1e5 - 71) -> grid_data
stack1_grid <- inla.stack(
  tag = "grid", ## tag
  data = list(Rain = NA), ## response
  A = list(A1.grid$proj$A, 1), ## projector matrices (SPDE and fixed effects)
  effects = list(
    list(site = seq_len(spde1$n.spde)), ## random field index
    grid_data  ## covariates at grid locations
  )
)
```

### Refit the model with the combinated stack

```{r fit_grid}
stack_all <- inla.stack(stack1, stack1_grid)
model_grid <- inla(Rain ~ 0 + Intercept + Xc + Yc + f(site, model = spde1), 
                   family = "gamma", data = inla.stack.data(stack_all),
                   control.predictor = list(A = inla.stack.A(stack_all), 
                                            link = 1),
                   control.compute = list(waic = TRUE),
                   control.mode = list(theta = model_spde1$mode$theta, 
                                       restart = FALSE),
                   control.results = list(return.marginals.random = FALSE,
                                          return.marginals.predictor = FALSE)
                   )
```

### Plot grid {.allowframebreaks}

```{r plot-grid}
si <- inla.stack.index(stack_all, "grid")$data
grid_data %>%
  bind_cols(model_grid$summary.fitted.values[si, ]) %>%
  `coordinates<-`(~X + Y) %>%
  `proj4string<-`(CRS("+init=epsg:5880")) -> gd
gd[!is.na(over(gd, boundary)), ] %>%
  as.data.frame() %>%
  ggplot() + geom_tile(aes(x = X, y = Y, fill = mean)) + coord_fixed()
```

### Using inlabru

```{r}
pred_mesh <- predict(bru_spde1, pixels(mesh1), ~exp(Intercept + Xc + Yc + site))
ggplot() + gg(pred_mesh) + gg(boundary)
```

### Challenge 5

- Fit the model using the SPDE
- Plot a map of the GMRF
- Plot a map of the predictions and their credible interval
