---
title: "Let's create an interactive map!"
description: "Creating interactive maps with leaflet"
author: "Jeroen Speybroeck"
date: 2018-06-14
categories: ["r"]
tags: ["gis", "r", "maps"]
output: 
    md_document:
        preserve_yaml: true
        variant: gfm
---

_General note: migration to the more actively developed `sf` package is currently advised by the `sp` maintainer._
_The `sp` package, used in this tutorial, is still maintained in order to support the newest versions of the GDAL and PROJ backends._

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The required packages are **leaflet** and **sp**.

```{r message=FALSE, warning=FALSE}
library(leaflet)
library(sp)
```

## Dummy data

Let's create a dumy `data.frame` to play around, i.e. the three locations of INBO:

```{r}
names <- c("VAC HT","Geraardsbergen","Linkebeek")
lat <- c(50.865664, 50.760201, 50.767950)
lon <- c(4.349944, 3.874300, 4.333044)
data <- data.frame(names,lat,lon)
```

We created three points:

```{r}
plot(data$lon, data$lat)
```

## Creating a map

We need to convert the `data.frame` to a `SpatialPointsDataFrame`:

```{r}
crs_wgs84 <- CRS(SRS_string = "EPSG:4326")
pts <- SpatialPointsDataFrame(data[c("lon","lat")], 
                              data[!(names(data) %in% c("lon","lat"))],
                              proj4string = crs_wgs84)

```

The `leaflet` package is ideal to create a basic interactive map:

```{r}
library(leaflet)
leaf_map <- leaflet(pts) %>% 
  addTiles(group = "OSM") %>%
  addCircleMarkers()
```

```{r eval=TRUE}
leaf_map
```


**Nice, no?!**

More information is provided [at the leaflet information website](https://rstudio.github.io/leaflet/)!
