---
title: "How to use the GeoTIFF raster file format in R"
description: "A simple tutorial to demonstrate the use of GeoTIFF files in R"
author: "Floris Vanderhaeghe"
date: 2019-11-20
categories: ["r", "gis"]
tags: ["gis", "r"]
output: 
    md_document:
        preserve_yaml: true
        variant: gfm
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE
)
set.seed(123456789)
unlink("art*")
```


This tutorial uses a few basic functions from the [dplyr](https://dplyr.tidyverse.org) and [raster](https://rspatial.org/raster/pkg) packages.
While only a few functions are used, you can use the previous hyperlinks to access the tutorials (vignettes) of these packages for more functions and information.

```{r warning=FALSE, message=FALSE}
options(stringsAsFactors = FALSE)
library(raster)
library(tidyverse)
library(inborutils)
```

You find bit more background about 'why and what', regarding the open GeoTIFF standard, in [a separate post](../../articles/geospatial_standards/) on this website.

## How to make and use GeoTIFF files (`*.tif`)

### Making a mono-layered GeoTIFF file from a `RasterLayer` R object

Let's create a small dummy `RasterLayer` object from scratch, for some area in Belgium (using the CRS ^[CRS = coordinate reference system] Belgian Lambert 72, i.e. [EPSG-code 31370](https://epsg.io/31370)):

```{r}
artwork <- 
  raster(extent(188500, 190350, 227550, 229550), # xmin, xmax, ymin, ymax
         res = 50, # resolution of 50 meters
         crs = CRS("+init=epsg:31370")) %>% 
  setValues(runif(ncell(.)))  # fill with random values
```

What does this look like?

```{r}
artwork
```

A simple trick to plot this raster:

```{r artwork}
spplot(artwork)
```

To write this `RasterLayer` object as a GeoTIFF, you can use the `raster::writeRaster()` function.
In the background, it uses the GeoTIFF driver of the powerful [GDAL](https://gdal.org) library.

```{r}
artwork %>% 
  writeRaster("artwork.tif")
```

And now?\
_**Say HURRAY!!**_

### Making a multi-layered GeoTIFF file from a `RasterBrick` R object

Let's create a `RasterBrick` object of three layers:

```{r}
arts <- brick(artwork) # RasterBrick with one layer (the RasterLayer from above)
arts[[2]] <- artwork + 10 # Add second layer, e.g. based on first one
arts[[3]] <- calc(arts[[2]], function(x) 20 ^ x) # Making third layer from second
names(arts) <- paste0("layer", 1:3)
```

_Note: in real-life cases, you will see that the `calc()` function is more efficient than simple algebraic expressions such as for layer2._

How does the result look like?

```{r}
arts
```

```{r arts}
arts %>% 
  as.list %>% 
  lapply(spplot)
```

So now what?\
_**Let's write it!**_

```{r}
arts %>% 
  writeRaster("arts.tif")
```

But, I want to add 20 extra layers!\
_**(...)**_

```{r}
arts2 <- 
  calc(artwork, 
       function(x) {-1:-20 * x}, # first layer = -1 * artwork
                                 # second layer = -2 * artwork
                                 # ....
       forceapply = TRUE)
names(arts2) <- paste0("neg_layer", 1:20)
# adding it to arts:
arts <- brick(list(arts, arts2))
# saving layer names for later use:
mynames <- names(arts)
```

```{r}
nlayers(arts)
```

```{r}
names(arts)
```


**Overwrite** the earlier written file:

```{r}
arts %>% 
  writeRaster("arts.tif",
              overwrite = TRUE)
```

That's about it!

### Reading a GeoTIFF file

Nothing can be more simple...

Reading a mono-layered GeoTIFF file with `raster()` gives back the `RasterLayer`:

```{r}
artwork_test <- raster("artwork.tif")
```

```{r}
artwork_test
```

Reading a multi-layered GeoTIFF file with `brick()` returns the `RasterBrick`:

```{r}
arts_test <- brick("arts.tif")
```

However:

```{r}
names(arts_test)
```

As you see, layer names are not saved in the GeoTIFF.
You define them in R:

```{r}
names(arts_test) <- mynames
arts_test
```
 
That's what we wanted!

The actual data are not loaded into memory, but read in chunks when performing operations.
This makes it convenient when using larger rasters:

```{r}
inMemory(arts_test)
```

Selecting a specific layer by its name:

```{r}
arts_test$neg_layer20
```


### Homework: explore the amazing `stars` package

Enter deep hyperspace and meet the `stars` package, which stores multidimensional hypercubes...
Really, visit its [website](https://r-spatial.github.io/stars/) and never look (or turn?) back!

```{r warning=FALSE, message=FALSE}
library(stars)
```


For now, my time's up and I'll just demonstrate how easy it is to transform a `Raster*` object into a `stars` object:

```{r paged.print=FALSE}
interstellar <- 
  arts[[1:5]] %>% 
  st_as_stars()
interstellar
```

It does make sense, right?

What about:

```{r paged.print=FALSE}
interstellar %>% split("band")
```


The `stars` package has a number of efficient geospatial algorithms that make it worth using, even for simple raster layers!


```{r include=FALSE}
unlink("art*")
```

