--- 
title: "INLA and inlabru with spatial patterns"
author: "Thierry Onkelinx"
site: bookdown::bookdown_site
output:
  bookdown::pdf_book:
    base_format: INBOmd::inbo_slides
    toc: FALSE
    slide_level: 3
    theme: inboenglish
    flandersfont: TRUE
    cover: aerial-aerial-shot-aerial-view-753868.jpg # Photo by Stephan Müller from Pexels https://www.pexels.com/photo/herd-of-brown-and-white-cows-on-green-grass-field-753868/
    cover_horizontal: FALSE
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  dev = "cairo_pdf"
)
library(tidyverse)
library(INLA)
library(inlabru)
library(gstat)
library(INBOtheme)
if (interactive()) {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 12, 
      transparent = "plot"
    )
  )
} else {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 8, 
      transparent = "plot"
    )
  )
  update_geom_defaults("text", list(size = 3))
}
set.seed(200190307)
```

# Checking spatial autocorrelation

## Pearson residuals

### Definition

> - components?
> - observed value ($y_i$), fitted value ($\hat{y}_i$), mean squared error ($MSE$)
> - formula
> - $$pr_i = \frac{y_i - \hat{y}_i}{\sqrt{MSE}}$$
> - $MSE$ (variance) depends on distribution! Check it using `inla.doc("name_of_your_distribution")`.

### Example data: rainfall in Parana state, Brazil

```{r rain-raw, echo = FALSE}
data(PRprec)
PRprec %>%
  transmute(Longitude, Latitude, Altitude, Rain = d0701) %>%
  filter(!is.na(Rain), Rain > 0) -> dataset
ggplot(dataset, aes(x = Longitude, y = Latitude, colour = Rain, size = Rain)) +
  geom_point() + coord_map()
```

```{r project, echo = FALSE}
coordinates(dataset) <- ~ Longitude + Latitude
dataset@proj4string <- CRS("+proj=longlat")
spTransform(dataset, CRS("+init=epsg:5880")) %>%
  as.data.frame() %>%
  rename(X = Longitude, Y = Latitude) %>%
  mutate(X = X / 100000, Y = Y / 100000) %>%
  mutate(X = X - mean(X), Y = Y - mean(Y)) -> dataset
```

### Calculate Pearson residuals

```{r model-iid}
model_iid <- inla(Rain ~ X + Y, family = "gamma", data = dataset, 
                  control.compute = list(waic = TRUE))
dataset %>%
  mutate(
    mu = model_iid$summary.fitted.values$mean,
    sigma2 = mu ^ 2 / model_iid$summary.hyperpar[1, "mean"],
    Pearson_iid = (Rain - mu) / sqrt(sigma2)
  ) -> dataset
```

## Variogram

### Definition

```{r vg-default}
vg_default <- variogram(Pearson_iid ~ 1, locations = ~X + Y, data = dataset, 
                cressie = TRUE)
```

```{r vg-default-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 1.7, nugget = 0.375) %>%
  fit.variogram(object = vg_default) -> vgm_default
characteristics <- tribble(
  ~id, ~label, ~x, ~xend, ~y, ~yend,
  1, "range", 0, 3 * vgm_default$range[2], 0.25, 0.25,
  3, "nugget", 0, 0, 0, vgm_default$psill[1],
  2, "sill", 0.1, 0.1, 0, sum(vgm_default$psill),
  4, "partial sill", 0, 0, vgm_default$psill[1], sum(vgm_default$psill)
)
vgl_default <- variogramLine(vgm_default, maxdist = max(vg_default$dist))
ggplot(vg_default, aes(x = dist, y = gamma)) +
  geom_line(data = vgl_default) +
  geom_point() +
  scale_x_continuous("distance (100 km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Important characteristics

```{r vg-default-char, echo = FALSE}
ggplot(vg_default, aes(x = dist, y = gamma)) +
  geom_line(data = vgl_default) +
  geom_point() +
  geom_hline(yintercept = cumsum(vgm_default$psill), linetype = 2) +
  geom_vline(xintercept = 3 * vgm_default$range[2], linetype = 3) +
  geom_segment(data = characteristics, arrow = arrow(ends = "both"),
               aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_label(data = characteristics, hjust = 0,
             aes(x = (x + xend) / 2, y = (y + yend) / 2, label = id)) +
  scale_x_continuous("distance (100 km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Important characteristics

```{r vg-default-char-lab, echo = FALSE}
ggplot(vg_default, aes(x = dist, y = gamma)) +
  geom_line(data = vgl_default) +
  geom_point() +
  geom_hline(yintercept = cumsum(vgm_default$psill), linetype = 2) +
  geom_vline(xintercept = 3 * vgm_default$range[2], linetype = 3) +
  geom_segment(data = characteristics, arrow = arrow(ends = "both"),
               aes(x = x, xend = xend, y = y, yend = yend)) +
  geom_label(data = characteristics, hjust = 0,
             aes(x = (x + xend) / 2, y = (y + yend) / 2, label = label)) +
  scale_x_continuous("distance (100 km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Projected example data

```{r rainfall-projected, echo = FALSE}
ggplot(dataset, aes(x = X, y = Y, colour = Rain, size = Rain)) +
  geom_point() + coord_fixed() + labs(x = "X (100 km)", y = "Y (100 km)")
```

### Increased cutoff

```{r vg-large-cutoff}
vg_large <- variogram(Pearson_iid ~ 1, locations = ~X + Y, data = dataset,
                      cutoff = 6, cressie = TRUE)
```

```{r vg-large-cutoff-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 1.7, nugget = 0.375) %>%
  fit.variogram(object = vg_large) -> vgm_large
vgl_large <- variogramLine(vgm_large, maxdist = max(vg_large$dist))
ggplot(vg_large, aes(x = dist, y = gamma)) +
  geom_line(data = vgl_large) +
  geom_point() +
  scale_x_continuous("distance (100 km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Number of point pairs is important

```{r vg-large-np, echo = FALSE}
ggplot(vg_large, aes(x = dist, y = gamma)) +
  geom_line(data = vgl_large) + geom_point(aes(size = np)) + 
  scale_size("number of\npoint pairs", limits = c(0, NA)) +
  scale_x_continuous("distance (100 km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA))
```

### Too small width leads to unstable variograms

```{r vg-small}
vg_small <- variogram(Pearson_iid ~ 1, locations = ~X + Y, data = dataset, 
                      width = 0.01, cressie = TRUE)
```

```{r vg-small-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 1.7, nugget = 0.375) %>%
  fit.variogram(object = vg_small) -> vgm_small
vgl_small <- variogramLine(vgm_small, maxdist = max(vg_default$dist))
ggplot(vg_small, aes(x = dist, y = gamma)) +
  geom_line(data = vgl_small) +
  geom_point(aes(colour = cut(np, c(0, 10, 20, 50, 100, 200, 500, Inf)))) +
  scale_x_continuous("distance (100 km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA)) +
  labs(colour = "number of\npoint pairs")
```

### Sensible small width yields the most informative variogram

```{r vg-final}
vg_final <- variogram(Pearson_iid ~ 1, locations = ~X + Y, data = dataset, 
                      width = 0.1, cressie = TRUE)
```

```{r vg-final-plot, echo = FALSE}
vgm(psill = 0.5, model = "Mat", range = 1.7, nugget = 0.375) %>%
  fit.variogram(object = vg_final) -> vgm_final
vgl_final <- variogramLine(vgm_final, maxdist = max(vg_default$dist))
ggplot(vg_final, aes(x = dist, y = gamma)) +
  geom_line(data = vgl_final) +
  geom_point(aes(colour = cut(np, c(0, 10, 20, 50, 100, 200, 500, Inf)))) +
  scale_x_continuous("distance (100 km)", limits = c(0, NA)) +
  scale_y_continuous("variance", limits = c(0, NA)) +
  labs(colour = "number of\npoint pairs")
```
