--- 
title: "Random intercept models with INLA"
authors: [thierryo]
site: bookdown::bookdown_site
output:
  bookdown::pdf_book:
    base_format: INBOmd::inbo_slides
    toc: FALSE
    slide_level: 2
    theme: inboenglish
    flandersfont: TRUE
    cover: pexels-photo-167698.jpeg # Photo by Lum3n.com from Pexels https://www.pexels.com/photo/conifer-dawn-daylight-evergreen-167698/?utm_content=attributionCopyText&utm_medium=referral&utm_source=pexels
    cover_horizontal: FALSE
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  dev = "cairo_pdf"
)
library(tidyverse)
library(INBOtheme)
if (interactive()) {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 12, 
      transparent = "plot"
    )
  )
} else {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 8, 
      transparent = "plot"
    )
  )
  update_geom_defaults("text", list(size = 3))
}
set.seed(200190307)
```

## Concept of the workshop

- 5 generic challenges
- everybody tries to tackle the challenge with own data
- stuck? ask your neighbour for help
- still stuck? ask me

slides, code, data and HackMD: https://inbo.github.io/tutorials/tutorials/r_inla/

# Fixed effect model

## Challenge 1

1. fit fixed effect model
2. extract $WAIC$ from the model
3. display fixed effect parameters in a table

## Prepare data

```{r prepare_data}
cars <- data.frame(
  fuel_consumption = 3.785411784 / 0.01609344 / mtcars$mpg, # liter / 100 km
  cc = mtcars$disp * 2.54 ^ 3, # engine displacement in cm³
  gearbox = factor(mtcars$am, levels = 0:1, labels = c("auto", "manual"))
)
summary(cars)
```

## Solution 1

```{r solution-1}
library(INLA, quietly = TRUE)
model <- inla(fuel_consumption ~ cc * gearbox, data = cars, 
              control.compute = list(waic = TRUE))
model$waic$waic
```

## Parameters

```{r oplossing-1-parameters, echo = FALSE}
model$summary.fixed %>%
  select(1, lcl = 3, ucl = 5) %>%
  kable(caption = "model parameters", digits = 5)
```

## Scaling

- based on mean and standard deviation (standardise)
```{r standardise}
cars$cc_std <- scale(cars$cc)
attr(cars$cc_std, "scaled:center") # reference (cc)
attr(cars$cc_std, "scaled:scale") # scale (cc)
model_std <- inla(fuel_consumption ~ cc_std * gearbox, data = cars, 
                  control.compute = list(waic = TRUE))
```

- based on carefully picked values

```{r scale}
cars$liter <- cars$cc / 1000
cars$liter_c <- cars$liter - 4 # reference = 4 liter = 4000 cc
model_liter <- inla(fuel_consumption ~ liter_c * gearbox, data = cars, 
                    control.compute = list(waic = TRUE))
```

## Effect on parameters

```{r parameter-scale, echo = FALSE}
model$summary.fixed %>%
  select(mean = 1) %>%
  rownames_to_column("parameter") %>%
  bind_cols(
    model_std$summary.fixed %>%
      select(mean_std = 1),
    model_liter$summary.fixed %>%
      select(mean_liter = 1)
  ) %>%
  kable(
    caption = "model parameters for different scaling", 
    digits = 5
  )
```

## Challenge 2

1. what is the default prior for a fixed effect (`?control.fixed`)
2. use a custom prior for a fixed effect (`?inla`)
3. specify two linear combinations [r-inla.org, FAQ 17](http://www.r-inla.org/faq#TOC-I-have-some-linear-combinations-of-the-nodes-in-the-latent-field-that-I-want-to-compute-the-posterior-marginal-of-is-that-possible-)

## Default prior for fixed effect

- $\mu = 0$
- $\tau = 0.001 \Rightarrow \sigma^2 = 1/\tau = 1000 \Rightarrow \sigma = 31.63$

```{r fixed-prior-default, echo = FALSE}
data.frame(x = seq(-100, 100, length = 101)) %>%
  mutate(density = dnorm(x, 0, sqrt(1 / 0.001))) %>%
  ggplot(aes(x = x, y = density)) +
  geom_line() +
  geom_vline(xintercept = 0, linetype = 2)
```

---

```{r extreme-parameter}
cars$extreme <- cars$liter / 100
model_extreme <- inla(fuel_consumption ~ extreme * gearbox, data = cars)
z <- model_liter$marginals.fixed$liter_c
z[, "x"] <- z[, "x"] * 100
z[, "y"] <- z[, "y"] / 100
plot(model_extreme$marginals.fixed$extreme, type = "l")
lines(z, col = "blue")
```

## Change fixed effect prior

```{r fixed-prior}
model_extreme2 <- inla(fuel_consumption ~ extreme * gearbox, data = cars, 
  control.fixed = list(
    mean = c(extreme = 100), 
    prec = c(extreme = 1e-7, "extreme:gearboxmanual" = 1e-7)))
plot(model_extreme$marginals.fixed$extreme, type = "l")
lines(z, col = "blue")
points(model_extreme2$marginals.fixed$extreme, col = "red")
```

## Linear combinations with fixed effects

```{r lincomb}
combinations <- expand.grid(liter = pretty(cars$liter), 
                            gearbox = unique(cars$gearbox)) %>%
  mutate(liter_c = liter - 4)
model.matrix(~ liter_c * gearbox, combinations) %>%
  as.data.frame() %>%
  inla.make.lincombs() %>%
  setNames(paste(combinations$gearbox, combinations$liter, sep = ":")) -> lc
model_lc <- inla(fuel_consumption ~ liter_c * gearbox, data = cars, 
                 lincomb = lc, control.compute = list(waic = TRUE))
```

---

```{r lincomb-derived}
model_lc$summary.lincomb #see ?control.inla
model_lc$summary.lincomb.derived
```

## Plot linear combinations

```{r plot-lincomb, echo = FALSE}
model_lc$summary.lincomb.derived %>%
  select(2, lcl = 4, ucl = 6) %>%
  bind_cols(combinations) %>%
  ggplot(aes(x = liter, y = mean, ymin = lcl, ymax = ucl)) +
  geom_ribbon(aes(fill = gearbox), alpha = 0.1) +
  geom_line(aes(colour = gearbox)) +
  xlab("engine displacement (liter)") + 
  ylab("estimated fuel_consumption (liter / 100 km)")
```

# Random intercept model ('iid')

## Challenge 3

1. fit a model with one or more random intercepts (`model = 'iid'`)
1. what is the default prior for `'iid'` (`inla.doc('iid')`)
1. calculate $\sigma$ for the random intercept
1. display the BLUP the random intercept

## Default prior

$$\log\tau \sim logGamma(1, 0.00005)$$
$$\tau \sim Gamma(1, 0.00005)$$

```{r iid-prior-precision, echo = FALSE}
data.frame(tau = seq(0, 1000, length = 10000)) %>%
  mutate(densiteit = dgamma(tau, 1, 5e-5)) %>%
  ggplot(aes(x = tau, y = densiteit)) +
  geom_line() +
  xlab("precisie") + ylim(c(0, NA))
```

## $\sigma$ random intercept

```{r solution-3-sigma}
read.delim("ButterfliesNEggs_V4.txt") %>%
  mutate(TreeHeight = TreeHeight / 100 - 1,
         Distance2Edge = Distance2Edge / 10 - 1,
         SmallOakAbundance = SmallOakAbundance / 10 - 0.2) -> butterfly
model <- inla(NEggs ~ NLowBranches + TreeHeight + SmallOakAbundance + 
                f(Area, model = "iid"), family = "poisson", data = butterfly)
model$summary.hyperpar
to_sigma <- function(tau){sqrt(1/tau)}
model$marginals.hyperpar$`Precision for Area` %>%
  inla.tmarginal(fun = to_sigma) %>%
  inla.zmarginal()
```

## Best Linear Unbiased Predictor (BLUP)

```{r solution-3-blup}
glimpse(model$summary.random$Area)
```

## Challenge 4

1. Think about the relevant magnitude of $\sigma$ for your random effect
1. Use a custom "pc.prec" prior with that $\sigma$ (`inla.doc("pc.prec")`)

## Penalised Complexity prior

$Prob(\sigma > u) = \alpha$ met $u > 0$ en $0 < \alpha < 1$

```{r pcprior, echo = FALSE}
expand.grid(
  sigma = seq(0, 3, length = 101),
  alpha = c(0.05, 0.1, 0.5, 0.9)
) %>%
  mutate(
    prob = inla.pc.pprec(sigma ^ -2, u = 0.5, alpha = alpha),
    f = factor(sprintf("%.2f", alpha))
  ) %>%
  ggplot(aes(x = sigma, y = prob, colour = f)) +
  geom_line() +
  geom_hline(aes(yintercept = alpha, colour = f), linetype = 2) + 
  geom_vline(xintercept = 0.5, linetype = 2) +
  labs(x = expression(sigma), colour = expression(alpha), title = "u = 0.5",
       y = expression(paste("Prob(", sigma, "|u)")))
```

## `inlatools` package

https://inlatools.netlify.com

```{r install-inlatools, eval = FALSE}
remotes::install_github("inbo/inlatools")
```

- assessing $\sigma$ random intercept
- assessing $\sigma$ random walk
- check dispersion
- check distribution
- extract fitted values, observed values, Pearson residuals

## Assessing $\sigma$ random intercept

```{r plot-sim-iid}
library(inlatools)
plot(simulate_iid(sigma = 2))
```

## Assessing $\sigma$ random intercept with `link` and `center`

```{r plot-sim-iid-2}
plot(simulate_iid(tau = 16), link = "log", center = "bottom")
```

## Solution 4

```{r solution-4}
model_pc <- inla(NEggs ~ NLowBranches + TreeHeight + SmallOakAbundance + 
                  f(Area, model = "iid", 
                    hyper = list(
                      theta = list(prior = "pc.prec", param = c(0.1, 0.05)))), 
                 family = "poisson", data = butterfly)
model_pc$marginals.hyperpar$`Precision for Area` %>%
  inla.tmarginal(fun = to_sigma) %>%
  inla.zmarginal()
```

## Check dispersion

```{r disp-check-error, error = TRUE}
plot(dispersion_check(model_pc))
model_pc <- inla(NEggs ~ NLowBranches + TreeHeight + SmallOakAbundance + 
                  f(Area, model = "iid", 
                    hyper = list(
                      theta = list(prior = "pc.prec", param = c(0.1, 0.05)))), 
                 family = "poisson", data = butterfly,
                 control.predictor = list(compute = TRUE))
```

## Clear overdispersion

```{r disp-check-poisson}
plot(dispersion_check(model_pc))
```

## Excess of zeros

```{r distr-check}
plot(fast_distribution_check(model_pc))
```

## Linear combinations including random intercept

Create a matrix for each random effect

- 1 row per linear combination
- 1 column per random effect level

```{r model-lc-random}
Area <- matrix(0, nrow = 3, ncol = 22)
Area[1, 5] <- 1
Area[2, c(3, 4)] <- c(1, -1)
Area[3, ] <- 1 / 22
lc1 <- inla.make.lincombs(NLowBranches = c(2, 0, 1), Area = Area)
names(lc1) <- c("Area 5", "Area 3 - Area 4", "average area")
lc2 <- inla.make.lincombs(NLowBranches = 1)
names(lc2) <- "fixed only"
lc <- c(lc1, lc2)
str(lc, 1)
```

---

```{r model-lc-random-fit}
model_lc <- inla(NEggs ~ NLowBranches + TreeHeight + SmallOakAbundance + 
                  f(Area, model = "iid", 
                    hyper = list(
                      theta = list(prior = "pc.prec", param = c(0.1, 0.05)))), 
                 family = "poisson", data = butterfly, lincomb = lc)
model_lc$summary.lincomb.derived # estimate are always on the link scale!
```

## Back transform to natural scale

```{r back-transform}
exp(model_lc$summary.lincomb.derived["average area", 4:6])
inla.tmarginal(exp, model_lc$marginals.lincomb.derived$`average area`) %>%
  inla.zmarginal()
```


# First order random walk model ('rw1')

## Definition

$$\Delta x_i = x_i - x_{i-1} \sim \mathcal{N}(0, \sigma^2)$$

$$x_i \sim \mathcal{N}(x_{i-1}, \sigma^2)$$


```{r eval = FALSE}
inla.doc("rw1")
```

- Useful in case of non-linear patterns in discrete variables (year, day, ...)
- Works with discretised continuous variables (e.g. after rounding)

## Divergent series

```{r rw1-divergence}
rw1 <- simulate_rw(sigma = 0.1)
plot(select_divergence(rw1), link = "log")
```

## Series with frequency change in direction

```{r rw1-jojo}
plot(select_change(rw1), link = "log")
```

## Series with central maximum

```{r rw1-poly-2}
plot(select_poly(rw1, coefs = c(0, -1)), link = "log")
```

## Challenge 5

1. pick a relevant variable for an 'rw1' model
1. ponder on a relevant $\sigma$ for that model
1. fit model with 'rw1' component and `pc.prec` prior

## Default data exploration `NLowBranches`

```{r eda-nlowbranches, warning = FALSE, message = FALSE}
ggplot(butterfly, aes(x = NLowBranches, y = NEggs)) +
  geom_smooth() + geom_point()
```

## Better data exploration `NLowBranches`

```{r eda-nlowbranches2}
ggplot(butterfly, aes(x = NLowBranches, y = NEggs)) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs", k = 4), 
              method.args = list(family = poisson)) + geom_point(alpha = 0.1)
```

## Relevant $\sigma$ for prior `NLowBranches`

```{r rw1-sigma}
simulate_rw(sigma = 0.25, start = 0, length = 4) %>%
  select_poly(coef = c(1, 1)) %>%
  plot(link = "log", center = "bottom")
```

## Solution 5

```{r solution-5}
model_rw1 <- inla(NEggs ~ TreeHeight + SmallOakAbundance + 
                   f(NLowBranches, model = "rw1",
                     hyper = list(
                       theta = list(prior = "pc.prec", param = c(0.25, 0.05)))) +
                   f(Area, model = "iid", 
                     hyper = list(
                       theta = list(prior = "pc.prec", param = c(0.1, 0.05)))), 
                 family = "poisson", data = butterfly,
                 control.predictor = list(compute = TRUE))
```

## Non-linear pattern `NLowBranches`

```{r solution-5-rw1, echo = FALSE}
model_rw1$marginals.hyperpar$`Precision for NLowBranches` %>%
  inla.emarginal(fun = to_sigma) -> est_sigma
model_rw1$summary.random$NLowBranches %>%
  select(ID, mean, lcl = 4, ucl = 6) %>%
  mutate_at(c("mean", "lcl", "ucl"), exp) %>%
  ggplot(aes(x = ID, y = mean, ymin = lcl, ymax = ucl)) + 
  geom_ribbon(alpha = 0.2) + geom_line() +
  labs(title = bquote(sigma ~ "=" ~ .(round(est_sigma, 3))), 
       x = "NLowBranches", y = "relative effect")
```
