---
title: "Using WMS service in R"
description: "How to use WMS (raster) GIS services within R scripts"
authors: [thierryo]
date: 2018-02-19
categories: ["r", "gis"]
tags: ["gis", "webservice", "r", "maps"]
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE
)
```

WMS stands for [Web Map Service](https://en.wikipedia.org/wiki/Web_Map_Service). The service provides prerendered tiles at different scales. This makes it useful to include them as background images in maps. 

We will use the `leaflet` package to make interactive maps.

```{r}
library(leaflet)
```

First, we define some WMS URLs for Flanders and Belgium to play with:

```{r}
# Flanders:
wms_grb <- "https://geoservices.informatievlaanderen.be/raadpleegdiensten/GRB-basiskaart/wms"
wms_ortho <- "https://geoservices.informatievlaanderen.be/raadpleegdiensten/OMWRGBMRVL/wms"
wms_inbo <- "https://geoservices.informatievlaanderen.be/raadpleegdiensten/INBO/wms"
wms_hunting <- "https://geoservices.informatievlaanderen.be/raadpleegdiensten/Jacht/wms"
# Belgium:
wms_cartoweb_be <- "https://cartoweb.wms.ngi.be/service"
wms_ortho_be <- "https://wms.ngi.be/inspire/ortho/service"
```

- `wms_grb` links to the WMS of the [GRB-basiskaart](http://www.geopunt.be/catalogus/webservicefolder/aa04ae22-2297-98c3-1ffd-3440-5aff-bd2c-8a0cc151), the Flemish cadastral map. It depicts land parcels, buildings, watercourses, roads and railroads.

- `wms_ortho` contains a mosaic of recent [orthophotos](http://www.geopunt.be/catalogus/webservicefolder/418e8e4a-12c1-80a8-8306-fcf4-799c-581d-c4e38594) made during the winter. The layer `Ortho` contains the images, the layer `Vliegdagcontour` detail on the time when the pictures were taken.

- `wms_inbo` is a WMS providing [several layers]().

- `wms_hunting` displays [hunting grounds](http://www.geopunt.be/catalogus/webservicefolder/525f1e17-c7d8-3bf3-550c-82c4-7fb3-e97c-a9bc3a6b) in Flanders.

- `wms_cartoweb_be` provides several [cartographic layers from the National Geographic Institute](https://www.ngi.be/website/aanbod/digitale-geodata/cartoweb-be/) (NGI).

- `wms_ortho_be` provides recent [orthophotos for the whole Belgian territory](https://www.geo.be/catalog/details/29238f19-ac79-4a4a-a797-5490226381ec?l=en), compiled by NGI.

In the WFS tutorial, we present a handy [**overview** of websites]((../spatial_wfs_services/#useful-overviews-of-web-feature-services)) with WMS (and WFS) services.


# Simple maps

WMS layers can be added to a `leaflet` map using the `addWMSTiles()` function.

Note that it is required to define the map view with `setView()`, by providing a map center (as longitude and latitude coordinates) and a zoom level.

```{r fig.cap = "Leaflet map with the GRB-basiskaart as background"}
leaflet() %>% 
  setView(lng = 4.287638, lat = 50.703039, zoom = 15) %>% 
  addWMSTiles(
    wms_grb,
    layers = "GRB_BSK",
    options = WMSTileOptions(format = "image/png", transparent = TRUE)
  )
```

_Note: run the code to see this and the following interactive maps._

```{r fig.cap = "Leaflet map with the orthophoto mosaic as background"}
leaflet() %>% 
  setView(lng = 4.287638, lat = 50.703039, zoom = 15) %>% 
  addWMSTiles(
    wms_ortho,
    layers = "Ortho",
    options = WMSTileOptions(format = "image/png", transparent = TRUE)
  )
```

```{r fig.cap = "Leaflet map with the potential natural vegetation"}
leaflet() %>% 
  setView(lng = 4.287638, lat = 50.703039, zoom = 15) %>% 
  addWMSTiles(
    wms_inbo,
    layers = "PNVeg",
    options = WMSTileOptions(format = "image/png", transparent = TRUE)
  )
```

```{r fig.cap = "Leaflet map of hunting grounds with the OpenStreetMap in the background"}
leaflet() %>% 
  setView(lng = 4.287638, lat = 50.703039, zoom = 14) %>% 
  addTiles(group = "OSM") %>%
  addWMSTiles(
    wms_hunting,
    layers = "Jachtterr",
    options = WMSTileOptions(format = "image/png", transparent = TRUE)
  )
```

By adding `addWMSTiles()` multiple times, several WMSs can be placed on top of each other.

```{r fig.cap = "Leaflet map with the GRB-basiskaart, hunting ground and the OpenStreetMap (OSM) as background"}
leaflet() %>% 
  setView(lng = 4.287638, lat = 50.703039, zoom = 14) %>% 
  addTiles(group = "OSM") %>%
  addWMSTiles(
    wms_grb,
    layers = "GRB_BSK",
    options = WMSTileOptions(format = "image/png", transparent = TRUE),
    group = "GRB"
  ) %>%
  addWMSTiles(
    wms_hunting,
    layers = "Jachtterr",
    options = WMSTileOptions(format = "image/png", transparent = TRUE),
    group = "hunting<br>grounds"
  ) %>%
  addLayersControl(
    baseGroups = "OSM",
    overlayGroups = c("GRB", "hunting<br>grounds"),
    options = layersControlOptions(collapsed = FALSE)
  )
```
