---
title: Estimating densities from a point pattern
author: [thierryonkelinx, falkmielke]
date: 2025-03-12
categories: ["r", "gis"]
tags: ["gis", "r", "maps", "density"]
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
---

# Update 2025: Geospatial Analysis

Not *everything* was better in the good old times.

This tutorial was originally written, some time in 2017, to demonstrate the calculation of spatial point densities.
It got stuck on a version 4 update of the [`PROJ`](https://proj.org/en/stable) project, a system library which handles geospatial coordinate projections.


Time moved on, and at the time of writing, about eight years later, `PROJ` has progressed to version 9.5, and geospatial analysis in R has evolved.
Some major new packages to consider are:

- "simple features" ([`sf`](https://r-spatial.github.io/sf/articles/sf1.html))
- "Spatiotemporal Arrays: Raster and Vector Datacubes" ([`stars`](https://r-spatial.github.io/stars/)).
- [`terra`](https://rspatial.github.io/terra/)
- [`gstat`](https://github.com/r-spatial/gstat)

They can be found on the [r-spatial repository](https://github.com/r-spatial).


The need for point density analysis in undiminished.
Noone updated this PR in the meantime.
Until someone took pity to rebase and write a minor update.


You will find a mini-intro dedicated to solving the task of the original tutorial here at the top.
A much more comprehensive and general starting point for geospatial analysis can be found [here](https://r-spatial.github.io/sf/articles/sf1.html) and [here](https://www.youtube.com/watch?v=u5j13L-n7qQ), for example.
Below is the original attempt as an archive, to make you appreciate how simple geospatial analysis has become in 2025 (though it does not always feel simple).


Make sure to also scan the [other geospatial tutorials](https://tutorials.inbo.be/tags/gis).


# Density Calculation

You probably measured an array of spatial locations, to begin with.

```{r load-points}
points <- read.delim("point-pattern/points.txt", sep = " ")
knitr::kable(dplyr::sample_n(points, 4), digits = 5)
```

For sure, you know which coordinate system those were recorded in. Do you?
Then best transform those points into the right CRS.
The library `sf` can handle geospatial objects, and forces but also supports you in handling CRS.

```{r sf-coordinates}
library("sf")
points_sf <- sf::st_as_sf(points, coords = c("lon", "lat"), crs = 4326)

```

To briefly check your points on a map, you can use `mapview`.

```{r mapview}
library("mapview")
mapview(points_sf, cex = 1, legend = FALSE)

```

And you might like to change CRS.
```{r transform-crs}
points_sf_lambert <- sf::st_transform(points_sf, 31370)

```

By the way, in case you get an `sf` geometry object and want to use the coordinates as base R numbers, you can extract either the geometry (point coordinates) or the associated data with `sf`.

```{r retrieve-geometry}
library("dplyr")
points_lambert <- dplyr::as_tibble(sf::st_coordinates(points_sf_lambert)) %>%
  dplyr::rename(c("lon" = X, "lat" = Y))

# a bit too general: here is no info besides the coordinates
df <- cbind(sf::st_drop_geometry(points_sf_lambert), points_lambert)

knitr::kable(dplyr::sample_n(df, 4), digits = 5)
```


Now the song for which you bought the album: density calculation with `MASS`.


```{r density-mass}

calculate_density_2d <- function(points_sf, cellsize = 10){
  stopifnot(dplyr = require("dplyr"),
          MASS = require("MASS"))

  # extract coordinates
  points <- sf::st_coordinates(points_sf)
  pt_x <- points[, 1]
  pt_y <- points[, 2]

  # range in both dimensions
  xlim <- range(pretty(pt_x)) + c(-100, 100)
  ylim <- range(pretty(pt_y)) + c(-100, 100)

  # number of grid points
  n <- c(
      diff(xlim),
      diff(ylim)
    ) / cellsize + 1

  # calculate density
  dens <- kde2d(
    x = pt_x,
    y = pt_y,
    n = n,
    lims = c(xlim, ylim)
  )

  # differentials
  dx <- diff(dens$x[1:2])
  dy <- diff(dens$y[1:2])
  # sum(dens$z * dx * dy)

  # collect the density grid
  dens <- expand.grid(
      lon = dens$x,
      lat = dens$y
    ) %>%
    mutate(
      density = as.vector(dens$z) * length(points_sf),
      id = seq_along(density)
    ) %>%
    merge(
      data.frame(
        x = dx * (c(0, 0, 1, 1, 0) - 0.5),
        y = dy * (c(0, 1, 1, 0, 0) - 0.5)
      )
    ) %>%
    mutate(
      lon = lon + x,
      lat = lat + y
    )

  return(dens)
}

point_density <- calculate_density_2d(points_sf_lambert)
knitr::kable(dplyr::sample_n(point_density, 5), digits = 2)
```


The last challenge is to plot these densities.
Mapview *can* do it...

```{r density-mapview}
point_density_sf <- sf::st_as_sf(point_density, coords = c("lon", "lat"), crs = 31370)
mapview(
  point_density_sf,
  cex = "density",
  zcol = "density"
)

```

... another versatile library is `stars`.

```{r plot-density-stars}
library("stars")

dens_stars <- stars::st_as_stars(point_density_sf, curvilinear=list(X1=x,X2=y))
plot(dens_stars)
```

... and, not the least, there are `ggplot` and [`ggspatial`](https://paleolimbot.github.io/ggspatial).
But still, these are only grid points.


What you actually might want to achieve is an *interpolation*,
for example [inverse distance weighted interpolation](https://www.paulamoraga.com/book-spatial/spatial-interpolation-methods.html#idw-inverse-distance-weighting).


```{r plot-density-tmap}
library("gstat")
library("terra")
library("tmap")

res <- gstat(formula = density ~ 1, locations = point_density_sf,
             nmax = 16, #or use nrow(point_density_sf) to use all the neighbors locations
             set = list(idp = 1)) # beta = 1 

# raster grid covering map
grid <- terra::rast(point_density_sf, nrows = 256, ncols = 256)
# coordinates of all cells
xy <- terra::xyFromCell(grid, 1:terra::ncell(grid))
coop <- st_as_sf(as.data.frame(xy), coords = c("x", "y"),
                 crs = st_crs(point_density_sf))

resp <- predict(res, coop)
resp$x <- st_coordinates(resp)[,1]
resp$y <- st_coordinates(resp)[,2]
resp$pred <- resp$var1.pred

pred <- terra::rasterize(resp, grid, field = "pred", fun = "mean")

tmap::tm_shape(pred) +
  tmap::tm_raster(
    #col_alpha = 0.6,
    col.scale = tmap::tm_scale_continuous(values = "viridis")
    )
```

There are many other options to plot maps and data.
You may find some more hints in [the November 2024 session](https://coding-club.inbo.be/sessions/20241128_visualize_spatial_data_in_r.html) 
of the [INBO coding club](https://coding-club.inbo.be).


Geospatial data analysis is a living field, and a lot has changed recently.
We hope that this tutorial has given you some directions.


# ARCHIVE: Density Calculation in 2017

In this example we focus on a set of 10450 coordinates in a small area.
The goal is to estimate the local density of points, expressed as the number of points per unit area.
The raw coordinates are given in [WGS84 (EPSG:4326)](https://epsg.io/4326), which is a geodetic coordinate reference system.
That is not suited for calculating distances, so we need to re-project the points into a local projected coordinate reference system.
In this case we use [Belgian Lambert72 (EPSG:3170)](https://epsg.io/31370). 
Next we calculate the density.
To visualise the density, we have to transform the results back to WGS84. 

The data used in this example is real data but centred to a different location for privacy reasons. The dataset is available on [GitHub](https://github.com/ThierryO/my_blog/tree/master/data/20170628).

First we must read the data into R. Plotting the raw data helps to check errors in the data.

```{r read-data, fig.cap = "Raw data", eval=FALSE}
points <- read.delim("point-pattern/points.txt", sep = " ")
library(ggmap)
map <- get_map(
  location = c(lon = mean(points$lon), lat = mean(points$lat)),
  zoom = 17,
  maptype = "satellite",
  source = "google"
)
ggmap(map) +
  geom_point(data = points, alpha = 0.1, colour = "blue", shape = 4)
```

The next step is to convert the dataset to a `SpatialPoints` object with WGS84 projection and re-project it to Belgian Lambert72.
`sp::CRS()` defines the coordinate reference systems (CRS).
`sp::coordinates()<-` is an easy way to convert a `data.frame` into a `SpatialPointsDataFrame`, but without specifying a CRS.
Therefore we need to override the CRS with the correct one.

`sp` will derive a 'WKT2 string' representation from an EPSG-code [^epsg] that we provide, in order to represent the CRS.
The WKT2 string (well known text) is a recent open standard by the Open Geospatial Consortium to represent a CRS, and it replaces the older (deprecated) 'PROJ.4 string'.
Currently, the function to set the CRS is still called `proj4string()`.
`sp::spTransform()` converts the spatial object from the current CRS to another CRS.

[^epsg]: Most coordinate reference systems have an [EPSG](https://en.wikipedia.org/wiki/International_Association_of_Oil_%26_Gas_Producers#European_Petroleum_Survey_Group) code which you can find at http://epsg.io/.

```{r crs-objects, eval=FALSE}
library(sp)
crs_wgs84 <- CRS(SRS_string = "EPSG:4326")
crs_lambert <- CRS(SRS_string = "EPSG:31370")
```

The warning above once again demonstrates that some PROJ.4 information is not supported anymore.

```{r reproject, eval=FALSE}
coordinates(points) <- ~lon + lat
# proj4string() - still the only available function to set the CRS - may in the
# future get a more general name:
proj4string(points) <- crs_wgs84
points_lambert <- spTransform(points, crs_lambert)
```

Once we have the points into a projected coordinate system, we can calculate the densities. We start by defining a grid. `cellsize` is the dimension of the square grid cell in the units of the projected coordinate system. Meters in case of Lambert72. The boundaries of the grid are defined using `pretty()`, which turns a vector of numbers into a "pretty" vector with rounded numbers. The combination of the boundaries and the cell size determine the number of grid cells `n` in each dimension. `diff()` calculates the difference between to adjacent numbers of a vector. The density is calculated with `MASS::kde2d()` based on the vectors with the longitude and latitude, the number of grid cells in each dimension and the boundaries of the grid. This returns the grid as a list with elements `x` (a vector of longitude coordinates of the centroids),  `y` (a vector of latitude coordinates of the centroids) and `z` (a matrix with densities). The values in `z` are densities for the 'average' point per unit area. When we multiply the value `z` with the area of the grid cell and sum all of them we get 1. So if we multiple `z` with the number of points we get the density of the points per unit area.

We use [`dplyr::mutate()`](http://dplyr.tidyverse.org/) to convert it into a `data.frame`. The last two steps convert the centroids into a set of coordinates for square polygons.

```{r density, eval=FALSE}
library(MASS)
library(dplyr)
xlim <- range(pretty(points_lambert$lon)) + c(-100, 100)
ylim <- range(pretty(points_lambert$lat)) + c(-100, 100)
n <- c(
  diff(xlim),
  diff(ylim)
) / params$cellsize + 1
dens <- kde2d(
  x = points_lambert$lon,
  y = points_lambert$lat,
  n = n,
  lims = c(xlim, ylim)
)
dx <- diff(dens$x[1:2])
dy <- diff(dens$y[1:2])
sum(dens$z * dx * dy)
dens <- expand.grid(
  lon = dens$x,
  lat = dens$y
) %>%
  mutate(
    density = as.vector(dens$z) * length(points_lambert),
    id = seq_along(density)
  ) %>%
  merge(
    data.frame(
      x = dx * (c(0, 0, 1, 1, 0) - 0.5),
      y = dy * (c(0, 1, 1, 0, 0) - 0.5)
    )
  ) %>%
  mutate(
    lon = lon + x,
    lat = lat + y
  )
```

In order to visualise the result, we have to re-project the coordinates back to WGS84. Then we can display the raster with a web based background image.

```{r ggmap, fig.cap = "Static image of density", eval=FALSE}
coordinates(dens) <- ~lon + lat
proj4string(dens) <- crs_lambert
dens_wgs <- spTransform(dens, crs_wgs84) %>%
  as.data.frame()
ggmap(map) +
  geom_polygon(data = dens_wgs, aes(group = id, fill = density), alpha = 0.5) +
  scale_fill_gradientn(
    "density\n(#/m²)",
    colours = rev(rainbow(100, start = 0, end = .7)),
    limits = c(0, NA)
  )
```

Using `leaflet` to generate a map was a bit more laborious.
Using the `data.frame dens_wgs`directly failed.
So we converted the `data.frame` to a `SpatialPolygonsDataframe`, which is a combination of a `SpatialPolygons` and a `data.frame`.
The `SpatialPolygons` consists of a list of `Polygons`, one for each row of the `data.frame`.
A `Polygons` object consists of a list of one or more `Polygon` objects.
In this example it is a single polygon which represents the grid cell.

```{r convert-to-Spatial-Polygons, eval=FALSE}
dens_sp <- lapply(
  unique(dens_wgs$id),
  function(i){
    filter(dens_wgs, id == i) %>%
      select(lon, lat) %>%
      Polygon() %>%
      list() %>%
      Polygons(ID = i)
  }
) %>%
  SpatialPolygons() %>%
  SpatialPolygonsDataFrame(
    data = dens_wgs %>%
      distinct(id, density),
    match.ID = FALSE
  )
```

`leaflet` requires a predefined function with a colour palette.
We use `leaflet::colorNumeric()` to get a continuous palette.
Setting `stroke = FALSE` removes the borders of the polygon.
`fillOpacity` sets the transparency of the polygons.

```{r leaflet, fig.cap = "Dynamic map of density", eval=FALSE}
library(leaflet)
pal <- colorNumeric(
  palette = rev(rainbow(100, start = 0, end = .7)),
  domain = c(0, dens_sp$density)
)
leaflet(dens_sp) %>%
  addTiles() %>%
  addPolygons(color = ~pal(density), stroke = FALSE, fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~density)
```

