---
title: Estimating densities from a point pattern
maintainer: Thierry Onkelinx
date: '2017-06-28'
output: 
  html_document: default
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
params:
  cellsize: 20
---

In this example we focus on a set of 10450 coordinates in a small area.
The goal is to estimate the local density of points, expressed as the number of points per unit area.
The raw coordinates are given in [WGS84 (EPSG:4326)](https://epsg.io/4326), which is a geodetic coordinate reference system.
That is not suited for calculating distances, so we need to re-project the points into a local projected coordinate reference system.
In this case we use [Belgian Lambert72 (EPSG:3170)](https://epsg.io/31370). 
Next we calculate the density.
To visualise the density, we have to transform the results back to WGS84. 

The data used in this example is real data but centred to a different location for privacy reasons. The dataset is available on [GitHub](https://github.com/ThierryO/my_blog/tree/master/data/20170628).

First we must read the data into R. Plotting the raw data helps to check errors in the data.

```{r read-data, fig.cap = "Raw data"}
points <- read.delim("point-pattern/points.txt", sep = " ")
library(ggmap)
map <- get_map(
  location = c(lon = mean(points$lon), lat = mean(points$lat)),
  zoom = 17,
  maptype = "satellite",
  source = "google"
)
ggmap(map) +
  geom_point(data = points, alpha = 0.1, colour = "blue", shape = 4)
```

The next step is to convert the dataset to a `SpatialPoints` object with WGS84 projection and re-project it to Belgian Lambert72.
`sp::CRS()` defines the coordinate reference systems (CRS).
`sp::coordinates()<-` is an easy way to convert a `data.frame` into a `SpatialPointsDataFrame`, but without specifying a CRS.
Therefore we need to override the CRS with the correct one.

`sp` will derive a 'WKT2 string' representation from an EPSG-code [^epsg] that we provide, in order to represent the CRS.
The WKT2 string (well known text) is a recent open standard by the Open Geospatial Consortium to represent a CRS, and it replaces the older (deprecated) 'PROJ.4 string'.
Currently, the function to set the CRS is still called `proj4string()`.
`sp::spTransform()` converts the spatial object from the current CRS to another CRS.

[^epsg]: Most coordinate reference systems have an [EPSG](https://en.wikipedia.org/wiki/International_Association_of_Oil_%26_Gas_Producers#European_Petroleum_Survey_Group) code which you can find at http://epsg.io/.

```{r crs-objects}
library(sp)
crs_wgs84 <- CRS(SRS_string = "EPSG:4326")
crs_lambert <- CRS(SRS_string = "EPSG:31370")
```

The warning above once again demonstrates that some PROJ.4 information is not supported anymore.

```{r reproject}
coordinates(points) <- ~lon + lat
# proj4string() - still the only available function to set the CRS - may in the
# future get a more general name:
proj4string(points) <- crs_wgs84
points_lambert <- spTransform(points, crs_lambert)
```

Once we have the points into a projected coordinate system, we can calculate the densities. We start by defining a grid. `cellsize` is the dimension of the square grid cell in the units of the projected coordinate system. Meters in case of Lambert72. The boundaries of the grid are defined using `pretty()`, which turns a vector of numbers into a "pretty" vector with rounded numbers. The combination of the boundaries and the cell size determine the number of grid cells `n` in each dimension. `diff()` calculates the difference between to adjacent numbers of a vector. The density is calculated with `MASS::kde2d()` based on the vectors with the longitude and latitude, the number of grid cells in each dimension and the boundaries of the grid. This returns the grid as a list with elements `x` (a vector of longitude coordinates of the centroids),  `y` (a vector of latitude coordinates of the centroids) and `z` (a matrix with densities). The values in `z` are densities for the 'average' point per unit area. When we multiply the value `z` with the area of the grid cell and sum all of them we get 1. So if we multiple `z` with the number of points we get the density of the points per unit area.

We use [`dplyr::mutate()`](http://dplyr.tidyverse.org/) to convert it into a `data.frame`. The last two steps convert the centroids into a set of coordinates for square polygons.

```{r density}
library(MASS)
library(dplyr)
xlim <- range(pretty(points_lambert$lon)) + c(-100, 100)
ylim <- range(pretty(points_lambert$lat)) + c(-100, 100)
n <- c(
  diff(xlim),
  diff(ylim)
) / params$cellsize + 1
dens <- kde2d(
  x = points_lambert$lon,
  y = points_lambert$lat,
  n = n,
  lims = c(xlim, ylim)
)
dx <- diff(dens$x[1:2])
dy <- diff(dens$y[1:2])
sum(dens$z * dx * dy)
dens <- expand.grid(
  lon = dens$x,
  lat = dens$y
) %>%
  mutate(
    density = as.vector(dens$z) * length(points_lambert),
    id = seq_along(density)
  ) %>%
  merge(
    data.frame(
      x = dx * (c(0, 0, 1, 1, 0) - 0.5),
      y = dy * (c(0, 1, 1, 0, 0) - 0.5)
    )
  ) %>%
  mutate(
    lon = lon + x,
    lat = lat + y
  )
```

In order to visualise the result, we have to re-project the coordinates back to WGS84. Then we can display the raster with a web based background image.

```{r ggmap, fig.cap = "Static image of density"}
coordinates(dens) <- ~lon + lat
proj4string(dens) <- crs_lambert
dens_wgs <- spTransform(dens, crs_wgs84) %>%
  as.data.frame()
ggmap(map) +
  geom_polygon(data = dens_wgs, aes(group = id, fill = density), alpha = 0.5) +
  scale_fill_gradientn(
    "density\n(#/m²)",
    colours = rev(rainbow(100, start = 0, end = .7)),
    limits = c(0, NA)
  )
```

Using `leaflet` to generate a map was a bit more laborious.
Using the `data.frame dens_wgs`directly failed.
So we converted the `data.frame` to a `SpatialPolygonsDataframe`, which is a combination of a `SpatialPolygons` and a `data.frame`.
The `SpatialPolygons` consists of a list of `Polygons`, one for each row of the `data.frame`.
A `Polygons` object consists of a list of one or more `Polygon` objects.
In this example it is a single polygon which represents the grid cell.

```{r convert-to-Spatial-Polygons}
dens_sp <- lapply(
  unique(dens_wgs$id),
  function(i){
    filter(dens_wgs, id == i) %>%
      select(lon, lat) %>%
      Polygon() %>%
      list() %>%
      Polygons(ID = i)
  }
) %>%
  SpatialPolygons() %>%
  SpatialPolygonsDataFrame(
    data = dens_wgs %>%
      distinct(id, density),
    match.ID = FALSE
  )
```

`leaflet` requires a predefined function with a colour palette.
We use `leaflet::colorNumeric()` to get a continuous palette.
Setting `stroke = FALSE` removes the borders of the polygon.
`fillOpacity` sets the transparency of the polygons.

```{r leaflet, fig.cap = "Dynamic map of density"}
library(leaflet)
pal <- colorNumeric(
  palette = rev(rainbow(100, start = 0, end = .7)),
  domain = c(0, dens_sp$density)
)
leaflet(dens_sp) %>%
  addTiles() %>%
  addPolygons(color = ~pal(density), stroke = FALSE, fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~density)
```

