---
title: "Lexical scoping in R"
description: ""
author: "Thierry Onkelinx"
date: "`r Sys.Date()`"
categories: ["r"]
tags: ["r"]
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

This is a reply to [The root of all evil in data science](https://www.linkedin.com/pulse/what-r-just-hoax-ticking-bomb-dr-markus-von-ins) by [Markus von Ins](https://www.linkedin.com/in/dr-markus-von-ins-5935a724).

Markus von Ins claims that it easy to write "viruses" in R because one can rewrite existing functions. 
IMHO, Markus von Ins fails to understand lexical scoping in R.

All code snippets from Markus start with a comment `# from Markus`

## Changing `var()`

First Markus sets a baseline for comparison.

```{r}
# from Markus
x <- c(10.4, 5.6, 3.1, 6.4, 21.7)
var(x)
length(x)
```

Let's see what is in the global environment and how `var()` is defined.

```{r}
ls()
var
```

Then Markus 'overwrites' the `var()` with a different version.

```{r}
# from Markus
var <- function(x){
  n <- length(x)
  mu <- mean(x)
  v <- sum((x - mu) ^ 2) / n
  v
}
var(x)
```

Let's look again at the global environment. Notice that there is a new object `var` which wasn't defined before. 
The new object is the function as defined by Markus. 
The original `var()` from the `stats` package is still available and unchanged. 

```{r}
ls()
var
stats::var
stats::var(x)
```

So we have **not** changed `stats::var()` but masked it by the version in the global environment. 
The one in the global environment is selected because the global environment is always first on the search path.

```{r}
searchpaths()
```

## Changing `length()`

In a second example Markus creates a new function `length()` in the global environment.

```{r}
# from Markus
var(x)
length <- function(x){
  le_2 <- sum(x ^ 2)
  le <- sqrt(le_2)
  le
}
length(x)
var(x)
```

The result of `var(x)` is changed because `var()` in the global environment calls the function `length()`. 
Because the new `length()` is first on the search path, that one is used.

```{r}
ls()
base::length(x)
stats::var(x)
```

## 'Virus' packages can only mask functions

Loading a new package puts it on the search path just after the global environment. In case one of the exported functions has the same name a function further on the search path, the user gets a warning message. 
The user is able the inspected the masking function and test whether it influences his code.

```{r}
library(lme4)
searchpaths()
identical(sigma, lme4::sigma)
identical(sigma, stats::sigma)
library(concordance)
identical(sigma, lme4::sigma)
identical(sigma, stats::sigma)
```

Note that the masking is only affecting the global environment. 
The lexical scope within a package is limited to the package itself and the packages it imports, depends on or suggests. 
A secure way of programming functions in a package is by explicitly importing each external function in the NAMESPACE.

## Save version of user defined `var()`

When we use the `foo::bar()` notation, then masking the function has no effect because our function will always use the function `bar()` from the package `foo`.

```{r}
var(x)
var <- function(x){
  n <- base::length(x)
  mu <- base::mean(x)
  v <- base::sum((x - mu) ^ 2) / n
  v
}
mean <- function(x){0}
sum <- function(x){Inf}
var(x)
stats::var(x)
identical(var, stats::var)
identical(length, base::length)
identical(mean, base::mean)
identical(sum, base::sum)
length(x)
mean(x)
sum(x)
```

## Some guidelines

1. Don't mask existing functions.
1. Check functions masking other functions.
1. Place stable functions into a package.

## More information

- "Advanced R" by Hadley Wickham has a section on [lexical scoping](http://adv-r.had.co.nz/Functions.html#lexical-scoping).
- "R packages" by Hadley Wickham has a section on [NAMESPACE](http://r-pkgs.had.co.nz/namespace.html).
