---
title: "The UTM grid and the MGRS grid: not quite the same"
description: ""
authors: [florisvdh]
date: 2023-09-22
categories: ["gis"]
tags: ["gis", "utm", "mgrs", "grids"]
bibliography: ../spatial.yaml
link-citations: true
csl: '`r cslfile <- file.path(tempdir(), "research-institute-for-nature-and-forest.csl"); download.file("https://github.com/inbo/styles/raw/inbocsl-0.2.0/research-institute-for-nature-and-forest.csl", cslfile); cslfile`'
urlcolor: blue
linkcolor: blue
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
---

```{r setup, include=FALSE}
library(sf)
library(stars)
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(stringr)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
theme_set(theme_bw())
theme_update(
  panel.background = element_blank(),
  panel.ontop = TRUE,
  panel.grid = element_line(colour = "grey40"),
  axis.title = element_blank()
)
countries <- ne_countries(scale = "large", continent = "Europe", returnclass = "sf")
countries <- st_transform(countries, "EPSG:3035")
world <- ne_countries(returnclass = "sf")
world <- st_transform(world, "ESRI:54032")
```

```{r eval=FALSE, include=FALSE}
provinces <- httr::parse_url("https://geo.api.vlaanderen.be/VRBG/wfs") |>
  purrr::list_merge(query = list(request = "GetFeature",
                                 typeName = "VRBG:Refprv",
                                 srsName = "EPSG:3812")) |>
  httr::build_url() |>
  read_sf(crs = 3812)
st_write(provinces, "provinces.gpkg")
```

```{r include=FALSE}
provinces <- read_sf("provinces.gpkg")
```


#### Summary

Grids are widely used in biodiversity monitoring to define the basic spatial units for data collection, data processing or mapping.
In this post I explain differences between the UTM and MGRS grids, provide references to their full definition and refer to some software implementations for MGRS.

Main points:

- The UTM grid and the MGRS grid outside the polar regions are both derived from the UTM (Universal Transverse Mercator) map projection system.
Both grids have a world-wide coverage, with exception of the polar regions in the case of the UTM grid.
In the polar regions, the MGRS derives its grid from the UPS (Universal Polar Stereographic) map projection.
- UTM grid references consist solely of digits (Arabic numerals), while MGRS grid references are alphanumeric (combination of digits & Latin characters).
- When decreasing the precision of a position, coordinates are to be rounded while in (UTM and MGRS) grid references the numeric parts representing easting and northing are to be truncated.
- Projected coordinate reference systems (CRSs) specify a geodetic datum so that coordinates can be linked to a physical location on the Earth's surface. Likewise, you need to specify (and be aware of) the geodetic datum when using a (UTM or MGRS) grid reference for positioning.

#### Terminological notes

The NGA (National Geospatial-Intelligence Agency) Office of Geomatics (<https://earth-info.nga.mil>), part of the US Department of Defense, maintains definitions of the UTM map projection system and the MGRS grid [@national_geospatial-intelligence_agency_universal_2014], as did its predecessor organisations since the inception of these systems in the 1940s [@department_of_the_army_map_1956; @department_of_the_army_map_2001; @palombo_military_2021; @stott_utm_1977].
Their usage of the term 'UTM grid' is not distinguished from the UTM map projection system, hence comes down to using numeric CRS coordinates in meters.

At least in the past, a 'UTM grid' reference system was implemented on _civil_ maps, distinct from mere UTM CRS coordinates and with grid referencing rules to allow different levels of precision (see section [UTM grid](#utm-grid)).
Hence I prefer to maintain the distinction between the terms 'UTM projection' and 'UTM grid', also because 'UTM grid' is sometimes erroneously used to refer to the 'MGRS grid', to which it is related.

Confusion between UTM and MGRS is further compounded by the fact that some (civil) sources that distinguish these grids, erroneously describe the division of UTM zones by 8° wide latitude bands as being part of the UTM map projection system, while this belongs solely to the MGRS.

## Introduction

In biodiversity monitoring the Military Grid Reference System (MGRS) is used a lot, although sometimes it is confusingly referred to as 'the UTM grid'.
Example projects that use this spatial reference system are the Atlas Florae Europaea [@lampinen_utm_2013] and the European Invertebrate Survey [@van_nieukerken_utm_1991; @van_veen_eis-nieuws_2000].
In Belgium various faunistic inventories use this grid.

In this post I explain differences between the UTM and MGRS grids, provide references to their full definition and refer to some software implementations for MGRS.
The full definitions are not given here since the primary aim is to draw attention to important properties and differences.
For the same reason, the UPS portion of MGRS is not further explained here.
This post results from a limited literature study on these topics (see [Bibliography](#bibliography)).

Coordinate reference systems ([CRSs](../../tutorials/spatial_crs_coding/#coordinate-reference-systems-minimal-background)) that are widely used, e.g. on regional, national or global level, have obtained standardized definitions in a coordinated manner (e.g. in the [EPSG dataset](https://epsg.org/)).
However, grid reference systems—associated with a projected CRS, a map projection [^map_projection], or sets of these—seem not as well globally coordinated.

A 'grid' in the context of maps refers to the horizontal and vertical _lines_ chosen in the cartesian coordinate system of a projected CRS, usually in a regular manner (equal distance between all lines).
It may also refer to the rectangular cells that emerge from these lines [@iliffe_datums_2008; @stott_utm_1977].
The distance between the grid lines on a map depends on the used map scale.

[^map_projection]: A map projection defines the conversion from geodetic to projected coordinates (geodetic coordinate system (CS, not CRS) + the projection (conversion) method + numeric projection (conversion) parameters), but without specifying the geodetic datum.


```{r grid, fig.cap="A map of Europe in the projected CRS 'EPSG:3035' ('ETRS89-extended / LAEA Europe'). The grid is from the same CRS."}
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") +
  coord_sf(datum = "EPSG:3035", xlim = c(2.5e6, 6e6), ylim = c(1.5e6, 4.5e6))
```

```{r grid2, fig.cap="The same map of Europe in the projected CRS 'EPSG:3035'. The grid however is from 'EPSG:3034' ('ETRS89-extended / LCC Europe'), reprojected in 'EPSG:3035'."}
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") +
  coord_sf(datum = "EPSG:3034", xlim = c(2.5e6, 6e6), ylim = c(1.5e6, 4.5e6))
```

```{r graticule, fig.cap="The lines that represent parallels and meridians of a geographic CRS are called the graticule. This map still uses the projected CRS 'EPSG:3035' but displays the graticule of geographic CRS 'EPSG:4258' ('ETRS89'), projected in CRS 'EPSG:3035'."}
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") + 
  coord_sf(datum = "EPSG:4258", xlim = c(2.5e6, 6e6), ylim = c(1.5e6, 4.5e6))
```



## UTM projection, UTM grid and MGRS grid

### UTM projection

UTM (Universal Transverse Mercator) was developed during World War II and subsequently refined by the US Army [@buchroithner_geodetic_2017].

UTM is a map projection system for the Earth, composed of 120 single Transverse Mercator map projections (conformal, transverse cylindrical projection) for 60 '**UTM zones**' (numbered 1 - 60), each 6 degrees of longitude wide and reaching from 80°S to 84°N (originally 80°N; @department_of_the_army_map_1956).
120 map projections are defined since for each UTM zone a separate map projection is needed for the northern and the southern hemisphere, the only difference being the false northing value.
For example, the largest part of Belgium (and the whole of Flanders) is situated in UTM zone 31, and there the map projection 'UTM zone 31N' applies [^31N], with 'N' referring to the northern hemisphere.

[^31N]: OGC URN identifier of map projection [UTM zone 31N](https://epsg.org/conversion_16031/UTM-zone-31N.html): `urn:ogc:def:coordinateOperation:EPSG::16031`. Using [`projinfo`](https://proj.org/en/latest/apps/projinfo.html), you can compare with `urn:ogc:def:coordinateOperation:EPSG::16131` for map projection [UTM zone 31S](https://epsg.org/conversion_16131/UTM-zone-31S.html).
More information about OGC URN identifiers can be found in the [OGC URN policy](https://www.ogc.org/about-ogc/policies/ogc-urn-policy/) and more extensively in the OGC Best Practice 'Definition identifier URNs in OGC namespace', the latest version of which can be found on the [OGC Best Practices](https://www.ogc.org/standards/bp/) webpage.


```{r utm-zones-world, fig.cap="A map centered on Greenwich, showing how meridians define the borders of UTM zones.", fig.height=7}
utm_zones <- 
  tibble(x = seq(-57, 93, 6), y = 36, zone = 21:46) |> 
  st_as_sf(coords = c("x", "y"), crs = 4258)
ggplot(data = world) +
  geom_sf(fill = "grey97", colour = "grey50") + 
  coord_sf(datum = "EPSG:4258", xlim = c(-3e6, 4e6), ylim = c(-4e6, 7e6)) +
  geom_sf_label(
    data = utm_zones, 
    aes(label = zone), 
    label.size = 0,
    fill = NA
  ) +
  scale_x_continuous(breaks = seq(-60, 90, 6)) +
  scale_y_continuous(breaks = -85)
```

```{r utm-zones-europe, fig.cap="UTM zones in the European area. The UTM zones are narrower in the north."}
utm_zones <- 
  tibble(x = seq(-57, 93, 6), y = 55, zone = 21:46) |> 
  st_as_sf(coords = c("x", "y"), crs = 4258)
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") + 
  coord_sf(datum = "EPSG:4258", xlim = c(2.5e6, 6e6), ylim = c(1.5e6, 4.5e6)) +
  geom_sf_label(
    data = utm_zones, 
    aes(label = zone), 
    label.size = 0,
    size = 6,
    fill = NA
  ) +
  scale_x_continuous(breaks = seq(-24, 48, 6)) +
  scale_y_continuous(breaks = 0)
```


UTM projects geographical coordinates (degrees) to cartesian coordinates (**meters**), so the result is XY coordinates (eastings and northings) like in other map projections, often resulting in high numbers, especially for Y as it covers the whole range of 0° to 80° or 84° latitude.

In the case of a lower precision, coordinates are rounded (not truncated) following the usual rounding rules of science and engineering [@national_geospatial-intelligence_agency_universal_2014], as would happen in other coordinate systems.

As an example, we calculate the projected easting (X) and northing (Y) of a point in Belgium using the `sf` package in R, at the millimeter precision.
The CRS uses the geodetic datum 'ETRS89' [^ETRS89] and the 'UTM zone 31N' map projection (CRS 'EPSG:25831').

```{r message=FALSE, echo=TRUE}
library(sf)
point <- st_sfc(st_point(c(4.1, 51.1)), crs = "EPSG:4258")
st_transform(point, "EPSG:25831") |> st_coordinates() |> format(nsmall = 3, trim = TRUE)
```

[^ETRS89]: OGC URN identifier of the [ETRS89 ensemble datum](https://epsg.org/datum_6258/European-Terrestrial-Reference-System-1989-ensemble.html): `urn:ogc:def:ensemble:EPSG::6258`.

```{r flanders, fig.cap="A map in the Belgian CRS 'EPSG:3812' ('ETRS89 / Belgian Lambert 2008'), showing provinces of Flanders. The point with geographical coordinates '51.1°N 4.1°E', projected in 'EPSG:3812', is shown in red. It is situated in the province 'Oost-Vlaanderen'. The superposed UTM grid lines and UTM coordinates are from the UTM CRS 'EPSG:25831' ('ETRS89 / UTM zone 31N')—reprojected in 'EPSG:3812'—with lines every 5 km."}
ggplot(provinces) +
  geom_sf() +
  geom_sf(
    data = st_sfc(st_point(c(4.1, 51.1)), crs = "EPSG:4258"), 
    colour = "red"
  ) +
  coord_sf(datum = "EPSG:25831", xlim = c(57e4, 67e4), ylim = c(66e4, 72e4)) +
  scale_x_continuous(breaks = seq(5e5, 6.5e5, 5e3)) +
  scale_y_continuous(breaks = seq(5.6e6, 6e6, 5e3)) +
  theme(
    axis.text.x = element_text(angle = 90),
    panel.grid = element_line(colour = "orchid")
  )
```


### UTM grid

A UTM grid refers to the grid lines and the resulting grid cells portrayed on a map, using the cartesian coordinate system(s) obtained by UTM projection.
These lines can for example be shown every 1, 2 or 10 km, depending on the map scale.
This concept complies with the general description of a 'grid' in the context of mapping (see [Introduction](#introduction)).

The UTM grid lines are labelled with the corresponding coordinate values, often expressed as kilometers except in the SE corner of the map where full coordinates are given in meters, as explained by @stott_utm_1977, and by @national_geospatial-intelligence_agency_universal_2014-1 in the context of the MGRS system.
In the case of kilometer grid labels, the (anterior) digits _before_ the last two (principal) digits are printed in superscript as they generally do not change within the map.
This approach matches general rules of printing the various rectangular grids that emerged since World War I.

```{r}
pts <- expand.grid(
  x = seq(50e3, 950e3, 100e3), 
  y = 3e6 + seq(50e3, 4950e3, 100e3)
)
grids <- tibble(crs = 32626:32639,
                colour = rep(viridisLite::plasma(9)[c(5,6)], 7),
                star = rep(list(st_as_stars(pts)), 14)) |> 
  mutate(
    grid = purrr::map2(star, crs, function(x, crs) {
      st_crs(x) <- st_crs(crs)
      zone_from_30 <- crs - 32630
      zone_bb <- st_bbox(
        c(xmin = (zone_from_30 - 1) * 6, xmax = zone_from_30 * 6, 
          ymin = -85, ymax = 85), 
        crs = st_crs(4326)
      )
      st_as_sf(x) |> 
        st_transform(4326) |> 
        st_intersection(st_as_sfc(zone_bb))
    }),
    geom_3035 = purrr::map2(
      grid, 
      colour,
      ~ geom_sf(data = st_transform(.x, 3035), colour = .y, fill = NA)
      )
  )
```

```{r utm-grid-europe, fig.cap="UTM grid in the European area, using cells of 100 km. As UTM zones (in one hemisphere) each have their own map projection, an angle arises between grid lines in two different UTM zones."}
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") + 
  grids$geom_3035[1:14] +
  coord_sf(datum = "EPSG:4326", xlim = c(2.5e6, 6e6), ylim = c(1.5e6, 4.5e6)) +
  scale_x_continuous(breaks = seq(-24, 48, 6)) +
  scale_y_continuous(breaks = 0)
```

```{r utm-grid-belgium, fig.cap="UTM grid in the area around Belgium, using cells of 100 km."}
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") + 
  grids$geom_3035[5:8] +
  coord_sf(datum = "EPSG:4326", xlim = c(3.5e6, 4.5e6), ylim = c(2.8e6, 3.5e6)) +
  scale_x_continuous(breaks = seq(-24, 48, 6)) +
  scale_y_continuous(breaks = 0)
```

The UTM grid referencing system to identify _grid cells_ at a specific precision follows the following numerical syntax, as explained on civil maps that use it [@stott_utm_1977] [^usarmy_utm]:

```
[UTM zone number] [Truncated integer X-coordinate] [Truncated integer Y-coordinate]
```

[^usarmy_utm]: This UTM grid reference notation is also followed in the 2001 version of US Army's Field Manual 3-25.26 (Map Reading and Land Navigation) for the notation of UTM coordinates [@department_of_the_army_map_2001], of course without truncation.

The spaces can be omitted or kept, or replaced by a point or a slash.
The integers can be truncated at the required precision, hence representing 10^p^ meters.
As the coordinates are always truncated (not rounded) to integers, this effectively means that a UTM grid reference represents a square cell with a size that reflects the chosen precision, potentially clipped by the UTM zone border.
For cells not clipped by the UTM zone border, this also means that their truncated coordinates match the UTM coordinates of their southwest corner; this holds for any precision.

Hence for the Belgian point location referred above with UTM coordinates <code>`r st_transform(point, "EPSG:25831") |> st_coordinates() |> format(nsmall = 3, trim = TRUE) |> paste(collapse = ", ")`</code>, the UTM grid reference at 10-meter precision is <code>`r paste("31", floor(st_transform(point, "EPSG:25831") |> st_coordinates() / 10) |> paste(collapse = " "))`</code>.

The above grid reference can be further shortened by dropping the UTM zone number and anterior digits, e.g. if the context is within the area of one map.
When not dropping prefixes, the UTM grid reference is unique at the global level _if_ it is also clear which hemisphere—i.e. which of the two map projections per UTM zone—applies.[^hemisphere]

[^hemisphere]: Some sources [e.g. @uf_geoplan_center_mgrs_2009] therefore mention the possibility to extend the UTM zone with 'N' or 'S' depending on the hemisphere, resulting in e.g. <code>`r paste("31N", floor(st_transform(point, "EPSG:25831") |> st_coordinates() / 10) |> paste(collapse = " "))`</code>. However this approach is now officially obsoleted, and _full_ (i.e. globally unique) UTM coordinates should contain an MGRS grid zone prefix (see MGRS section) instead (@national_geospatial-intelligence_agency_universal_2014-1: appendix A).

The concepts of UTM grid portrayal on maps and UTM grid referencing with anterior and principal digits match those of other rectangular grids that have also long been in use on civil maps, and these are all named 'civil' grids by @stott_utm_1977.

Note that the UTM grid defines _no_ subdivisions of the cartesian coordinate system by means of latitude bands, nor does it describe an alphanumeric geocoding system to uniquely identify further subdivisions.
These topics are part of the Military Grid Reference System (MGRS) specification.

### MGRS grid (Military Grid Reference System)

The Military Grid Reference System (MGRS) was designed by the end of the 1940s by the US Army and is used by NATO militaries for locating positions worldwide up to 1 m precision.
The MGRS specification is maintained by the NGA Office of Geomatics ^[URL: <https://earth-info.nga.mil>.] [@national_geospatial-intelligence_agency_universal_2014].
Best use the foregoing reference when referring to the MGRS.
Other resources describe the MGRS too, including @wikipedia_contributors_military_2023.

Biodiversity monitoring and perhaps other projects sometimes claim to use UTM grid references, while they actually use MGRS grid references.
Hence it is good to understand the difference; see also the [Terminological notes](#terminological-notes).

The coordinate system obtained by **UTM** projection is used in the MGRS definition between 80°S and 84°N, and MGRS combines this with a **UPS** (Universal Polar Stereographic) projection to define its grid in the polar regions.
In what follows, _only the UTM portion of MGRS_ is considered.

The MGRS grid pattern between 80°S and 84°N has two important differences with the pattern of the [UTM grid](#utm-grid)—apart from using a different referencing (geocoding) system:

- In MGRS the UTM grid cell pattern is _intersected_ by latitude bands of generally 8° wide (starting at the equator).
Intersecting [UTM zones](#utm-projection) with latitude bands results in large cells called _**MGRS grid zones**_ of generally 6° longitude by 8° latitude large.
- In MGRS, some of the MGRS grid zones are wider or narrower than 6 longitudinal degrees.
This has been done to avoid grid zone junctions in specific areas, e.g. Bergen and Oslo (Norway) belong to the same grid zone.

```{r utm-grid-latbands-europe, fig.cap="The intersection of the UTM grid by latitude bands of 8° wide forms the basis of the MGRS. This map does not include the subsequent changes to some MGRS grid zones in the north, e.g. in Norway."}
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") + 
  grids$geom_3035[1:14] +
  coord_sf(datum = "EPSG:4326", xlim = c(2.5e6, 6e6), ylim = c(1.5e6, 4.5e6)) +
  scale_x_continuous(breaks = seq(-24, 48, 6)) +
  scale_y_continuous(breaks = seq(0, 80, 8)) +
  theme(panel.grid = element_line(colour = "blue"))
```

Below the grid zone level, it is still the UTM grid pattern that drives the further subdivision in MGRS between 80°S and 84°N, down to the required precision, just as in other grid systems.
However, an important difference at this level is that MGRS uses an _**alphanumeric grid reference**_ instead of a numeric one as in civil grids.
In MGRS, Latin characters are used in referencing both the applicable grid zone and the first subdivision thereof, as follows.

An MGRS 'grid reference' (sometimes called MGRS coordinate) is defined as:

```
[Grid Zone Designation] [Grid Square ID] [Truncated integer X-coordinate] [Truncated integer Y-coordinate]
```

```{r include=FALSE}
utm_coords <- st_transform(point, "EPSG:25831") |> 
  st_coordinates() |> 
  as.vector()
space_mgrs <- function(mgrsref) {
  gzd <- str_sub(mgrsref, end = 3)
  gsid <- str_sub(mgrsref, 4, 5)
  coord <- str_sub(mgrsref, 6)
  x <- str_sub(coord, end = nchar(coord) / 2)
  y <- str_sub(coord, nchar(coord) / 2 + 1)
  paste(gzd, gsid, x, y)
}
```


Hence for the Belgian point location referred earlier this comes down to <code>`r mgrs::utm_to_mgrs(31, "N", utm_coords[1], utm_coords[2]) |> space_mgrs()`</code> at the 1 m precision, which resolves to <code>`r mgrs::utm_to_mgrs(31, "N", utm_coords[1], utm_coords[2], 2) |> space_mgrs()`</code> at 1 km precision.

The spaces can be omitted or kept, but preferably omitted in a computational environment.
The integer coordinates are truncated depending on the required precision, up to _zero_ digits (precision 100 km), with the _maximum_ of 5 digits per coordinate corresponding to a precision of 1 meter (i.e. the unit of the UTM coordinate system).
Consequently, MGRS grid references need less digits per coordinate for 1 m precision than UTM grid references, and this is seen as an advantage of MGRS.

```{r utm-grid-latbands-belgium, fig.cap="The 'U' latitude band goes from 48°N to 56°N. Two resulting MGRS grid zones are fully shown here: grid zones 31U (with most of Belgium) and 32U (with most of Germany)."}
ggplot(data = countries) +
  geom_sf(fill = "grey97", colour = "grey50") + 
  grids$geom_3035[5:8] +
  coord_sf(datum = "EPSG:4326", xlim = c(3.5e6, 4.5e6), ylim = c(2.7e6, 3.7e6)) +
  scale_x_continuous(breaks = seq(-24, 48, 6)) +
  scale_y_continuous(breaks = seq(0, 80, 8)) +
  theme(panel.grid = element_line(colour = "blue"))
```

In the UTM area of MGRS (between 80°S and 84°N), an MGRS grid reference is built as follows:

- The **Grid Zone Designation** (GZD), e.g. `31U`, combines the UTM zone (31) with a letter code (U) that corresponds to a latitude band of generally 8° wide; hence each intersection defines the _**MGRS grid zone**_.
The letter codes in the GZD represent the latitude bands and run from 'C' in the south to 'X' in the north, dropping 'I' and 'O' in order to not confuse with numeric digits.
The X band is 12° high while all other bands are 8° high.
- Each grid zone is further subdivided into _**grid squares**_ (cells) of size 100 km, using the _UTM coordinate system_ to define the cell border coordinates as multiples of 100 000 m.
As grid zones get wider towards the equator, more grid squares are present in zones near the equator than in zones located much further north or south.
The **Grid Square ID** is a code of 2 characters that is unique _within_ the grid zone, e.g. `ES`, with 'E'—the 'easting letter'— referring to a column and with 'S'—the 'northing letter'—referring to a row.
These letters follow the Latin alphabet (excluding 'I' and 'O') in eastern and northern direction within a UTM zone, neglecting parallel grid zone borders and repeating the sequence as needed in northern direction.
The starting letters at UTM coordinates `0, 0` determine both letter sequences in a UTM zone, but depend on the UTM zone number as specified in the rules of the lettering scheme [^scheme].
- To specify locations _**with precision < 100 km**_, the **truncated integer coordinates** of the [UTM grid reference](#utm-grid) are added, thereby _dropping_ the digits that represent the hundred-thousands and higher.
At this level, a precision can be chosen of 1, 10, 100, 1000 or 10 000 meters.

[^scheme]: Moreover, two different lettering schemes are in use (called 'AA' and 'AL'), and it is the ellipsoid used in the geodetic datum that defines which one will be used!

Since coordinates are always truncated (not rounded), resulting in 0 to 5 integers per coordinate, this effectively means that an MGRS grid reference represents a square cell with its size reflecting the chosen precision, although potentially clipped by the UTM zone border or the latitude band.
E.g. in `31U ES 1 4` X = 1 refers to the half-open interval [**1**0, 20) km.

## What is the relation to coordinate reference systems (CRSs)?

Projected CRSs combine a map projection (such as 'UTM 31N') with a geodetic datum that relates actual positions on the Earth's surface to the (unprojected, ellipsoidal) geodetic coordinate system, which also implies defining the ellipsoid and prime meridian.
The geodetic datum is a property of the geodetic CRS associated with the projected CRS.

This also means that without a geodetic datum, you cannot determine the physical location on the Earth's surface that corresponds to a pair of UTM coordinates or to a UTM or MGRS grid reference [^ed50].
See the [CRS tutorial](../../tutorials/spatial_crs_coding/#coordinate-reference-systems-minimal-background) for more information.

[^ed50]: Older series of Belgian maps distributed by the National Geographical Institute (<https://www.ngi.be>) applied the ED50 geodetic datum (European Datum 1950), which affects the position of coordinates (hence MGRS grid) by about 90 m in E-W direction and about 200 m in N-S direction compared to the ETRS89 and WGS 84 datums. OGC URN identifiers of these datums:\
[ED50 datum](https://epsg.org/datum_6230/European-Datum-1950.html): `urn:ogc:def:datum:EPSG::6230`\
[ETRS89 ensemble datum](https://epsg.org/datum_6258/European-Terrestrial-Reference-System-1989-ensemble.html): `urn:ogc:def:ensemble:EPSG::6258`\
[WGS 84 ensemble datum](https://epsg.org/datum_6326/World-Geodetic-System-1984-ensemble.html): `urn:ogc:def:ensemble:EPSG::6326`.

Although the aim of MGRS is positioning and although it needs a compound CRS for its implementation, MGRS is a geocoding standard, not a map projection or a CRS as it adds complexity on top of a compound CRS.

In Belgium, the most relevant UTM CRSs for UTM zone 31 are:

- `r st_crs("EPSG:25831")$Name` (`EPSG:25831`)
- `r st_crs("EPSG:32631")$Name` (`EPSG:32631`)
- `r st_crs("EPSG:23031")$Name` (`EPSG:23031`) which was applied in older topographic maps

Run `sf::st_crs("EPSG:25831")` in R to inspect the WKT string.

## Distribution of the MGRS grid

The NGA Office of Geomatics [distributes](https://earth-info.nga.mil/index.php?dir=coordsys&action=coordsys) worldwide WGS 84 vector layers of MGRS grid polygons [up to 1 km](https://earth-info.nga.mil/index.php?dir=coordsys&action=mgrs-1km-polyline-dloads) precision.

The UF Geoplan Center provides an [MGRS-data site](http://mgrs-data.org) that serves as a library for MGRS data and information, including [geospatial layers](http://mgrs-data.org/data/).

The match of these layers with any WGS 84 UTM CRS and with the WGS 84 graticule can be easily verified post-hoc.

## Criticisms about the MGRS

Criticisms about the MGRS are mainly related to the complexity of the system itself.

- The geocoding system is quite excentric compared to CRSs, especially in current times where computing power is available and handling long numeric coordinates is automated and mathematically simpler.
- The complexity may be a source of mistakes.
A renowned geodesist has advocated on the PROJ mailing list to not use the MGRS outside of the United States.

The @national_geospatial-intelligence_agency_universal_2014 advises:

> Because various characteristics of MGRS are unhelpful to analytical work (see Subsection 11.8), this document suggests (but does not mandate) the following division of labor between MGRS and UTM/UPS when both are under consideration.
UTM/UPS should be be used for calculations, analytical work, and storage & retrieval of geographic
information; MGRS should be limited to notations on maps and charts, displays on end-user devices and person-to-person or person-to-machine communication.

## Software implementations

MGRS and corresponding conversion methods have not been implemented in PROJ because of the technical non-fit.

They have been implemented in geospatial software though.
The following list provides open-source implementations and may be incomplete:

- the [GEOTRANS](https://earth-info.nga.mil) executables and C libraries of the NGA Office of Geomatics.
They provide various coordinate conversions and transformations.
- the C++ library [GeographicLib](https://geographiclib.sourceforge.io).
It grew from a desire to improve on the GEOTRANS package for transforming between geographic and MGRS coordinates.
GeographicLib also provides utilities that can be run in a shell environment, such as `GeoConvert` for converting coordinates to or from MGRS.
These utilities also have an [online](https://geographiclib.sourceforge.io/doc/library.html#online-utilities) implementation.
- a silent (unexposed) implementation as C code inside the [GDAL driver for NITF](https://gdal.org/drivers/raster/nitf.html), an imagery format of the NGA Geospatial-Intelligence Standards Working Group (GWG).
- the R package [mgrs](https://github.com/hrbrmstr/mgrs) which uses GDAL's MGRS implementation in the NITF-driver.
- the Python package `mgrs` ([PyPi](https://pypi.org/project/mgrs/) / [conda-forge](https://anaconda.org/conda-forge/mgrs)) which uses the GEOTRANS C library.
- the QGIS plugin [Lat Lon Tools](https://plugins.qgis.org/plugins/latlontools/) by the US National Security Agency ([code repo](https://github.com/NationalSecurityAgency/qgis-latlontools-plugin)).
It provides several geoprocessing algorithms, which can also be accessed outside the QGIS GUI, e.g. from R [^qgisprocess].
The plugin also provides a GUI for easy coordinate conversion based on interaction with the map.

[^qgisprocess]: Beside R's `mgrs` package, another possibility to translate to/from MGRS grid references in R is by accessing the 'Lat Lon Tools' QGIS plugin using the [qgisprocess](https://r-spatial.github.io/qgisprocess) package, e.g. by calling the algorithm `latlontools:point2mgrs`.

Note that some of these implementations do not require a geodetic datum (hence CRS), but of course (converted) coordinates still need a geodetic datum to refer to an actual position on the Earth's surface and to use them with other geospatial data.

Since it has not been implemented in PROJ, regular geospatial R packages cannot convert to or from MGRS grid references.

Various online web applications allow to interactively determine MGRS coordinates, or display the MGRS grid.
One example is [map.army](https://www.map.army/doc/en/map/coordinategrid/).

## Acknowledgements

NGI / INBO

## Bibliography



