---
title: "Random slopes with perfect correlation"
description: "Illustrate differences between lme4 and INLA when faced with perfectly correlated random slopes"
author: "Thierry Onkelinx"
date: "`r Sys.Date()`"
categories: ["r", "statistics"]
tags: ["r", "analysis", "mixed model"]
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
---



# Read the data

```{r read-data}
library(readr)
library(rprojroot)
dataset <- read_csv(find_root_file(
  "content/tutorials/r_perfect_correlation",
  "correlated-random-effects.csv",
  criterion = is_git_root)
  )
dataset$region <- factor(dataset$region)
dataset$country <- factor(dataset$country)
```


```{r}
head(dataset)
```


```{r}
summary(dataset)
```

# Aggregate binary data

This makes the data more compact and easier to analyse.

```{r prepare_data}
library(dplyr)
dataset <- dataset %>% 
  group_by(region, country, wave) %>% 
  summarise(
    Present = sum(y == 1), 
    N = n(), 
    x1 = mean(x1, na.rm = TRUE), 
    x2 = mean(x2)
  )
dataset$countrywave <- interaction(dataset$country, dataset$wave)
dataset$regionid <- as.integer(dataset$region)
dataset$regionid1 <- dataset$regionid + max(dataset$regionid)
dataset$regionid2 <- dataset$regionid1 + max(dataset$regionid)
dataset$cwave <- dataset$wave - min(dataset$wave)
n.region <- n_distinct(dataset$region)
```


```{r}
dataset
```


```{r}
summary(dataset)
```


# Analyse the model with `lme4`

```{r illustrate-problem}
library(lme4)
system.time(
  model <- glmer(
    cbind(Present, N - Present) ~ 
      x1 + x2 + (1 + x1 + x2 | region) + (1 | country) + (1 | countrywave), 
    data = dataset, 
    family = binomial
  )
)
model
```

This gives perfectly correlated random effects because there is not enough information in the dataset to estimate random slopes at the region level.

# Analyse the model with `INLA`

```{r inla-equivalent}
library(INLA)
system.time(
  m1 <- inla(
    Present ~ x1 + x2 + 
        # INLA equivalent of (1 + x1 + x2|region)
          f(regionid, model = "iid3d", n = 3 * n.region) + 
          f(regionid1, x1, copy = "regionid") + 
          f(regionid2, x2, copy = "regionid") + 
        # INLA equivalent of (1 + x1 + x2|region)
      f(country, model = "iid") + 
      f(countrywave, model = "iid"),
    Ntrials = N,
    data = dataset,
    family = "binomial",
    control.compute = list(dic = TRUE)
  )
)
```


```{r}
summary(m1)
```


```{r}
plot(
  m1$marginals.hyperpar$`Rho1:2 for regionid`, 
  type = "l", 
  main = "Correlation intercept : x1"
)
```


```{r}
plot(
  m1$marginals.hyperpar$`Rho1:3 for regionid`, 
  type = "l", 
  main = "Correlation intercept : x2"
)
```


```{r}
plot(
  m1$marginals.hyperpar$`Rho2:3 for regionid`, 
  type = "l", 
  main = "Correlation x1 : x2"
)
```


Note that the correlations are now around zero with rather wide credible intervals.

`INLA` is a bit slower than `lme4`, but still very fast for a Bayesian technique.

## Alternative models

```{r other-inla-models}
# Remove random effect of region
m2 <- inla(
  Present ~ x1 + x2 + f(country, model = "iid") + f(countrywave, model = "iid"),
  Ntrials = N,
  data = dataset,
  family = "binomial",
  control.compute = list(dic = TRUE)
)
```


```{r}

# Use a random walk along wave per country
m3 <- inla(
  Present ~ x1 + x2 + f(country, model = "iid") + f(cwave, replicate = as.integer(country), model = "rw1"),
  Ntrials = N,
  data = dataset,
  family = "binomial",
  control.compute = list(dic = TRUE)
)
```


```{r}
# Remove random effect of countr:wave
m4 <- inla(
  Present ~ x1 + x2 + f(country, model = "iid"),
  Ntrials = N,
  data = dataset,
  family = "binomial",
  control.compute = list(dic = TRUE)
)
```


```{r}
c(m1$dic$dic, m2$dic$dic, m3$dic$dic, m4$dic$dic)
```

Based on the DIC criterion, there is not that much difference between the models with and without the region effects. 
The effect of wave per country is important.

