--- 
title: "inlabru vs INLA"
author: "Thierry Onkelinx"
site: bookdown::bookdown_site
output:
  bookdown::pdf_book:
    base_format: INBOmd::inbo_slides
    toc: FALSE
    slide_level: 2
    theme: inboenglish
    flandersfont: TRUE
    cover: pexels-photo-1310755.jpeg # Photo by sasha igrevsky from Pexels https://www.pexels.com/photo/bird-s-eye-view-black-and-white-colors-forest-1310755/
    cover_horizontal: FALSE
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE,
  dev = "cairo_pdf"
)
library(tidyverse)
library(INBOtheme)
if (interactive()) {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 12, 
      transparent = "plot"
    )
  )
} else {
  theme_set(
    theme_inbo(
      base_family = "Flanders Art Sans",
      base_size = 8, 
      transparent = "plot"
    )
  )
  update_geom_defaults("text", list(size = 3))
}
set.seed(200190307)
```

# Introduction

## Concept

- first a comparison between `INLA` and `inlabru`
- then time to redo the challenges from the first workshop using `inlabru`

Slides, code and data available on https://inbo.github.io/tutorials/tutorials/r_inla/

## `INLA` or `inlabru`?

- `inlabru` wrapper around `INLA`
- taylored towards spatial data
    - spatially stuff will handled in the next tutorial
- some stuff is easier / better
- some stuff is harder / more awkward

## Toy data set

- Tundra bean goose ([_Anser fabalis subsp. rossicus_](https://www.gbif.org/species/6178319))
- Subset of wintering waterbirds in Flanders (https://doi.org/10.15468/lj0udq)

```{r load-data}
readRDS("anser_fabalis_rossicus.Rds") %>%
  mutate(cyear = year - max(year)) -> goose
glimpse(goose)
```

# Fixed effects only

## Similar syntax

- $WAIC$ and $DIC$ are calculated by default

```{r fixed-syntax}
library(INLA)
m0_inla <- inla(count ~ cyear, data = goose, family = "nbinomial", 
                control.compute = list(waic = TRUE, dic = TRUE))
library(inlabru)
m0_inlabru <- bru(count ~ cyear, data = goose, family = "nbinomial")
```

## `inlabru` returns augmented `INLA` object

```{r fixed-output}
class(m0_inla)
class(m0_inlabru)
all(names(m0_inla) %in% names(m0_inlabru))
names(m0_inlabru)[!names(m0_inlabru) %in% names(m0_inla)]
```

## Careful with factor variables

```{r inlabru-factor-wrong, warning = TRUE}
m1_wrong <- bru(count ~ cyear + month, data = goose, family = "nbinomial")
m1_wrong$summary.fixed
```

## `inlabru` requires dummy variables in case of factors

```{r inlabru-factor-correct}
model.matrix(~month, goose) %>% # create dummy variable for month
  as.data.frame() %>%
  select(-1) %>% # drop intercept
  bind_cols(goose) -> goose
m1_inlabru <- bru(count ~ cyear + monthdec + monthjan + monthfeb, data = goose, 
                  family = "nbinomial")
m1_inlabru$summary.fixed
```

# Random effects

## `inlabru` works slightly different

- you can use any name instead of `f()`
- use `map` to link the random effect to the data (cfr. `ggplot2::aes()`)
- use integer indices in case of a factor
- you need to provide the number of levels in case of a factor

## Example of random intercept

```{r random-intercept}
comp_inla <- count ~ cyear + month + f(location_id, model = "iid")
m2_inla <- inla(comp_inla, data = goose, family = "nbinomial", 
                control.compute = list(waic = TRUE, dic = TRUE))
goose <- mutate(goose, loc_id = as.integer(factor(location_id)))
n_loc <- max(goose$loc_id)
comp_inlabru <- count ~ cyear + monthdec + monthjan + monthfeb + 
  site(map = loc_id, model = "iid", n = n_loc)
m2_inlabru <- bru(comp_inlabru, data = goose, family = "nbinomial")
```

## `inlabru` allows to reuse a variable

```{r inla-unique}
mutate(goose, cyear = cyear - min(cyear) + 1, cyear2 = cyear) -> goose2
comp_inla <- count ~ cyear + f(cyear2, model = "iid") + month + 
  f(location_id, model = "iid")
m3_inla <- inla(comp_inla, data = goose2, family = "nbinomial", 
                control.compute = list(waic = TRUE, dic = TRUE))

n_year <- max(goose2$cyear)
comp_inlabru <- count ~ cyear + rtrend(map = cyear, model = "iid", n = n_year) +
  monthdec + monthjan + monthfeb + site(map = loc_id, model = "iid", n = n_loc)
m3_inlabru <- bru(comp_inlabru, data = goose2, family = "nbinomial")
```

## `map` applies function on the fly

```{r map-function}
comp_inlabru <- count ~  lintrend(map = cyear, model = "linear") + 
  quadtrend(map = cyear ^ 2, model = "linear") +
  rtrend(map = cyear, model = "iid", n = n_year) +
  monthdec + monthjan + monthfeb + site(map = loc_id, model = "iid", n = n_loc)
m3_inlabru2 <- bru(comp_inlabru, data = goose2, family = "nbinomial")
m3_inlabru2$summary.fixed
```

# Plotting the model

## Prepare a model with rw1 and iid component

```{r plot-intercept}
pc_prior <- list(theta = list(prior = "pc.prec", param = c(1, 0.01)))
goose %>%
  mutate(iyear = cyear - min(cyear) + 1) -> goose
n_year <- max(goose$iyear)
comp_inlabru <- count ~ monthdec + monthjan + monthfeb + 
  trend(map = iyear, model = "rw1", n = n_year, hyper = pc_prior) +
  site(map = loc_id, model = "iid", n = n_loc, hyper = pc_prior)
m5_inlabru <- bru(comp_inlabru, data = goose, family = "nbinomial")
```

## Plot the model

```{r plot-intercept}
plot(m5_inlabru)
```

## Plotting different fixed effect

```{r plot-dec}
plot(m5_inlabru, "monthdec") + 
  geom_vline(xintercept = 0, linetype = 2)
```

## Combined plots with `multiplot()`

```{r plot-fixed, echo = FALSE}
p_intercept <- plot(m5_inlabru) + 
  geom_vline(xintercept = 0, linetype = 2) +
  xlim(-1, 4)
p_monthdec <- plot(m5_inlabru, "monthdec") + 
  geom_vline(xintercept = 0, linetype = 2) +
  xlim(-1, 4)
p_monthjan <- plot(m5_inlabru, "monthjan") + 
  geom_vline(xintercept = 0, linetype = 2) +
  xlim(-1, 4)
p_monthfeb <- plot(m5_inlabru, "monthfeb") + 
  geom_vline(xintercept = 0, linetype = 2) +
  xlim(-1, 4)
multiplot(p_intercept, p_monthdec, p_monthjan, p_monthfeb)
```

## Plotting a random effect

```{r plot-trend}
plot(m5_inlabru, "trend")
```

# Predictions

## `inlabru` has a `predict()` method

- no need to refit the model for predicting new data!
- works for `inla()`, `bru()` and `lgcp()` models
- you can specify which components of the models to be used

```{r pred-bru}
goose_trend <- distinct(goose, year, iyear)
pred_trend_log <- predict(m5_inlabru, data = goose_trend, formula = ~ trend)
glimpse(pred_trend_log)
```

## Predictions are easy to plot

```{r pred-bru-plot}
ggplot() + gg(pred_trend_log) + geom_hline(yintercept = 0, linetype = 2)
```

## prediction formula allows functions

```{r}
pred_trend <- predict(m5_inlabru, data = goose_trend, formula = ~ exp(trend))
ggplot() + gg(pred_trend) + geom_hline(yintercept = 1, linetype = 2)
```

## Predictions can use multiple components

```{r pred-multi}
goose_new <- distinct(goose, year, month, iyear, monthdec, monthjan, monthfeb)
pred_goose <- predict(
  m5_inlabru, data = goose_new, 
  formula = ~ exp(Intercept + trend + monthdec + monthjan + monthfeb)
)
```

## Multiple covariates require a bit more work to plot

```{r pred-multi-plot}
ggplot(pred_goose, aes(x = year, y = mean, ymin = q0.025, ymax = q0.975)) +
  geom_ribbon(aes(fill = month), alpha = 0.1) + geom_line(aes(colour = month))
```

## Even aggregations are possible

```{r pred-aggregate}
goose_sum <- filter(goose, month == "jan")
predict(m5_inlabru, data = goose_sum, 
        formula = ~aggregate(
          exp(Intercept + trend + site), 
          by = list(year = goose_sum$year), 
          FUN = sum))
```
