---
title: "How to use open vector file formats in R: GeoPackage & GeoJSON"
description: "A simple tutorial to demonstrate the use of *.gpkg and *.geojson files in R"
authors: [florisvdh]
date: 2020-02-03
categories: ["r", "gis"]
tags: ["gis", "r"]
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
---

```{r include = FALSE}
library(knitr)
library(googlesheets4)
source("token_option.R")
opts_chunk$set(
  echo = TRUE
)
unlink(c("sac.*", "habitatstreams.*"))
```


This tutorial uses a few basic functions from the [dplyr](https://dplyr.tidyverse.org) and [sf](https://r-spatial.github.io/sf/) packages.
While only a few functions are used, you can use the previous hyperlinks to access the tutorials (vignettes) of these packages for more functions and information.

```{r warning=FALSE, message=FALSE}
options(stringsAsFactors = FALSE)
library(tidyverse)
library(sf)
library(inborutils)
```

You will find a bit more background about 'why and what', regarding the considered open standards, in [a separate post](../../articles/geospatial_standards/) on this website.

In short, the GeoPackage and GeoJSON formats are ideal for exchange, publication, interoperability & durability and to open science in general.

The below table compares a few vector formats that are currently used a lot.
This tutorial focuses on the open formats.

```{r message=FALSE, echo=FALSE, warning=FALSE}
sheets_read("15KVm87QvVis--vyCHa8EQBcXAeQ23M1oaCL_dENempk", 
            sheet = "Vector",
            col_types = "c") %>% 
  kable(align = "lcccc")
```


## How to make and use GeoPackages (`*.gpkg`)

### Making a GeoPackage from a geospatial `sf` object in R

As an example, we download a geospatial layer of Special Areas of Conservation in Flanders (version _sac_2013-01-18_) from Zenodo:

```{r message=FALSE}
# meeting a great function from the 'inborutils' package:
download_zenodo(doi = "10.5281/zenodo.3386815")
```

_Did you know this: you can visit a website of this dataset by just prefixing the DOI ^[DOI = Digital Object Identifier. See <https://www.doi.org>.] with `doi.org/`!_

The data source is a shapefile, in this case consisting of 6 different files.
Read the geospatial data into R as an `sf` object, and let's just keep the essentials (though it doesn't matter for the GeoPackage):

```{r}
sac <- 
  read_sf("sac.shp") %>% 
  select(sac_code = GEBCODE,
         sac_name = NAAM,
         subsac_code = DEELGEBIED,
         polygon_id = POLY_ID)
```

Have a look at its contents by printing the object:

```{r paged.print=FALSE}
sac
```

To write the GeoPackage, we just use the GPKG driver of the powerful [GDAL](https://gdal.org) library (supporting most open and some closed formats), which can be elegantly accessed through `sf::st_write()`:

```{r}
sac %>% 
  st_write("sac.gpkg")
```

Is that all?\
_**YES :-)**_

Really?\
_**YES :-)**_

Well, hmmm, if you really want to know a little bit more...

A GeoPackage can contain many layers.
So, it is good practice to explicitly define the layer name within the GeoPackage (above, it was automatically called 'sac').
For example:

```{r}
sac %>% 
  st_write("sac.gpkg",
           layer = "special_areas_conservation",
           delete_dsn = TRUE)
```

Note, `delete_dsn` was set as `TRUE` to replace the whole GeoPackage.
(There is also a `delete_layer` argument to overwrite an existing _layer_ with the same name.)

Let's extract a selection of features from the `special_areas_conservation` layer, and add it as a second layer into the GeoPackage file:

```{r}
sac %>% 
  filter(str_detect(sac_name, "Turnhout")) %>% # only polygons having 'Turnhout' in their name field
  st_write("sac.gpkg",
           layer = "turnhout")
```

So yes, adding layers to a GeoPackage is done simply by `st_write()` again to the same GeoPackage file (by default, `delete_dsn` is `FALSE`), and defining the new layer's name.

So, which layers are available in the GeoPackage?

```{r}
gdalUtils::ogrinfo("sac.gpkg") %>% 
  cat(sep = "\n")
```

_**You see?**_

### Reading a GeoPackage file

Can it become more simple than this?

```{r}
# (note: the 'layer' argument is unneeded if there's just one layer)
sac_test <- st_read("sac.gpkg",
                    layer = "special_areas_conservation")
```

Ready!

`st_read()` is a function of the great `sf` package -- hence the result is an `sf` object again.

Also other geospatial software will (or should) be able to open the GeoPackage format.
It is an open standard, after all!


## How to make and use GeoJSON files (`*.geojson`)

### Making a GeoJSON file from a geospatial `sf` object in R

As another example, let's download a shapefile of stream habitat 3260 in Flanders (version _2018_):

```{r message=FALSE}
download_zenodo(doi = "10.5281/zenodo.3386246")
```

_Again: you can visit a website of this dataset by just prefixing the DOI with `doi.org/`!_

The data source is a shapefile again, in this case consisting of 4 different files.
Similar as above, we read the geospatial data into R as an `sf` object and select a few attributes to work with:

```{r}
habitatstreams <- 
  read_sf("habitatstreams.shp") %>% 
  select(river_name = NAAM,
         source = BRON)
```


```{r paged.print=FALSE}
habitatstreams
```

Nowadays, it is recommended to use the more recent and strict **RFC7946** implementation of GeoJSON.
The previous 'GeoJSON 2008' implementation is now obsoleted
(see [the post](../../articles/geospatial_standards/) on this tutorials website for a bit more background).

The RFC7946 standard is well supported by GDAL's GeoJSON driver, however GDAL must be given the explicit option `RFC7946=YES` in order to use it already ^[
Though GeoJSON 2008 is [obsoleted](http://geojson.org/geojson-spec.html), the now recommended [RFC7946](https://tools.ietf.org/html/rfc7946) standard is still officially in a _proposal_ stage.
That is probably the reason why GDAL does not yet default to RFC7946.
A somehow confusing stage, it seems.
].

Write the GeoJSON file as follows:

```{r}
habitatstreams %>% 
  st_write("habitatstreams.geojson",
           layer_options = "RFC7946=YES")
```

Done creating!

### Do I look good?

Hey wait, wasn't a GeoJSON file just a text file?\
_**Indeed.**_

So I can just open it as a text file to get an idea of its contents?\
_**Well seen :-)**_

Hence, also use it in versioned workflows?\
_**Didn't hear that. (Cool, though...)**_

Let's just look at the top 7 lines of the file:

```{r echo=FALSE, comment=""}
readLines("habitatstreams.geojson", n = 7) %>% 
  cat(sep = "\n")
```

You can see it basically lists the feature attributes and the coordinates of the lines' vertices, with each feature starting on a new line.

Compare the coordinates with those of the `sf` object `habitatstreams` above: the data have automatically been transformed to WGS84!

Note: in order to be still manageable (text file size, usage in versioning systems) it seems wise to use GeoJSON for more simple cases -- **points and rather simple lines and polygons** -- and use the binary GeoPackage format for larger (more complex) cases.

### Reading a GeoJSON file

Just do this:

```{r}
habitatstreams_test <- st_read("habitatstreams.geojson")
```

Same story as for the GeoPackage: other geospatial software will (or should) be able to open the GeoJSON format as well, as it's an open and well established standard.

From the message of `st_read()` you can see the CRS is WGS84 ([EPSG-code 4326](https://epsg.io/4326)) - this is always expected when reading a GeoJSON file.

If you want to transform the data to another CRS, e.g. Belgian Lambert 72 ([EPSG-code 31370](https://epsg.io/31370)), use `sf::st_transform()`:

```{r}
habitatstreams_test %>% 
  st_transform(31370)
```

```{r include=FALSE}
unlink(c("sac.*", "habitatstreams.*"))
```
