---
title: "Using WMS service in R"
date: 2018-02-19T12:09:22+01:00
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE
)
```

WMS stands for [Web Map Service](https://en.wikipedia.org/wiki/Web_Map_Service). The service provides prerendered tiles at different scales. This makes it useful to include them as background images in maps. 

There is currently only one example on this page. We plan to add more examples in the future. Feel free to add more examples through a pull request at https://github.com/inbo/tutorials/pulls. Or create an issue at https://github.com/inbo/tutorials/issues.

```{r}
wms_grb <- "https://geoservices.informatievlaanderen.be/raadpleegdiensten/GRB-basiskaart/wms"
```

# As background of interactive maps

WMS layers can be added to a `leaflet` map using the `addWMSTiles()` function.

```{r}
library(leaflet)
leaflet() %>% 
  setView(lng = 4.287638, lat = 50.703039, zoom = 15) %>% 
  addTiles(group = "OSM") %>%
  addWMSTiles(
    wms_grb,
    layers = "GRB_BSK",
    options = WMSTileOptions(format = "image/png", transparent = TRUE),
    group = "GRB"
  ) %>%
  addLayersControl(
    overlayGroups = c("OSM", "GRB", "CartoWeb"),
    options = layersControlOptions(collapsed = FALSE)
  )
```
