---
title: "Coordinate operations and maps in R"
description: "This tutorial shows how to do coordinate transformations and conversions in R. It also demonstrates how spatial data can be mapped with ggplot2 and mapview."
authors: [florisvdh, ambermertens]
date: 2021-09-16
categories: ["r", "gis"]
tags: ["gis", "r", "maps"]
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl
link-citations: TRUE
bibliography: bibliography.yaml
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  comment = ""
)
options(stringsAsFactors = FALSE)
git_root <- rprojroot::find_root(rprojroot::is_git_root)
tutorial_root <- file.path(git_root, "content/tutorials/spatial_transform_crs_2")
dir.create(file.path(tutorial_root, "data_gisclub"), showWarnings = FALSE)
gisclubdata_path <- file.path(tutorial_root, "data_gisclub")
```

## Introduction and setup

This tutorial originated while preparing an R-demonstration during a GIS club at INBO on 16 September 2021.
(Further material of the GIS club is available [here](https://drive.google.com/drive/folders/1s5HZ5VjnkTbCQSE1vfFyMJz6PVIl4mHW).)

A straightforward approach for transforming spatial data is available in another [tutorial](../spatial_transform_crs/).
The current tutorial tackles a few extra aspects.

The tutorial assumes you have some pre-existing knowledge:

-   Basic knowledge about coordinate reference systems (CRSs) and geodetic datums; see [another tutorial](../spatial_crs_coding/) and the references and links therein.
-   Knowing how to read geospatial files with [`sf`](https://r-spatial.github.io/sf).
There is another [tutorial](../spatial_standards_vector/) demonstrating some aspects.

In this tutorial we will apply conversions (without datum shift) and transformations (with datum shift) to `sf` and `sfc` objects.
In the background, these operations use the [PROJ](https://proj.org) library.

Let's load the needed R packages and prepare a more appropriate theme for mapping with [`ggplot2`](https://ggplot2.tidyverse.org/):

```{r message=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
theme_set(theme_bw())
theme_update(panel.grid = element_line(colour = "grey80"))
library(mapview)
```

Set up input data if needed:

```{r results="hide"}
if (!file.exists(file.path(gisclubdata_path, "gemeenten_belgie.shp"))) {
  googledrive::drive_download(googledrive::as_id("1-epL-fyKB8eS-WwuZhjsl8uyZYplAqFA"), 
                              file.path(tempdir(), "data_gisclub.zip"))
  unzip(file.path(tempdir(), "data_gisclub.zip"),
        exdir = tempdir())
  list.files(file.path(tempdir(), "data_gisclub"), full.names = TRUE) %>% 
    file.copy(gisclubdata_path, recursive = TRUE)
}
```

The code assumes that you have a `gisclubdata_path` object defined (directory path as a string).

## Joining and mapping polygon layers

For a given `municipalities` and `watercourses` dataset, we would like to append the municipality name to the watercourses by making a spatial join.

### Reading data and trying to join

Let's read the data file of Belgian municipalities:

```{r}
path_municipalities <- file.path(gisclubdata_path, "gemeenten_belgie.shp")
municipalities <- read_sf(path_municipalities)
```

It looks like this:

```{r paged.print=FALSE}
municipalities
```

It appears that the points have three coordinate dimensions (`XYZ`).
However the features have only two useful dimensions:

```{r}
st_dimension(municipalities)
```

Some tools -- e.g. spherical geoprocessing -- expect XY data.
So we will drop the Z dimension with `st_zm()`.
Meanwhile we select the only necessary attribute and rename it, using `select()`.

```{r}
municipalities <- 
  municipalities %>% 
  select(municipality = T_MUN_NL) %>% 
  st_zm()
```

We perform similar operations on the `watercourses` dataset.
Here, data are already defined as two dimensions.

```{r}
path_watercourses <- file.path(gisclubdata_path, "watergangen_antw.shp")
watercourses <- read_sf(path_watercourses)
```

```{r paged.print=FALSE}
watercourses
```

Note that not all columns are printed.
Let's preview all columns with `glimpse()`:

```{r}
glimpse(watercourses)
```

We only select some:

```{r}
watercourses <- 
  watercourses %>% 
  select(polygon_id = OIDN,
         vhag_code = VHAG,
         watercourse = NAAM)
```

```{r paged.print=FALSE}
watercourses
```

In the above prints of the `municipalities` and `watercourses` objects, you could already see their coordinate reference system (CRS).
With `st_crs()` you can extract the `crs` object; it is printed as:

```{r}
st_crs(municipalities)
```

The CRS name '`r st_crs(municipalities)$Name`' can also be returned by `st_crs(municipalities)$Name`.
As seen from the WKT string, this is a geographical CRS.

However for `watercourses` the CRS is missing:

```{r}
st_crs(watercourses)
```

Can we make the spatial join already?

```{r}
st_join(watercourses, municipalities) %>% try
```

Fortunately, this doesn't work: `st_join()` requires its inputs to belong to the same CRS.

Well, we know that `watercourses` is in the projected CRS `EPSG:31370` (`r st_crs(31370)$Name`), so we can just update it:

```{r}
st_crs(watercourses) <- "EPSG:31370"
```

Since the CRSs are still different, this should *not* be sufficient for the join.

```{r}
st_join(watercourses, municipalities) %>% try
```

Indeed!

Let's transform `municipalities` to `EPSG:31370`:

```{r}
municipalities_31370 <- st_transform(municipalities, "EPSG:31370")
```

And now...

```{r paged.print=FALSE}
(watercourses_comm <- st_join(watercourses, municipalities_31370))
```

Succeeded!

Two things to note here:

- `st_join()` has an argument '`join`' that defines the type of topological relationship  that is used to either join or _not_ join a geometry from the first layer with a geometry from the second. 
Various so-called 'binary predicates' can be used to define this relationship (see `?st_join`), but the default one is `st_intersects()`, i.e. an _intersection_ as defined by the '[DE-9IM](https://en.wikipedia.org/wiki/DE-9IM)' model of binary geometry relations.
- Since several watercourses intersect more than one municipality, several of them are repeated in `watercourses_comm`, each with another values for `municipality`.

We chose to make the spatial join in the `EPSG:31370` CRS.
However, we can equally do it on a spherical model of the Earth, when using the geographical coordinates.
We expect the resulting joined attributes to be equivalent.
Let's check:

```{r paged.print=FALSE}
st_join(st_transform(watercourses, "EPSG:4326"), municipalities) %>% 
  st_drop_geometry %>% 
  all.equal(watercourses_comm %>% st_drop_geometry)
```

### Mapping

While it has only one attribute and `r nrow(municipalities_31370)` rows, the object size of `municipalities_31370` is relatively large because its polygons have numerous vertices:

```{r}
object.size(municipalities_31370) %>% format(units = "MB")
```

This makes a simple map take long to render in R (not shown here; just try yourself):

```{r ggplot_map_heavy, eval=FALSE}
ggplot(data = municipalities_31370) + geom_sf()
```

For the sake of this exercise, we make a simplified derived object `comm_simpl_31370` with fewer vertices:

```{r}
comm_simpl_31370 <- st_simplify(municipalities_31370, dTolerance = 10)
```

Its size is much smaller:

```{r}
object.size(comm_simpl_31370) %>% format(units = "MB")
```

So it renders faster.

```{r ggplot_graticule_4326}
ggplot(data = comm_simpl_31370) + geom_sf()
```

However, mind that the simplification may be OK for mapping at an appropriate scale, but beware that information has been lost, so shapes have changed and gaps + overlaps will appear between polygons.
Hence don't use this simplified version if you process it further, or for detailed mapping.

Note the latitude-longitude graticule: the map follows the CRS of the plotted object, but the graticule is latitude-longitude by default.

We can plot the graticule of another CRS with `coord_sf()`:

```{r ggplot_graticule_31370}
ggplot(data = comm_simpl_31370) + geom_sf() + coord_sf(datum = "EPSG:31370")
```

The above is the most obvious choice here, since it's the CRS of the object, but we could as well plot the graticule of e.g.
`EPSG:3035` (`r st_crs(3035)$Name`):

```{r ggplot_graticule_3035}
ggplot(data = comm_simpl_31370) + geom_sf() + coord_sf(datum = "EPSG:3035")
```

Next to the [`ggplot2`](https://ggplot2.tidyverse.org/) package, which we used above, interactive maps can be made with the [`mapview`](https://r-spatial.github.io/mapview/) package.

Here too, time to render differs between:

```{r mapview_munic_heavy, eval=FALSE}
mapview(municipalities_31370, legend = FALSE)
```

and:

```{r mapview_munic_multicolour}
mapview(comm_simpl_31370, legend = FALSE)
```

_Note: run the code to see this and the following interactive maps._

In this case, `mapview` is using the `leaflet` package with some nice defaults, but in the above examples it makes a color map by default because there's only one attribute.
We don't want this and we make some further adjustments:

```{r mapview_munic_pink}
mapview(comm_simpl_31370, 
        col.regions = "pink", 
        alpha.regions = 0.3, 
        legend = FALSE,
        map.types = "OpenStreetMap")
```

Note that a separate tutorial about `leaflet` is [available](../spatial_create_leaflet_map/) on this website!

Beware that `mapview` and `leaflet` produce maps in the '`r st_crs(3857)$Name`' CRS (`EPSG:3857`).
For that to happen, `mapview` transforms spatial objects on-the-fly.
[Pseudo Mercator or Web Mercator](https://proj.org/operations/projections/webmerc.html) is a much used projection in web mapping because the coordinate projection is computed much faster.
However it should not be used in official mapping or as a projected CRS for geoprocessing, since it has more distortions (including shape) than e.g. the (truly) conformal Mercator projection (e.g. in CRS `EPSG:3395` -- `r st_crs(3395)$Name` -- which is best used in the equatorial region).

Let's show two layers with `mapview`:

```{r mapview_munic_water}
mapview(list(comm_simpl_31370, watercourses), 
        col.regions = c("pink", "blue"), 
        alpha.regions = c(0.2, 0.5),
        color = c("grey", NA), 
        legend = FALSE,
        map.types = "OpenStreetMap")
```

## Different transformation pipelines

We consider a layer of points that correspond to locations where characteristics were determined of Flemish surfacewater bodies:

```{r}
path_points <- file.path(gisclubdata_path, "meetplaatsen_oppwater.shp")
points <- read_sf(path_points)
```

Check the CRS:

```{r}
st_crs(points)
```

The `points` object has the projected '`r st_crs(points)$Name`' CRS (`EPSG:28992`), which was created for The Netherlands -- onshore.
See e.g.
[Wikipedia](https://nl.wikipedia.org/wiki/Rijksdriehoeksco%C3%B6rdinaten) for more background about this CRS.

Essentially we are dealing with a different projected CRS than `EPSG:31370`.
`EPSG:28992` uses the 'Amersfoort' geodetic datum, defined using the 'Bessel 1841' ellipsoid, while `EPSG:31370` has the 'Reseau National Belge 1972' datum, defined using the 'International 1924' ellipsoid [^1].

[^1]: Note that several European countries, including Belgium, now provide national projected CRSs that use the common ETRS89 datum (European Terrestrial Reference System 1989).
    An example is the Belgian CRS `EPSG:3812` (`r st_crs(3812)$Name`).

We will transform `points` (back) to `EPSG:31370` since the data are outside the intended area of usage.
Note that this dataset has been made for mere demonstrative purposes:
since at least a part of the points fall outside the area of usage (geographic bounding box) of the '`r st_crs(points)$Name`' CRS, the dataset is not appropriate to start from.
But let's suppose it is all one had available to work with!

### Standard approach

By default, the appropriate transformation pipeline (concatenated coordinate operations) can differ between different areas that are represented in a spatial dataset.
This is taken care of automatically, based on the bounding boxes of each CRS, but it will also depend on the availability of specific transformation grids.

```{r}
points_31370 <- st_transform(points, "EPSG:31370")
```

Notice the completely different coordinate ranges between both CRSs:

```{r collapse=TRUE, comment="#>"}
st_geometry(points)
```

```{r collapse=TRUE, comment="#>"}
st_geometry(points_31370)
```

A specific area shown with `ggplot2` in `EPSG:31370`:

```{r ggplot_water_points}
ggplot() + 
  geom_sf(data = watercourses, fill = "lightblue", colour = "lightblue") +
  geom_sf(data = points_31370, fill = "red", shape = 21) +
  lims(x = c(154e3, 155e3), y = c(218e3, 219e3)) +
  coord_sf(datum = "EPSG:31370")
```

### Enforcing a specific pipeline

We can request available transformation pipelines with `sf_proj_pipelines()`:

```{r}
pipelines <- sf_proj_pipelines("EPSG:28992", "EPSG:31370")
```

One can further limit this by specifying the area of interest (`aoi` argument), which we didn't do here.

It is a dataframe with following structure:

```{r}
glimpse(pipelines)
```

Some metadata about the pipelines:

```{r paged.print=FALSE}
pipelines %>% 
  as_tibble %>% 
  select(definition, description, grid_count, accuracy, instantiable)
```

We can see that the first one has best accuracy.
Several pipelines are based on one or two grids for a better accuracy.

When printing the `pipelines` object, its first row is printed in a more readable manner:

```{r}
pipelines
```

Since the rows of pipelines are sorted by `instantiable` and `accuracy`, it makes sense to designate the printed pipeline as _Best instantiable operation_.

It follows that the same information appears when manually selecting and then printing, except for the number of 'candidate operations found':

```{r}
pipelines[1,]
```

If you'd like to get even more metadata on the available pipelines, including the WKT definitions, you can also run PROJ's [`projinfo`](https://proj.org/apps/projinfo.html) command line program from within R if it's available in your system `PATH`.

```{r eval=FALSE}
system("projinfo -s EPSG:28992 -t EPSG:31370 --spatial-test intersects -o WKT2:2019")
```

The *default* operation returned by PROJ's `projinfo`, based on the union of the source and target CRS's bounding box, appears to be just one pipeline with so-called [ballpark accuracy](https://proj.org/glossary.html#term-Ballpark-transformation).
It's obtained by dropping the `--spatial-test intersects` option (or replacing `intersects` by `contains`, which is default)[^projinfo], and so it matches the pipeline where `accuracy` equals `NA` in our `pipelines` dataframe:

```{r}
pipelines[13,]
```

[^projinfo]: In the documentation of `projinfo`, we can read about the `--spatial-test` option: '_Specify  how the area of use of coordinate operations found in the database are compared to the area of use specified explicitly with `--area` or `--bbox`, or derived implicitly from the area of use of the source and target CRS.  By default, projinfo will only keep coordinate operations whose are of use is strictly within the area of interest (**contains** strategy).  If using the **intersects** strategy, the spatial test is relaxed, and any coordinate operation whose area of use at least partly intersects the area of interest is listed._'

The low accuracy of a ballpark transformation is because datum differences between source and target CRSs are neglected:

> It does not attempt any datum shift, hence the "ballpark" qualifier in its name.
> Its accuracy is unknown, and could lead in some cases to errors of a few hundreds of metres.

*(Source: <https://proj.org/glossary.html#term-Ballpark-transformation>)*

With `st_transform()` we can enforce a specific pipeline.
Note that it is only available for `sfc` objects, not `sf` objects.

Let's apply the most accurate pipeline to *all* points:

```{r}
# We select the pipeline with lowest accuracy (< 0.02), by filtering on accuracy.
# If the grids are installed, this should match the first line of the pipelines object.
chosen_pipeline_definition <- pipelines %>% filter(accuracy < 0.02) %>% pull(definition)
points_31370_strict <- 
  st_transform(st_geometry(points), "EPSG:31370",
               pipeline = chosen_pipeline_definition) %>% 
  st_sf(st_drop_geometry(points), geometry = .) %>% 
  as_tibble %>% 
  st_as_sf
```

And compare:

```{r collapse=TRUE, comment="#>"}
st_geometry(points_31370)
```


```{r collapse=TRUE, comment="#>"}
st_geometry(points_31370_strict)
```

Notice the **`r attr(st_geometry(points_31370_strict), "n_empty")` empty geometries** in case of `points_31370_strict`!

What is happening?

```{r ggplot_pipelines}
p <- 
  ggplot() + 
  geom_sf(data = points_31370, colour = "red") +
  geom_sf(data = points_31370_strict, colour = "green") 
p
```

It appears that points too far at sea are considered invalid when applying this pipeline, and their geometry was cleared (although the corresponding row was not dropped).

If we make a zoomable plot, there's more to discover.

```{r ggplotly_pipelines}
plotly::ggplotly(p)
```

Most points on the map are on the exact same location between both approaches.
We can see that points west and south of the *geographic bounding box* for usage of the Dutch CRS have a different position depending on the applied approach.

It can be further investigated with `mapview` (note that this does a further reprojection to `EPSG:3857`):

```{r mapview_pipelines}
mapview(list(points_31370, points_31370_strict),
        col.regions = c("red", "green"))
```

Here it appears that the western and southern points of `points_31370` were transformed with the ballpark-accuracy pipeline:
they are way off the surfacewater bodies!

In this case, one might prefer enforcing the specific pipeline outside the geographic bounding box of `EPSG:28992`, as we did for `points_31370_strict`.
However best explore the other pipelines as well, e.g. the most accurate one that doesn't depend on a grid, and consider using that outside the geographic bounding box.

Bottomline is that you should be suspicious about coordinates outside the CRS's geographic bounding box!
So keep an eye on them when applying a default transformation procedure.


## Further reading and more packages

Much more information can be found in the online books by @pebesma_spatial_2021 and @lovelace_geocomputation_2019.
The `sf` package has a nice website at <https://r-spatial.github.io/sf>; version 0.6-1 was introduced in @pebesma_simple_2018.

There are several more R packages to make beautiful maps.
We specifically mention [`tmap`](https://r-tmap.github.io/tmap/) [@tennekes_tmap_2018] and [`mapsf`](https://riatelab.github.io/mapsf/) for making production-ready thematic maps.
`tmap` is used a lot in @lovelace_geocomputation_2019.

Further, there is the [`plotKML`](https://plotkml.r-forge.r-project.org/) package, which renders spatial objects in Google Earth [@hengl_plotkml:_2015]!

## Bibliography
