---
title: "Observation level random effects"
description: "A way to capture overdispersion in a mixed model with a Poisson distribution"
author: "Thierry Onkelinx"
date: "`r Sys.Date()`"
categories: ["r", "statistics"]
tags: ["r", "analysis", "mixed model"]
output: 
    md_document:
        preserve_yaml: true
        variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Definition

An observation level random effect (OLRE) is a random effect with a different level for each observation. 
Combined with a Poisson distribution it can capture moderate overdispersion.

## Example

The model below is a simple mixed model. 
$\delta_{ijk}$ is the OLRE.

$$Y_{ijk} \sim Poisson(\mu_{ijk})$$

$$\log(\mu_{ijk}) = \eta_{ijk}$$

$$\eta_{ijk} = \beta_0 + \beta_1 X_i + b_j + \delta_{ijk}$$

$$b_j \sim N(0, \sigma_b)$$

$$\delta_{ijk} \sim N(0, \sigma_{olre})$$

Let's generate some overdispersed data using a negative binomial distribution.

```{r echo = TRUE}
set.seed(324)
n.i <- 10
n.j <- 10
n.k <- 10
beta.0 <- 1
beta.1 <- 0.3
sigma.b <- 0.5
theta <- 5
dataset <- expand.grid(
  X = seq_len(n.i),
  b = seq_len(n.j),
  Replicate = seq_len(n.k)
)
rf.b <- rnorm(n.j, mean = 0, sd = sigma.b)
dataset$eta <- beta.0 + beta.1 * dataset$X + rf.b[dataset$b]
dataset$mu <- exp(dataset$eta)
dataset$Y <- rnbinom(nrow(dataset), mu = dataset$mu, size = theta)
dataset$OLRE <- seq_len(nrow(dataset))
```

```{r}
library(ggplot2)
ggplot(dataset, aes(x = X, y = Y, group = Replicate)) +
  geom_line() +
  geom_point() +
  facet_wrap(~b)
```

Next we fit the model with an OLRE.

```{r}
library(lme4)
m1 <- glmer(Y ~ X + (1 | b) + (1 | OLRE),
  data = dataset,
  family = poisson
)
summary(m1)
```

The estimates for $\beta_0$, $\beta_1$ and $\sigma_b$ are reasonably close to the ones used to generate the data.

## Check the sanity of the model

Models with an OLRE should be used carefully. 
Because OLRE can have a very strong influence on the model. 
One should always check the standard deviation of the ORLE. 
High standard deviations are a good indication for problems. 
Let's show why.

The main model (without OLRE) is $\gamma_{ijk} = \beta_0 + \beta_1 X_i + b_j$. 
The OLRE correct this for overdispersion $\eta_{ijk} = \gamma_{ijk} + \delta_{ijk}$.
Since $\delta_{ijk} \sim N(0, \sigma_{olre})$, the 2.5% and 97.5% quantiles of this distribution define a range in which 95% of the plausible OLRE values are. 
The 2.5% quantile ($\delta_{lcl}$) indicates a strong but plausible downward correction, the 97.5% quantile  ($\delta_{ucl}$) an upward correction. 
So due to the corrections we have $\eta_{ijk} = (\gamma_{ijk} + \delta_{lcl}; \gamma_{ijk} + \delta_{ucl}) = \gamma_{ijk} + (\delta_{lcl};\delta_{ucl})$.
In a Poisson distribution with log-link we have $\log(\mu_{ijk}) = \eta_{ijk}$ or $\mu_{ijk} = e ^{\eta_{ijk}} = e ^{\gamma_{ijk} + (\delta_{lcl};\delta_{ucl})}$. 
The ratio $r_{ijk}$ between the upper and lower boundary of $\mu_{ijk}$ is:

$$r_{ijk} = \frac{e ^{\gamma_{ijk} + \delta_{ucl}}}{e ^{\gamma_{ijk} + \delta_{lcl})}}$$
