---
title: "Zero-inflation"
description: "Various ways in which zero-inflation can be detected and handled."
authors: [thierryo]
date: "`r Sys.Date()`"
categories: ["r", "statistics"]
tags: ["r", "analysis", "mixed model"]
output: 
  md_document:
    preserve_yaml: true
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE
)
```

## Used packages

```{r}
library(ggplot2)
library(scales)
library(pscl)
library(MASS)
library(tidyr)
library(dplyr)
library(lme4)
```



# Lots of zero's doesn't imply zero-inflation

```{r}
set.seed(1234)
n <- 1e4
n.sim <- 1e3
mean.poisson <- 0.05
mean.zeroinfl <- 10000
prop.zeroinfl <- 0.1
```


```{r}
dataset <- data.frame(
  Poisson = rpois(n, lambda = mean.poisson)
)
```


```{r}
ggplot() +
  geom_linerange(
    data = dataset %>%
      filter(Poisson == 0) %>%
      count(Poisson),
    mapping = aes(
      x = Poisson,
      ymin = 0,
      ymax = n
    )
  ) +
  geom_histogram(
    data = dataset %>%
      filter(Poisson > 0),
    mapping = aes(x = Poisson),
    boundary = 0
  )
```


```{r}
table(dataset$Poisson)
```


The example above generates values from a Poisson distribution with mean `r mean.poisson`. 
Although it has no zero-inflation, `r sprintf("%.1f%%", 100 * mean(dataset$Poisson == 0))` of the values are zero.

# Zero-inflation can also occur with low number of zero's

```{r}
dataset$ZeroInflatedPoisson <- rbinom(n, size = 1, prob = 1 - prop.zeroinfl) *
  rpois(n, lambda = mean.zeroinfl)
```


```{r message=FALSE}
ggplot() +
  geom_linerange(
    data = dataset %>%
      filter(ZeroInflatedPoisson == 0) %>%
      count(ZeroInflatedPoisson),
    mapping = aes(
      x = ZeroInflatedPoisson,
      ymin = 0,
      ymax = n
    )
  ) +
  geom_histogram(
    data = dataset %>%
      filter(ZeroInflatedPoisson > 0),
    mapping = aes(x = ZeroInflatedPoisson),
    boundary = 0
  )
```


```{r}
table(dataset$ZeroInflatedPoisson == 0)
```


The second example generates a data from a zero-inflated Poisson distribution with mean `r mean.zeroinfl` and `r sprintf("%.0f, %%", 100 * prop.zeroinfl)` excess zero's. 
The actual proportion of zero's is `r sprintf("%.1f%%", 100 * mean(dataset$ZeroInflatedPoisson == 0))`.

# How to test for zero-inflation

```{r}
dataset <- expand.grid(
  Mean = c(mean.poisson, mean.zeroinfl),
  Rep = seq_len(n)
)
dataset$Poisson <- rpois(nrow(dataset), lambda = dataset$Mean)
dataset$ZeroInflatedPoisson <- rbinom(nrow(dataset), size = 1, prob = 1 - prop.zeroinfl) *
  rpois(nrow(dataset), lambda = dataset$Mean)
```


```{r}
ggplot() +
  geom_linerange(
    data = dataset %>%
      filter(Poisson == 0) %>%
      count(Poisson),
    mapping = aes(
      x = Poisson,
      ymin = 0,
      ymax = n
    )
  ) +
  geom_histogram(
    data = dataset %>%
      filter(Poisson > 0),
    mapping = aes(x = Poisson),
    boundary = 0
  )
```


```{r}
ggplot() +
  geom_linerange(
    data = dataset %>%
      filter(ZeroInflatedPoisson == 0) %>%
      count(ZeroInflatedPoisson),
    mapping = aes(
      x = ZeroInflatedPoisson,
      ymin = 0,
      ymax = n
    )
  ) +
  geom_histogram(
    data = dataset %>%
      filter(ZeroInflatedPoisson > 0),
    mapping = aes(x = ZeroInflatedPoisson),
    boundary = 0
  )
```


For this example we generate a new dataset with two levels of Mean: `r sprintf("%.2f", c(mean.poisson, mean.zeroinfl))`. 
We generate both a Poisson and a zero-inflated Poisson variable. 
The latter has `r sprintf("%.0f, %%", 100 * prop.zeroinfl)` excess zero's. 
The proportion of zero's the variables are respectively `r sprintf("%.1f%%", 100 * mean(dataset$Poisson == 0))` and `r sprintf("%.1f%%", 100 * mean(dataset$ZeroInflatedPoisson == 0))`.

Let's assume that our hypothesis is that all observations share the same mean. 
Note that is clearly not the case. 
If this hypothesis would hold, then we fit a simple intercept-only model.

```{r}
m.pois.simple <- glm(Poisson ~ 1, data = dataset, family = poisson)
m.zero.simple <- glm(ZeroInflatedPoisson ~ 1, data = dataset, family = poisson)
summary(m.pois.simple)
summary(m.zero.simple)
```

After fitting the model we generate at random a new response variable based on the models and count the number of zero's. 
This is repeated several times so that we can get a distribion of zero's based on the model. 
Finally we compare the number of zero's in the original dataset with this distribution.

```{r}
simulated <- data.frame(
  PoissonSimple = apply(
    simulate(m.pois.simple, nsim = n.sim) == 0,
    2,
    mean
  ),
  ZeroInflatedSimple = apply(
    simulate(m.zero.simple, nsim = n.sim) == 0,
    2,
    mean
  )
)
```


```{r}
ggplot(simulated, aes(x = PoissonSimple)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = mean(dataset$Poisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
```


```{r}
ggplot(simulated, aes(x = ZeroInflatedSimple)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = mean(dataset$Poisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
```

In both cases none of the simulated values contains zero's. 
The red line indicates the observed proportion of zero's with is clearly greater. 
This indicates that we have either a poor model fit or zero-inflation.

# Improving the model

In the case of this example we have two groups with very different means: one group with low mean generating lots of zero's and one group with large mean generating no zero's. 
Adding this grouping improves the model.

```{r}
m.pois.complex <- glm(Poisson ~ factor(Mean), data = dataset, family = poisson)
m.zero.complex <- glm(ZeroInflatedPoisson ~ factor(Mean), data = dataset, family = poisson)
summary(m.pois.complex)
summary(m.zero.complex)
```

```{r}
simulated$PoissonComplex <- apply(
  simulate(m.pois.complex, nsim = n.sim) == 0,
  2,
  mean
)
simulated$ZeroInflatedComplex <- apply(
  simulate(m.zero.complex, nsim = n.sim) == 0,
  2,
  mean
)
```


```{r}
ggplot(simulated, aes(x = PoissonComplex)) +
  geom_histogram(binwidth = 0.0005) +
  geom_vline(xintercept = mean(dataset$Poisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
```


```{r}
ggplot(simulated, aes(x = ZeroInflatedComplex)) +
  geom_histogram(binwidth = 0.0005) +
  geom_vline(xintercept = mean(dataset$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
```


In case of the Poisson variable, `r sprintf("%.1f%%", 100 * mean(simulated$PoissonComplex > mean(dataset$Poisson == 0)))` of the simulated dataset from the complex model has more zero's than the observed dataset. 
So this model can captures the zero's and there is not zero-inflation. 
In case of the zero-inflated Poisson variable `r sprintf("%.1f%%", 100 * mean(simulated$ZeroInflatedComplex > mean(dataset$ZeroInflatedPoisson == 0)))` of the simulated dataset from the complex model has more zero's than the observed dataset. 
Hence we conclude that the zero's are not modelled properly. 
We can't improve the model further with covariates, hence we'll have to treat is a a zero-inflated distribution.

```{r}
m.zero.zi <- zeroinfl(
  ZeroInflatedPoisson ~ factor(Mean) | 1,
  data = dataset,
  dist = "poisson"
)
summary(m.zero.zi)
sim.zero.zi <- function(model) {
  eta <- predict(model, type = "count")
  prob <- predict(model, type = "zero")
  new.value <-
    rbinom(length(eta), size = 1, prob = 1 - prob) *
      rpois(length(eta), lambda = eta)
  mean(new.value == 0)
}
simulated$ZeroInflatedZI <- replicate(n.sim, sim.zero.zi(m.zero.zi))
ggplot(simulated, aes(x = ZeroInflatedZI)) +
  geom_histogram(binwidth = 0.0005) +
  geom_vline(xintercept = mean(dataset$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
```

When we fit a proper zero-inflated model to the zero-inflated Poisson variable, then `r sprintf("%.1f%%", 100 * mean(simulated$ZeroInflatedZI > mean(dataset$ZeroInflatedPoisson == 0)))` of the simulated datasets have more zero's than the observed dataset. 
So the zero-inflation is properly handled.

# Modelling zero-inflation by overdispersion

```{r}
m.zero.nb <- glm.nb(ZeroInflatedPoisson ~ factor(Mean), data = dataset)
summary(m.zero.nb)
dataset$ID <- seq_along(dataset$Mean)
m.zero.olre <- glmer(
  ZeroInflatedPoisson ~ factor(Mean) + (1 | ID),
  data = dataset,
  family = poisson
)
summary(m.zero.olre)
```

```{r}
simulated$ZeroInflatedNB <- apply(
  simulate(m.zero.nb, nsim = n.sim) == 0,
  2,
  mean
)
```


```{r}
ggplot(simulated, aes(x = ZeroInflatedNB)) +
  geom_histogram(binwidth = 0.0005) +
  geom_vline(xintercept = mean(dataset$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
simulated$ZeroInflatedOLRE <- apply(
  simulate(m.zero.olre, nsim = n.sim) == 0,
  2,
  mean
)
```


```{r}
ggplot(simulated, aes(x = ZeroInflatedOLRE)) +
  geom_histogram(binwidth = 0.0005) +
  geom_vline(xintercept = mean(dataset$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
```

```{r}
mean.2 <- 5
dataset2 <- expand.grid(
  Mean = mean.2,
  Rep = seq_len(100)
)
dataset2$ZeroInflatedPoisson <-
  rbinom(nrow(dataset2), size = 1, prob = 1 - prop.zeroinfl) *
    rpois(nrow(dataset2), lambda = dataset2$Mean)
```


```{r}
ggplot(dataset2, aes(x = ZeroInflatedPoisson)) +
  geom_histogram(binwidth = 1)
m.zero2 <- glm(ZeroInflatedPoisson ~ 1, data = dataset2, family = poisson)
summary(m.zero2)
simulated$ZeroInflated2 <- apply(
  simulate(m.zero2, nsim = n.sim) == 0,
  2,
  mean
)
```


```{r}
ggplot(simulated, aes(x = ZeroInflated2)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = mean(dataset2$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
m.zero.nb2 <- glm.nb(ZeroInflatedPoisson ~ 1, data = dataset2)
summary(m.zero.nb2)
simulated$ZeroInflatedNB2 <- apply(
  simulate(m.zero.nb2, nsim = n.sim) == 0,
  2,
  mean
)
```


```{r}
ggplot(simulated, aes(x = ZeroInflatedNB2)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = mean(dataset2$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
dataset2$ID <- seq_along(dataset2$Mean)
m.zero.olre2 <- glmer(
  ZeroInflatedPoisson ~ (1 | ID),
  data = dataset2,
  family = poisson
)
```


```{r}
simulated$ZeroInflatedOLRE2 <- apply(
  simulate(m.zero.olre2, nsim = n.sim) == 0,
  2,
  mean
)
```


```{r}
ggplot(simulated, aes(x = ZeroInflatedOLRE2)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = mean(dataset2$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
m.zero.zi2 <- zeroinfl(
  ZeroInflatedPoisson ~ 1,
  data = dataset2,
  dist = "poisson"
)
```


```{r}
summary(m.zero.zi2)
```


```{r}
simulated$ZeroInflatedZI2 <- replicate(n.sim, sim.zero.zi(m.zero.zi2))
ggplot(simulated, aes(x = ZeroInflatedZI2)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept = mean(dataset2$ZeroInflatedPoisson == 0), colour = "red") +
  scale_x_continuous(label = percent)
```


A negative binomial distribution can capture some of the zero-inflation by overdispersion. 
Especially in cases with low mean and few data. 
When the mean is low, a reasonable portion of the zero origination comes from the poisson distribution. 
However, the shape of the distribution is different.

```{r}
distribution <- data.frame(
  Count = 0:20
)
distribution$Truth <- dpois(
  distribution$Count,
  lambda = mean.2
) + ifelse(
  distribution$Count == 0,
  prop.zeroinfl,
  0
)
distribution$Poisson <- dpois(
  distribution$Count,
  lambda = exp(coef(m.zero2))
)
distribution$ZIPoisson <- dpois(
  distribution$Count,
  lambda = exp(coef(m.zero.zi2)[1])
) + ifelse(
  distribution$Count == 0,
  plogis(coef(m.zero.zi2)[2]),
  0
)

se <- sqrt(VarCorr(m.zero.olre2)$ID)
z <- seq(qnorm(.001, sd = se), qnorm(.999, sd = se), length = 101)
dz <- dnorm(z, sd = se)
dz <- dz / sum(dz)
delta <- outer(
  distribution$Count,
  exp(z + fixef(m.zero.olre2)),
  FUN = dpois
)
distribution$OLRE <- delta %*% dz

distribution$NegBin <- dnbinom(
  distribution$Count,
  size = m.zero.nb2$theta,
  mu = exp(coef(m.zero.nb2))
)
long <- gather(distribution, key = "Model", value = "Density", -Count)
long$Truth <- long$Model == "Truth"
```


```{r}
ggplot(long, aes(x = Count, y = Density, colour = Model, linetype = Truth)) +
  geom_line()
```


