---
title: "Using WFS service in R"
description: "How to use WFS (vectors/features) GIS services within R scripts"
authors: [thierryo, hansvancalster, florisvdh]
date: "`r Sys.Date()`"
categories: ["r"]
tags: ["gis", "webservice", "r", "maps"]
bibliography: "../../articles/reproducible_research.bib"
output: 
    md_document:
        preserve_yaml: true
        variant: gfm
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
  cache = TRUE,
  autodep = TRUE
)
```

# What is WFS?

In computing, the [Open Geospatial Consortium (OGC) **Web Feature Service (WFS)**](http://www.opengeospatial.org/standards/wfs) Interface Standard provides an interface allowing requests for geographical features across the web using platform-independent calls. 
One can think of geographical features as the "source code" behind a map, whereas the Web Mapping Service (WMS) interface or online tiled mapping portals like Google Maps return only an image, which end-users cannot edit or spatially analyze. 

The standard output format for reading or downloading data from a WFS is the XML-based Geography Markup Language (GML), but other formats like shapefiles or geojson are also available. 
In early 2006 the OGC members approved the OpenGIS GML Simple Features Profile. 
This profile is designed both to increase interoperability between WFS servers and to improve the ease of implementation of the WFS standard. 
(Source: [Wikipedia](https://en.wikipedia.org/wiki/Web_Feature_Service))

Before we embark on using a WFS service in R, we would like to draw your attention to the following issue when using a WFS service in a workflow that needs to be reproducible in the longer term. 
A distributed, stable storage of the GIS data used in a workflow is necessary to make that workflow reproducible on a longer term. 
When using dynamic databases and web services, where the data they present are expected to evolve (e.g. by always pointing at the latest release in these WFS-cases), this is not guaranteed if you just use the WFS service to directly read GIS data from it. 
Instead of directly reading the GIS data, it is also possible to download the data from the WFS service to disk (and possibly archive on, e.g., Zenodo) in order to obtain a stable version that can be re-used later by reading it from disk.
Also, in case it takes a long time to get the GIS data from the WFS service, downloading instead of directly reading is often a better choice.
On the other hand, some types of workflows might instead require the most up to date GIS data that are available to be retrieved while no workflow reproducibility is needed.
Do note that publishing the workflow is always recommended for published work.
It serves the scientific community and it is also a way of proving scientific quality - hence reproducibility is often the best choice for published work.

Some of the material presented in this tutorial benefitted from a tutorial presented at the Use of R in Official Statistics conference in The Hague, September 2018 [Spatial Analysis in R with Open Geodata](https://github.com/TWIAV/Spatial_Analysis_in_R_with_Open_Geodata/), from @lovelace_geocomputation_2020 and from [this Dutch wordpress blogpost](https://friesewoudloper.wordpress.com/2015/05/17/het-bevragen-van-een-wfs-in-de-browser-voorbeelden-aan-de-hand-van-de-bag-service/).

# Useful overviews of web feature services

WFS services for Belgium and regions in Belgium:

- [overview of WFS services for Flanders region](https://overheid.vlaanderen.be/Webdiensten-Overdrachtdiensten)
- [overview compiled by Michel Stuyts](https://wfs.michelstuyts.be/), which is [also available on gitlab](https://gitlab.com/GIS-projects/Belgium-WFS/)
- [overview maintained by DOV Vlaanderen](https://dov.vlaanderen.be/page/interessante-webservices)

European portals:

- [inspire geoportal](https://inspire-geoportal.ec.europa.eu/): European portal for spatial data - some of which have a WFS service
- [environmental data for Europe](https://discomap.eea.europa.eu/Index/): many of the products listed have either a WMS or a WFS service 

Worldwide coverage:

- [spatineo directory](https://directory.spatineo.com/)


# Used packages

```{r}
library(sf) # simple features packages for handling vector GIS data
library(httr) # generic webservice package
library(tidyverse) # a suite of packages for data wrangling, transformation, plotting, ...
library(ows4R) # interface for OGC webservices

```

# Get to know what the service can do with `GetCapabilities`

First of all we need the URL of the service. 

```{r}
wfs_bwk <- "https://geoservices.informatievlaanderen.be/overdrachtdiensten/BWK/wfs"
```

Next, we append information to the URL address with the aid of `httr::parse_url` and `httr::build_url`.
The former function parses an URL into a list for easier programmatic addition of information to the URL.
The latter function does the reverse and builds the URL from the list object.

The `url$query` slot is where you instruct the WFS what information it should return.
It is constructed as a list with name-value pairs.
For now, we only need to specify the `GetCapabilities` request.
Other information such as passing `version = 2.0.0` can be added, but is not required (by default, the latest version of the WFS service will be chosen).

```{r}
url <- parse_url(wfs_bwk)
url$query <- list(service = "wfs",
                  #version = "2.0.0", # facultative
                  request = "GetCapabilities"
                  )
request <- build_url(url)
request
```

With `GetCapabilities`, we obtain a complete overview of all metadata for the web service.

To see all capabilities, you can visit [the request in the webbrowser](`r request`). 
For instance opening the page in the webbrowser and searching for "Filter_Capabilities" allows you to see all possible ways to filter the data from a WFS layer (e.g. restrict the downloaded data to a specified bounding box with `SpatialOperator name="BBOX"`).

Instead of searching the page on the web, there are several ways to access specific pieces of information programmatically. 
We will show here how to do this using functions in the `ows4R` package. 
The first thing we need to do is generate a connection to the WFS with the aid of `WFSClient$new()`.


```{r}
bwk_client <- WFSClient$new(wfs_bwk, 
                            serviceVersion = "2.0.0") #serviceVersion must be provided here
```

The resulting object `bwk_client` is an R6 object. 
If you are not familiar with R6 object, you might want to read [the R6 chapter in Advanced R](https://adv-r.hadley.nz/r6.html).

Printing ``bwk_client` looks like this:

```{r}
bwk_client
```

The features listed can be accessed using `$`.
We can see that some of them are again functions.
As a first example, the following code will list all available layers for that WFS.


```{r}
bwk_client$getFeatureTypes(pretty = TRUE)
```

Another way of accessing this information is:

```{r}
bwk_client$getFeatureTypes() %>%
  map_chr(function(x){x$getName()})

bwk_client$getFeatureTypes() %>%
  map_chr(function(x){x$getTitle()})
```

We are using `purrr::map()` functionalities here, because the output of `bwk_client$getFeatureTypes()` is a list and each element of the list contains information about a layer.


Let's turn back our attention to `getCapabilities()` and see what information we can extract from it.

```{r}
bwk_client$getCapabilities()
```

This is again an R6 class object and the `$` can be used to chain together several functions, much in the  same way as the pipe operator `%>%`. 
The following chunk illustrates its use. 
Try executing this incrementally (select and execute code up to first `$`, second `$`, ...) to see what happens.

```{r}
bwk_client$
  getCapabilities()$
  findFeatureTypeByName("BWK:Bwkhab")$
  getDescription() %>%
  map_chr(function(x){x$getName()})

# or shorter
bwk_client$
  describeFeatureType(typeName = "BWK:Bwkhab") %>%
  map_chr(function(x){x$getName()})
  
```

This lists all available fields for the layer "BWK:Bwkhab".

Here is how to get a character vector naming all available operations of the WFS:


```{r}
bwk_client$
  getCapabilities()$
  getOperationsMetadata()$
  getOperations() %>%
  map_chr(function(x){x$getName()})
```


The next chunk shows how we can extract the available output formats.
We will see later that `GetFeature` is the operation needed to read or download data from the WFS. 
The metadata for this operation has what we want and we can extract it with a combination of `purrr::map()` and ``purrr::pluck()`.

```{r}
bwk_client$
  getCapabilities()$
  getOperationsMetadata()$
  getOperations() %>%
  map(function(x){x$getParameters()}) %>%
  pluck(3, "outputFormat")
```

<!--
Another interesting piece of metadata we might be interested in is which spatial operations the WFS service is able to do. 
-->



Some more examples follow.
Let's extract the bounding boxes for all layers. 

```{r}
bwk_client$
  getCapabilities()$ 
  getFeatureTypes() %>%  
  map(function(x){x$getBoundingBox()})
```

As expected for this WFS, the bounding boxes are the same for all layers.

The final example shows how to get the abstract so we can read about the contents of the layers.

```{r}
bwk_client$
  getCapabilities()$ 
  getFeatureTypes() %>%  
  map_chr(function(x){x$getAbstract()})
```



# Read or download vector data from WFS: `GetFeature` 

## Example 1: an entire layer

The map of regions of Belgium.

```{r warning=FALSE}
wfs_regions <- "https://eservices.minfin.fgov.be/arcgis/services/R2C/Regions/MapServer/WFSServer"
regions_client <- WFSClient$new(wfs_regions, 
                            serviceVersion = "2.0.0")
regions_client$getFeatureTypes(pretty = TRUE)
```


```{r}
url <- parse_url(wfs_regions)
url$query <- list(service = "wfs",
                  #version = "2.0.0", # optional
                  request = "GetFeature",
                  typename = "regions",
                  srsName = "EPSG:4326"
                  )
request <- build_url(url)

bel_regions <- read_sf(request) #Lambert2008
```

```{r}
ggplot(bel_regions) +
  geom_sf()
```

## Example 2: filter by attributes

In this example, we only want to extract specific regions.
This can be done using either [standard OGC filter specification](https://www.ogc.org/standards/filter) or using a [Contextual Query Language (CQL) filter](https://www.loc.gov/standards/sru/cql/index.html),
for which a didactical explanation can be found [here](https://gcs-docs.s3.amazonaws.com/EVWHS/Miscellaneous/DevGuides/WFS/WFS_Query.htm).^[Note that CQL was formerly called Common Query Language.]
The latter, however, only works for WFS services that are hosted on a GeoServer!

In this example we also show how the previously used R code can be stitched together using the pipe (`%>%`) operator.


**Standard OGC filter**

Unfortunately, the standard OGC filter format is very verbose...

```{r}
wfs_regions %>%
  parse_url() %>%
  list_merge(query = list(service = "wfs",
                          #version = "2.0.0", # optional
                          request = "GetFeature",
                          typename = "regions",
                          srsName = "EPSG:4326",
                          filter = "<Filter><PropertyIsEqualTo><PropertyName>regions:NameDUT</PropertyName><Literal>'Vlaams Gewest'</Literal></PropertyIsEqualTo></Filter>")) %>%
  build_url() %>%
  read_sf() %>%
  ggplot() +
  geom_sf()
```


**CQL filter**

We use a different WFS service for which CQL works.
First we need to know the names of the fields by which we can filter.


```{r}
wfs_vrbg <- "https://geoservices.informatievlaanderen.be/overdrachtdiensten/VRBG/wfs"

vrbg_client <- WFSClient$new(wfs_vrbg, 
                            serviceVersion = "1.1.0")

vrbg_client$
  getCapabilities()$
  findFeatureTypeByName("VRBG:Refprv")$
  getDescription() %>%
  map_chr(function(x){x$getName()})

# another way of doing this:
wfs_vrbg %>%
  parse_url() %>%
  list_merge(query = list(service = "wfs",
                          #version = "1.1.0", # optional
                          request = "DescribeFeatureType",
                          typeName = "VRBG:Refprv")) %>%
  build_url() %>%
  GET()
```

The CQL filter format is much more human readable and easier to code:

```{r}
sf_prov <- wfs_vrbg %>% 
  parse_url() %>% 
  list_merge(query = list(service = "wfs",
                          #version = "1.1.0", # optional
                          request = "GetFeature",
                          typeName = "VRBG:Refprv",
                          srsName = "EPSG:31370",
                          cql_filter="NAAM='West-Vlaanderen'")) %>% 
  build_url() %>% 
  read_sf(crs = 31370)

sf_prov
```

Also check out [example 5](#example-5-extract-feature-data-at-particular-points) for a more advanced use of the CQL filter.

Note, the rather exotic geometry type that is returned (`MULTISURFACE`).
Some `sf` functions, such as `st_buffer()`, will not work out of the box for this type.
In this specific case, we need an intermediate step `st_cast(to = "GEOMETRYCOLLECTION")` to make it work.

```{r error=TRUE}
sf_prov %>% 
  st_buffer(dist = 100) # errors

sf_prov_buffer <- sf_prov %>% 
  st_cast(to = "GEOMETRYCOLLECTION") %>% 
  st_buffer(dist = 10000) # works

sf_prov %>% 
  ggplot() + 
  geom_sf(data = sf_prov_buffer) +
  geom_sf()
```

If you need to further convert to geometry type `POLYGON`, then use `st_collection_extract()` to extract the Polygon from the GeometryCollection:

```{r paged.print=FALSE}
sf_prov %>% 
  st_cast(to = "GEOMETRYCOLLECTION") %>% 
  st_collection_extract(type = "POLYGON")
```

## Example 3: restrict to a bounding box

This examples illustrates how you can read or download information from a WFS for further use in R. 

Let's start by reading in the "BWK:Bwkhab" layer for the Hallerbos area. 

The main part is defining the input and output. 
We store the data in [GeoJSON](https://en.wikipedia.org/wiki/GeoJSON) format which is an open standard format designed for representing simple geographical features, along with their non-spatial attributes. 
It is based on JSON, the JavaScript Object Notation.

We also add the bounding box from which we want to retrieve the data. 
This is very important to add. 
If you omit a bounding box, the service will return the entire map which can be very large.


```{r}
url <- parse_url(wfs_bwk)
url$query <- list(service = "WFS",
                  #version = "2.0.0", # optional
                  request = "GetFeature",
                  typename = "BWK:Bwkhab",
                  bbox = "142600,153800,146000,156900")
request <- build_url(url)
```

Let's check what we are about to read with `sf::st_layers()`.
This time, the function does return useful information, but the layer name BWK:Bwkhab seems to be overwritten.

```{r}
st_layers(request)
```


```{r}
bwk_hallerbos <- read_sf(request)
```


Note that the `layer = ...` argument of `sf:read_sf()` cannot be used to specify the layer when you pass a WFS GetFeature request to the function.


Let's make a simple plot of the object. 
Note that the object contains features outside of the bounding box. 
Those are features which have only some part within the bounding box.

```{r}
ggplot(bwk_hallerbos) + 
  geom_sf()
```

You can use `sf::st_write()` to save this layer in any format that is listed by `sf::st_drivers()`.

## Example 4: downloading data to disk

Continuing from the previous example, we can download the data with `httr::GET` and `httr::write_disk()`.
The only difference is that we add `outputFormat` to the request.
In this way we can deviate from the default output format and choose the output format we like.

```{r}
url <- parse_url(wfs_bwk)
url$query <- list(service = "WFS",
                  #version = "2.0.0", # optional
                  request = "GetFeature",
                  typename = "BWK:Bwkhab",
                  bbox = "142600,153800,146000,156900",
                  outputFormat = "application/json")
request <- build_url(url)
```


```{r}
file <- tempfile(fileext = ".geojson")
GET(url = request, 
    write_disk(file))
```

At this point, all features are downloaded and can be used in R as we would with any other local file. 
So we need to load the file with `read_sf()` from `sf`.

```{r}
bwk_hallerbos2 <- read_sf(file)
```

Downloading data could be important for reproducible workflows (e.g. by using the downloaded data as the 'real' starting point and distribute it).

## Example 5: extract feature data at particular points

In some situations, we do not need the spatial features (polygons, lines, points), but are interested in the data at a particular point (i.e. attribute table data) of the spatial feature. 
When working in a local GIS environment, one would use a spatial operator to extract the data (e.g. `within`, `intersects`, `contains`,...). 
As we discussed earlier, WFS supports certain spatial operators as part of the service to directly query this data and overcomes the need to download the spatial feature data first.


Consider the following use case: You want to extract the attribute data from a [soil map](http://www.geopunt.be/catalogus/datasetfolder/5c129f2d-4498-4bc3-8860-01cb2d513f8f) for a number of sampling points (point coordinates). 
This use case can be tackled by relying on the WFS service and the affiliated spatial operators.

Our example data point (in Lambert 72):

```{r}
x_lam <- 173995.67
y_lam <- 212093.44
```

From this point we know the data, so we can verify the result (in Dutch):

* Bodemtype:	s-Pgp3(v)
* Bodemserie:	Pgp
* Textuurklasse:	licht zandleem
* Drainageklasse:	uiterst nat, gereduceerd

Hence, we now want to extract these soil properties from the WFS, for the coordinates defined above. 
```{r}
properties_of_interest <- c("Drainageklasse",
                            "Textuurklasse",
                            "Bodemserie",
                            "Bodemtype")
```

The URL of the wfs service of the soil map of the Flemish region:
```{r}
wfs_bodemtypes <- "https://www.dov.vlaanderen.be/geoserver/bodemkaart/bodemtypes/wfs"
```

The essential part is to set up the proper query! 
The required data for the service is defined in the [metadata](https://www.dov.vlaanderen.be/geoserver/bodemkaart/bodemtypes/wfs?version=1.1.0&request=GetCapabilities&service=wfs) description. 
This can look a bit overwhelming at the start, but is a matter of looking for some specific elements of the (XML) document:

* `service` (WFS) and `request` (GetFeature) are mandatory fields (see below);  `version` (1.1.0) is optional
* `typeName`: Look at the different `<FeatureType...` enlisted and pick the `<Name>` of the one you're interested in. In this particular case `bodemkaart:bodemtypes` is the only one available.
* `outputFormat`: The supported output formats are enlisted in `<ows:Parameter name="outputFormat">`. As the service provides CSV as output, this is a straightforward option. `json` is another popular one.
* `propertyname`: A list of the attribute table fields (cfr. supra). A full list of the Flanders soil map is provided [here](https://www.dov.vlaanderen.be/geoserver/bodemkaart/bodemtypes/wfs?request=DescribeFeatureType).
* We also define the `CRS`, using the [EPSG code](http://spatialreference.org/).
* `CQL_FILTER`: Define the spatial operator, in this case `INTERSECTS` of the WFS `geom` and our `POINT` coordinate. The operators are enlisted in the `<ogc:SpatialOperators>` field.

Formatting all this information in a query and executing the request (`GET`) towards the service:

```{r}
query <- list(service = "WFS",
             request = "GetFeature",
             #version = "1.1.0", # optional
             typeName = "bodemkaart:bodemtypes",
             outputFormat = "csv",
             propertyname = as.character(paste(properties_of_interest,
                                               collapse = ",")),
             CRS = "EPSG:31370",
             CQL_FILTER = sprintf("INTERSECTS(geom,POINT(%s %s))",
                                  x_lam, y_lam)) 
result <- GET(wfs_bodemtypes, query = query)
result
```

The result is not yet formatted to be used as a dataframe. 
We need to use a small *trick* using the `textConnection` function to get from the result (bits) towards a readable output in a dataframe:

```{r}
df <- read.csv(textConnection(content(result, 'text')))
knitr::kable(df)
```

Which indeed corresponds to the data of the coordinate.

In order to make your coding work lighter, you can easily put the WFS R code (building the WFS request and retrieving its results) inside a function.
Just use [the code](https://github.com/inbo/inborutils/blob/8dad1ebe25ba80a3239d4a1ab9e918cf862b621e/R/extract_soil_map_data.R#L49-L91) of the `inborutils::extract_soil_map_data()` function (documented [here](https://inbo.github.io/inborutils/reference/extract_soil_map_data.html)) as a template for your own function.

Now this function makes it really easy to request the data:

```{r}
library(inborutils)
extract_soil_map_data(
  x_lam = x_lam, 
  y_lam = y_lam) %>%
  knitr::kable()

extract_soil_map_data(
  x_lam = x_lam, 
  y_lam = y_lam,
  properties_of_interest = c("Eenduidige_legende", "Textuurklasse")) %>%
  knitr::kable()

extract_soil_map_data(
  x_lam = x_lam, 
  y_lam = y_lam,
  properties_of_interest = properties_of_interest) %>%
  knitr::kable()

# multiple point locations
xy <- data.frame(id = c("loc1", "loc2"),
                 x = c(173995.67, 180000),
                 y = c(212093.4, 212000))

xy %>%
  group_by(id) %>%
  summarise(extract_soil_map_data(x, y, 
                                  properties_of_interest = properties_of_interest)) %>%
  knitr::kable()
```


## Example 6: pagination


When a layer of a WFS service contains many thousands of features, often the WFS service will restrict the number of features that can be obtained with one request.
This is done to optimize performance (responsiveness) of the WFS service: it is more efficient to transfer small packages of data and this is even more the case when multiple user requests are transmitted concurrently to the service.

In this example we show how to deal with this situation.
A technique called 'pagination' can be used to obtain all features one is interested in by sending multiple requests to the server.
The service is available from version '2.0.0' onwards.
Servers running older versions may or may not have support for 'pagination'.

We will use the 'Watervlakken' WFS service for this example.

```{r}
# this layer returns just 1000 features (for performance reasons)
wfs <- "https://gisservices.inbo.be/arcgis/services/Watervlakken/MapServer/WFSServer?"
url <- parse_url(wfs)
url$query <- list(service = "wfs",
                  version = "2.0.0", # facultative
                  request = "GetFeature",
                  typename = "Watervlakken:Watervlakken",
                  srsName = "EPSG:31370"
)
request <- build_url(url)
gd <- read_sf(request)
gd
```

As we can see from the output, only 1000 features are returned.

To know in advance how many features the WFS has in total, you can add `resultType="hits"` to the query:


```{r}
url$query <- list(service = "wfs",
                  version = "2.0.0", # facultative
                  request = "GetFeature",
                  typename = "Watervlakken:Watervlakken",
                  srsName = "EPSG:31370",
                  resultType = "hits"
)
request <- build_url(url)
result <- GET(request)
parsed <- xml2::as_list(content(result, "parsed"))
n_features <- attr(parsed$FeatureCollection, "numberMatched")
n_features
```

We see that there are `r n_features` features.
In the next chunk we show how `startIndex` and `count` can be used together to obtain more features (3000 in the example).

```{r}
# use count and startIndex to return more features (pagination)
get_watervlakken <- function(index = 0) {
  wfs <- "https://gisservices.inbo.be/arcgis/services/Watervlakken/MapServer/WFSServer?"
  request <- build_url(url)
  request # url in browser bekeken en nodige info in opgezocht
  url$query <- list(service = "wfs",
                    version = "2.0.0", # facultative
                    request = "GetFeature",
                    typename = "Watervlakken:Watervlakken",
                    srsName = "EPSG:31370",
                    startIndex=index,
                    count=1000)
  request <- build_url(url)
  gd <- read_sf(request)
  return(gd)
}

gd <- map_dfr(.x = seq(0, 2000, 1000),
              .f = get_watervlakken
        )

gd
```

Whether it is a good idea or not to request many thousands of features from a WFS really depends on what our further plans are with the spatial data.

Do we really need the attribute data?
If not, and the only goal is visualization, we should have used a WMS (Web Mapping Service) instead (see [this tutorial](https://inbo.github.io/tutorials/tutorials/spatial_wms_services/)).

Do we need all data from the WFS?
We need to think carefully which features and which attributes are really needed for whatever use case we might have.
When we have a clear idea which subset of the data we need, we can apply what we have learned in the examples in this tutorial to restrict the request to what we need.

If we find ourselves in the case where we think we really need to request many, many features from a WFS service, it may also be more advised to use a download service instead of a WFS (search) service.



# References
