---
title: "Git and GitHub Authentication on Windows"
description: "How to safely identify yourself, or a third party on behalf of you, to Git or GitHub"
authors: [hansvancalster]
date: "`r Sys.Date()`"
categories: ["version control"]
tags: ["git", "github", "rstudio"]
output: 
  md_document:
    preserve_yaml: true
    variant: gfm+footnotes+definition_lists
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


## Before we start

This tutorial assumes you have a GitHub account, and have Git, R and RStudio installed on your system.
The purpose of this tutorial is to make clear recommendations about authentication for employees at INBO who use Git version control in their workflows. 

If you do not have a GitHub account, you can [sign up for a new GitHub account](https://docs.github.com/en/get-started/signing-up-for-github/signing-up-for-a-new-github-account).
The installation of Git, R and RStudio at INBO, is done by IT administrators who follow [these installation instructions for admins](../../installation/administrator/).
In addition, the useR is asked to do some extra steps after installation of [Git](../../installation/user/user_install_git/), [R](../../installation/user/user_install_r/) and [RStudio](../../installation/user/user_install_rstudio/).

To authenticate means passing information (= credentials) that proves that you are exactly who you declare to be.
Authentication is about securely accessing your GitHub account's resources either directly by yourself or on behalf of you by a third party.
For instance, you can grant an R package the right to do stuff on your behalf (e.g. reading your e-mail address).
You determine what can and what cannot be done and this can be different for various third parties.

Most of the material in this tutorial is explained in length in https://happygitwithr.com/connect-intro.html.
We will mainly follow [the advice given in the usethis package](https://usethis.r-lib.org/articles/articles/git-credentials.html).

Authentication is a quite confusing and technical topic.
In the [TL;DR section](#tldr) (Too Long; Didn't read) we just provide a summary of the recipe to follow without any explanation.
Using this recipe will hopefully just work.
If you want to get some intuition about what you are doing, you will need to read the remainder of the tutorial too (and also check out the weblinks included in this tutorial).

## TL;DR

First, enable two-factor authentication (2FA) for your GitHub account: follow [these steps](https://docs.github.com/en/authentication/securing-your-account-with-two-factor-authentication-2fa/configuring-two-factor-authentication#configuring-two-factor-authentication-using-a-totp-mobile-app).
You need a time-based one-time password app for 2FA to work.
We recommend a mobile app which you can download on your smartphone such as `andOTP` (open source, Android) or `Google Authenticator` (closed source, Android or iOS).

Next, in R, install the `usethis` package (this will also install packages `gert` and `gitcreds` which we use below) and the `checklist` package:

```{r}
install.packages("usethis")
install.packages("checklist", repos = "https://inbo.r-universe.dev")
```

Use the following commands to configure some global git options (if you haven't done this already):

```{r}
View(gert::git_config_global()) # check if values are already set or not

if (checklist::yesno(
  paste0("Are you sure you want to execute this code?\n",
         "This action will overwrite Git global configuration settings.\n",
         "Make sure you changed 'your name' and 'your@email.com' in the code,\n",
         "if you didn't select negative answer."))) {
  usethis::use_git_config(
    scope = "user",
    user.name = "Your Name",
    user.email = "your@email.com",
    init.defaultBranch = "main"
  ) 
} else {
    message("Action aborted")
}
```

If you plan to use functions that create something on GitHub (a repo, an issue, a pull request, a branch, ...) and for all remote operations from the command line (`git pull`, `git push`, `git clone`, `git fetch`), you can use the following commands to create a Personal Access Token (PAT) and add it to the Git Credential Manager: 

```{r}
?usethis::create_github_token() # read the help file
usethis::create_github_token() #browser opens, follow instructions
```

Add the PAT to the Git Credential Manager:

```{r}
?gitcreds::gitcreds_set() # read the help file
gitcreds::gitcreds_set() #paste PAT
```

**WARNING**:  handle your Personal Access Token (PAT) as a secret.
Anyone who has your token, has access to your GitHub account.

Check if everything is OK:

```{r}
usethis::git_sitrep()
```

The default Git protocol should be 'https' (recommended on Windows).

The `usethis` package promotes the use of a global `.gitignore` file which prevents that some file types that could contain sensitive information (from your account / credentials) are tracked by the Git version control system.
However, we prefer a [project-specific `.gitignore` file](https://github.com/inbo/checklist/blob/main/inst/generic_template/gitignore) for this purpose.
This is one of the many things that the `checklist` package will take care off for you.
It is therefore good practice to use the `checklist` package to set up your RStudio projects for either R packages or regular R code projects.
We refer to the [`checklist` package documentation](https://inbo.github.io/checklist/) for further information.


## Modes of authentication

The way in which you need to authenticate depends on how you access your resources in GitHub.
For instance, to manage your repositories online you can sign-in to GitHub using a username and password followed by two-factor authentication.
As another example, a function like `checklist::new_branch()` will create a new branch on GitHub and locally.
In order to create a new branch, you will need a Personal Access Token with appropriate scopes if you use the HTTPS protocol.
A token with no assigned scopes can only access public information.

We will follow the [recommendations given in the `usethis` package](https://usethis.r-lib.org/articles/articles/git-credentials.html) for safe Git authentication on Windows, which cover multiple facets:

1. Turn on two-factor authentication for your GitHub account.
1. Adopt HTTPS as your Git transport protocol.
1. Use a personal access token (PAT) for all Git remote operations from the command line or from R.
1. Allow tools to store and retrieve your credentials from the Git Credential Manager.

This deserves some further explanation.
The first recommendation essentially is an extra layer of security compared to a simple username-and-password authentication on GitHub.
It is unrelated to git operations.
To enable two-factor authentication for your GitHub account, follow [these steps](https://docs.github.com/en/authentication/securing-your-account-with-two-factor-authentication-2fa/configuring-two-factor-authentication#configuring-two-factor-authentication-using-a-totp-mobile-app).
You need a time-based one-time password app for 2FA to work.
We recommend a mobile app which you can download on your smartphone such as `andOTP` (open source, Android) or `Google Authenticator` (closed source, Android or iOS).
The closed tools don't require users 'to read on' and make everything simple, while open tools will require some minimal responsibility e.g. to take care of personal backups.
Some open source desktop applications are available as well and are listed in https://github.com/andOTP/andOTP/wiki/Open-Source-2FA-Apps.
Mostly these open source tools are geared towards offline storage and give users maximum control over their credentials.
2FA is not that intrusive.
It only kicks in when you login from a new device or when your last login on a device was a long time ago.
If someone steals your username and password, they still can't login using that combination from their device due to 2FA.

The second recommendation is to use the [HTTPS protocol](#using-the-https-protocol) to transport information (data) from your local Git repositories to their remote (online) counterparts.
A good alternative is the SSH protocol.
We have also included in this tutorial how to use the [SSH protocol](#creating-public-private-ssh-key-pairs), but in our experience this is not necessary in most circumstances and more involved to implement in Windows.
Users of the Linux operating system, on the other hand, may prefer the SSH protocol.
It is always possible to switch between HTTPS and SSH protocols and we explain in [the SSH section](#creating-public-private-ssh-key-pairs) how to do that.

The third recommendation is mandatory when you use the HTTPS protocol.
GitHub no longer allows a simple password for Git remote operations.
The remote is the place where you store your code on a Git Server, for instance https://github.com/inbo/tutorials.
Git remote operations include, among others, `git clone`, `git fetch`, `git pull` and `git push`.

The fourth recommendation is the easiest, because it just works out of the box with recent versions of Git (2.29 or higher).
The installation of Git on Windows comes with a [Git Credential Manager](https://github.com/GitCredentialManager/git-credential-manager) which allows that web apps like [hackmd](https://hackmd.io/login) or R packages like `gert` and `gh` can ask for your credentials and get them from the credential store.
In earlier days, it was sometimes needed to store your PAT in a `.Renviron` text file which you store only locally (the file will in that case contain a line `GITHUB_PAT=`...).
This is no longer needed and should be avoided because of the risk to expose the PAT when that file is accidentally put online.
You can check if you have previously set a `GITHUB_PAT` in your `.Renviron` with the R command `usethis::edit_r_environ()`.

## Creating Personal Access Tokens (PAT)

We move the steps about creating a PAT upfront, because this includes guidelines to store the PAT with the Git Credential Manager so they can be discovered automatically.

Personal access tokens can only be used for HTTPS Git operations.
When you create a PAT, you will be asked which scopes should be allowed (i.e. what rights do you give).
We suggest that you stick with the scopes recommended by the `usethis` package.
Executing the following command takes your browser to a pre-filled form with the recommended scopes (repo, user, workflow) where you can create your PAT.

```{r}
usethis::create_github_token()
```

After you have done this, you can store this PAT in the Git Credential Manager using this code:

```{r}
gitcreds::gitcreds_set()
```

which will open a prompt where you can paste your PAT (or replace an old one).

If you want to know more about PATs read [this section from happy git with R](https://happygitwithr.com/https-pat.html#get-a-pat) and [this section from the usethis package](https://usethis.r-lib.org/articles/git-credentials.html#get-a-personal-access-token-pat).
In case your PAT is stolen, you should deactivate it ASAP on GitHub.

After you've done this, you can check your authentication settings.
The `usethis` package has a function to get a situation report on your current Git/GitHub status, including information about authentication:

```{r}
usethis::git_sitrep()
```


The situation report should normally report that the default Git protocol is `https`, which we will discuss next.

## Using the HTTPS protocol

The HTTPS protocol is the GitHub default URL transport protocol that uses the Transport Layer Security (TLP) encryption protocol to encrypt communications.
To learn more about what https is, [read this explanation](https://www.cloudflare.com/learning/ssl/what-is-https/).

The usual way to choose the protocol is when you clone a repository to your computer.
You can do this in various ways:

- Point-and-click approach:
    - Go to https://github.com/inbo/tutorials, press the clone button and copy the URL (default is HTTPS: `https://github.com/inbo/tutorials.git`)
    - Open RStudio: `File > New project > Version control > Git > paste https://github.com/inbo/tutorials.git > create project`
- Using git commands
    - Open a terminal
    - (Go to the folder where you want to clone the repository using `cd` command or open the terminal directly at that location)
    - type `git clone https://github.com/inbo/tutorials.git`
- Using R packages: choose one:
    - `usethis::create_from_github(repo_spec = "https://github.com/inbo/tutorials.git", destdir = "path/to/gitrepofolder")`
    - `gert::git_clone(url = "https://github.com/inbo/tutorials.git", path = "path/to/gitrepofolder")`

See also in [SSH protocol](#creating-public-private-ssh-key-pairs) the paragraph about the use of the `git remote` command. 

When you use these git commands or R functions, your PAT will be discovered from the Git Credential Manager and automatically authorize access (if you followed the steps in the previous section correctly) [^1].

[^1]: In case you did not follow the guidelines correctly, Git will prompt you to log in using a browser window and asks for your username and password.
In the password section you will need to provide your PAT (not your GitHub password, which is indeed confusing). If you enabled two-factor authentication (like you should have), you'll need to complete the 2FA challenge. Once you've authenticated successfully, your credentials are stored in the Git Credential Manager and will be used every time you clone an HTTPS URL. Git will not ask you for your credentials again unless you change your credentials.

## Creating public-private SSH-key pairs

Below we describe the steps you need to run in order to make SSH-keys (Secure SHell).
The SSH protocol is, next to the HTTPS-protocol, a way to safely interact with 
GitHub.
A general and easy to follow explanation about SSH can be read [here](https://dev.to/risafj/ssh-key-authentication-for-absolute-beginners-in-plain-english-2m3f).
Both HTTPS and SSH are secure ways to communicate with a server (pass information between your computer and a server).
For a discussion of technical differences and similarities on how both protocols handle security, [this resouce about SSH and TLS](https://www.ssl2buy.com/wiki/ssh-vs-ssl-tls) is useful.


A private key, that is specific to your computer, and a matching public key that
is stored on your GitHub account are needed.
The provided instructions are for Windows and are taken from
https://happygitwithr.com/ssh-keys.html.
The same source may be consulted for other operating systems.

1. Open RStudio: _Tools > Global Options…> Git/SVN > Create RSA Key..._ You can
optionally use a passphrase for extra protection of the key.
Without password, anyone who has a copy of your private key, can impersonate you when authenticating on GitHub.
When the private key is password protected they need your password too.
Click create and apply.

2. In RStudio, open a Rproject with Git version control and navigate to the Git
pane.
Open the Git Shell (_More > Shell_).
First we check if the SSH agent works with the following command
(The first $ is the prompt and you should not copy it. To paste in the shell
click the right mouse button.):

    ```
    $ eval $(ssh-agent -s)
    # which should give something like this:
    Agent pid 59566
    ```

    If that works, we need to add the SSH key (if you provided a passphrase, you
    will need it here.):

    ```
    $ ssh-add ~/.ssh/id_rsa
    ```

3. Restart RStudio. Next _Tools > Global Options…> Git/SVN_ and click
_View public key_ in the SSH-RSA section.
Copy to the clipboard.

4. Open your personal GitHub account _`https://github.com/<githubusername>`_
(you may need to login). 
Click your profile-icon in the upperright corner and go to
_Settings > SSH and GPG keys_.
Click _New SSH key_.
Paste the public key in the field and provide an informative title
(e.g. `<year>-<computername>`).
Click _Add SSH Key_.

5. To use the SSH protocol (or HTTPS for that matter), you need to set the remote of your repository correctly.
The following command is useful to find out which protocol your repository uses
(_Git pane > More > Shell_) (open the Shell from within the RStudio project
with Git version control which you want to switch to SSH):

    ```
    $ git remote -v
    ```
You will either see `https://github.com/<OWNER>/<REPO>.git` or
`git@github.com:<OWNER>/<REPO>.git`.
The first one is the HTTPS-protocol, the latter one the SSH-protocol.
Switching between protocols is possible at any time.
To switch from HTTPS-protocol to SSH-protocol, type:

    ```
    $ git remote set-url origin git@github.com:USERNAME/REPOSITORY.git
    ```

    Check if the remote is set correctly with `git remote -v`.
