--- 
title: "Classification and visualisation of estimates and their uncertainty"
author: "Thierry Onkelinx"
site: bookdown::bookdown_site
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: INBOmd::inbo_slides
    location: "2020-02-03"
    toc: FALSE
    slide_level: 2
    theme: inbo
    lang: english
    font_flanders: TRUE
    cover: photo-of-forest-with-fog-1671324.jpg #https://www.pexels.com/photo/photo-of-forest-with-fog-1671324
    cover_horizontal: FALSE
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  dev = "cairo_pdf",
  echo = FALSE
)
library(grid)
library(tidyverse)
library(scales)
library(gridExtra)
library(effectclass)
library(qrcode)
library(INBOtheme)
if (interactive()) {
  theme_set(
    theme_inbo(
      base_family = "Calibri",
      base_size = 12, 
      transparent = "plot"
    )
  )
} else {
  theme_set(
    theme_inbo(
      base_family = "Calibri",
      base_size = 10, 
      transparent = "plot"
    )
  )
  update_geom_defaults("text", list(size = 3))
}
```

# Introduction

## Interpretation of estimates without information of uncertainty

```{r example1}
example <- tibble(
  year = 2008:2019, 
  log_mean = c(
    0.0121739367571996, 0.0302943346128291, 0.0543294920787334, 
    0.0728212525993772, 0.0874232566229058, 0.0776393430910669, 
    0.0814711880023876, 0.0856695157160183, 0.0846196500557396, 
    0.100704508903153, 0.109449358322858, 0.0913570454765638
  ), 
  log_sd = c(
    0.0634628387844638, 0.0889342263613141, 0.111204490060904, 
    0.130038728018353, 0.139808205841085, 0.13830227064554, 
    0.148274205414618, 0.1488048815632, 0.145743583525741,
    0.160521118918554, 0.17050538360847, 0.170659007417255
  )
) %>%
  mutate(
    change = exp(log_mean),
    lcl = exp(qnorm(0.025, log_mean, log_sd)),
    ucl = exp(qnorm(0.975, log_mean, log_sd))
  )
p <- ggplot(example, aes(x = year, y = change)) +
  geom_line() +
  scale_x_continuous(breaks = pretty) +
  theme(axis.title.x = element_blank())
grid.arrange(
  p + scale_y_continuous("Change (2007 = 100%)", labels = percent),
  p + 
    scale_y_continuous(
      "Change (2007 = 100%)", labels = percent, limits = c(0.5, 3)
    ),
  nrow = 1
)
```

## Estimates with uncertainty

```{r example2}
p <- ggplot(example, aes(x = year, y = change, ymin = lcl, ymax = ucl)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  scale_x_continuous(breaks = pretty) +
  theme(axis.title.x = element_blank())
p + scale_y_continuous("Change (2007 = 100%)", labels = percent)
```

## Unapproriate limits in combination with uncertainty

```{r example3}
grid.arrange(
  p + 
    scale_y_continuous("Change (2007 = 100%)", labels = percent) +
    coord_cartesian(ylim = range(example$change)),
  p + 
    scale_y_continuous(
      "Change (2007 = 100%)", labels = percent, limits = c(0.5, 3)
    ),
  nrow = 1
)
```

# Uncertainty

## Point estimate

```{r point_estimate}
effect <- tibble(
  mean = 0,
  sd = 1,
  lcl95 = qnorm(0.025, mean, sd),
  ucl95 = qnorm(0.975, mean, sd)
)
tibble(
  p = seq(0, 1, by = 0.005)
) %>%
  mutate(
    effect = qnorm(p),
    density = dnorm(effect)
  ) %>%
  filter(is.finite(effect)) -> likelihood
plot_dens <- ggplot(likelihood, aes(x = effect, y = density)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
plot_effect <- ggplot(effect, aes(y = mean, x = 1)) +
  geom_point() +
  ylim(range(likelihood$effect)) +
  coord_flip() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
g_dens <- ggplotGrob(plot_dens)
g_effect <- ggplotGrob(plot_effect)
max_width <- as.list(unit.pmax(g_dens$widths[2:5], g_effect$widths[2:5]))
g_dens$widths[2:5] <- max_width
g_effect$widths[2:5] <- max_width
grid.arrange(
  grobs = list(g_effect, g_dens), 
  heights = c(1, 4)
)
```

## Confidence interval

```{r interval}
likelihood %>%
  filter(0.025 <= p, p <= 0.975) %>%
  bind_rows(
    likelihood %>%
      filter(pmin(abs(p - 0.025), abs(p - 0.975)) < 1e-4) %>%
      mutate(density = 0) %>%
      arrange(desc(p))
  ) -> likelihood_polygon
plot_dens <- ggplot(likelihood, aes(x = effect, y = density)) +
  geom_polygon(
    data = likelihood_polygon,
    fill = inbo.steun.blauw,
    alpha = 0.3
  ) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_vline(xintercept = c(effect$lcl95, effect$ucl95), linetype = 3) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
plot_effect <- ggplot(effect, aes(y = mean, x = 1, ymin = lcl95, ymax = ucl95)) +
  geom_point() +
  geom_errorbar() +
  ylim(range(likelihood$effect)) +
  coord_flip() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
g_dens <- ggplotGrob(plot_dens)
g_effect <- ggplotGrob(plot_effect)
g_dens$widths[2:5] <- max_width
g_effect$widths[2:5] <- max_width
grid.arrange(
  grobs = list(g_effect, g_dens), 
  heights = c(1, 4)
)
```

## Confidence interval

```{r interval2}
likelihood %>%
  filter(pmin(abs(p - 0.025), abs(p - 0.975)) < 1e-4) %>%
  mutate(density = c(0, diff(p) / diff(effect))) %>%
  pivot_wider(names_from = p, values_from = c(effect, density)) -> like_rect
plot_dens <- ggplot(likelihood, aes(x = effect, y = density)) +
  geom_rect(
    data = like_rect,
    aes(
      xmin = effect_0.025, ymin = density_0.025, 
      xmax = effect_0.975, ymax = density_0.975,
      x = NULL, y = NULL
    ),
    alpha = 0.3
  ) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_vline(xintercept = c(effect$lcl95, effect$ucl95), linetype = 3) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
plot_effect <- ggplot(effect, aes(y = mean, x = 1, ymin = lcl95, ymax = ucl95)) +
  geom_point() +
  geom_errorbar() +
  ylim(range(likelihood$effect)) +
  coord_flip() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
g_dens <- ggplotGrob(plot_dens)
g_effect <- ggplotGrob(plot_effect)
g_dens$widths[2:5] <- max_width
g_effect$widths[2:5] <- max_width
grid.arrange(
  grobs = list(g_effect, g_dens), 
  heights = c(1, 4)
)
```

## 30%, 60% and 90% confidence interval

```{r interval3}
tibble(
  p = c(0.05, 0.2, 0.35, 0.65, 0.8, 0.95)
) %>%
  mutate(
    end = qnorm(p, effect$mean, effect$sd),
    start = lag(end),
    interval = abs(1 - 2 * p) %>%
      lag(),
    width = c(0, diff(end)),
    delta_p = c(0, diff(p)),
    height = delta_p / width
  ) %>%
  filter(!is.na(height)) %>%
  transmute(start, end, bottom = 0, density = height) -> like_rect
plot_dens <- ggplot(likelihood, aes(x = effect, y = density)) +
  geom_rect(
    data = like_rect,
    aes(
      xmin = start, ymin = bottom, 
      xmax = end, ymax = density,
      x = NULL, y = NULL
    ),
    alpha = 0.2
  ) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_vline(
    xintercept = qnorm(
      c(0.05, 0.2, 0.35, 0.65, 0.8, 0.95), 
      effect$mean, 
      effect$sd
    ), 
    linetype = 3
  ) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
plot_effect <- ggplot(effect, aes(y = mean, x = 1, link_sd = sd)) +
  geom_point() +
  stat_fan(geom = "rect") +
  ylim(range(likelihood$effect)) +
  coord_flip() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
g_dens <- ggplotGrob(plot_dens)
g_effect <- ggplotGrob(plot_effect)
g_dens$widths[2:5] <- max_width
g_effect$widths[2:5] <- max_width
grid.arrange(
  grobs = list(g_effect, g_dens), 
  heights = c(1, 4)
)
```

## 9 confidence intervals (10% increments)

```{r interval9}
tibble(
  p = seq(0.05, 0.95, by = 0.05)
) %>%
  mutate(
    end = qnorm(p, effect$mean, effect$sd),
    start = lag(end),
    interval = abs(1 - 2 * p) %>%
      lag(),
    width = c(0, diff(end)),
    delta_p = c(0, diff(p)),
    height = delta_p / width
  ) %>%
  filter(!is.na(height)) %>%
  transmute(start, end, bottom = 0, density = height) -> like_rect
plot_dens <- ggplot(likelihood, aes(x = effect, y = density)) +
  geom_rect(
    data = like_rect,
    aes(
      xmin = start, ymin = bottom, 
      xmax = end, ymax = density,
      x = NULL, y = NULL
    ),
    alpha = 0.3
  ) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_vline(
    xintercept = qnorm(
      seq(0.05, 0.95, by = 0.05), 
      effect$mean, 
      effect$sd
    ), 
    linetype = 3
  ) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
plot_effect <- ggplot(effect, aes(y = mean, x = 1, link_sd = sd)) +
  geom_point() +
  stat_fan(geom = "rect", fine = TRUE) +
  ylim(range(likelihood$effect)) +
  coord_flip() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
g_dens <- ggplotGrob(plot_dens)
g_effect <- ggplotGrob(plot_effect)
g_dens$widths[2:5] <- max_width
g_effect$widths[2:5] <- max_width
grid.arrange(
  grobs = list(g_effect, g_dens), 
  heights = c(1, 4)
)
```

## Example

```{r example4}
grid.arrange(
  ggplot(example, aes(x = year, y = change, ymin = lcl, ymax = ucl)) +
    geom_ribbon(alpha = 0.2) +
    geom_line() +
    scale_x_continuous(breaks = pretty) +
    theme(axis.title.x = element_blank()) + 
    scale_y_continuous("Change (2007 = 100%)", labels = percent) +
    ggtitle("95% CI"),
  ggplot(example, aes(x = year, y = exp(log_mean), link_sd = log_sd)) +
    stat_fan(link = "log") +
    geom_line() +
    scale_x_continuous(breaks = pretty) +
    scale_y_continuous(
      "Change (2007 = 100%)", 
      labels = percent, 
      limits = c(min(example$lcl), max(example$ucl))
    ) +
    theme(axis.title.x = element_blank()) +
    ggtitle("30%, 60% & 90% CI"),
  nrow = 1
)
```

# Classification of estimates

## Significant effects

```{r reference}
ds <- data.frame(
  effect = factor(1:2, labels = c("positive\neffect", "negative\neffect")
  ),
  mean = c(2, -2),
  sd = 1
)
ggplot(ds, aes(x = effect, y = mean, link_sd = sd)) +
  stat_fan(geom = "rect") +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_y_continuous(
    breaks = 0, 
    labels = "reference"
  ) +
  theme(
    axis.title = element_blank()
  )
```

## Non significant effects

```{r reference2}
ds <- data.frame(
  effect = factor(1:2, labels = c("no\neffect", "unknown\neffect")
  ),
  mean = c(0.2, -0.5),
  sd = c(1, 2)
)
ggplot(ds, aes(x = effect, y = mean, link_sd = sd)) +
  stat_fan(geom = "rect") +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_y_continuous(
    breaks = 0, 
    labels = "reference"
  ) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_blank()
  )
```

## Non significant effects

```{r reference3}
ggplot(ds, aes(x = effect, y = mean, link_sd = sd)) +
  stat_fan(geom = "rect") +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-2, 2), linetype = 3) +
  scale_y_continuous(
    breaks = c(-2, 0, 2), 
    labels = c("lower\nthreshold", "reference", "upper\nthreshold")
  ) +
  theme(
    axis.title = element_blank()
  )
```

## Detailed classification of positive effects

```{r positive}
ds <- data.frame(
  effect = factor(
    1:3, 
    labels = c(
      "positive\neffect", "moderate\npositive\neffect", 
      "strong\npositive\neffect"
    )
  ),  
  mean = c(1.95, 1.05, 3.05),
  sd = c(1, 0.5, 0.5)
)
ggplot(ds, aes(x = effect, y = mean, link_sd = sd)) +
  stat_fan(geom = "rect") +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-2, 2), linetype = 3) +
  scale_y_continuous(
    breaks = c(-2, 0, 2), 
    labels = c("lower\nthreshold", "reference", "upper\nthreshold")
  ) +
  theme(
    axis.title = element_blank()
  )
```

## Detailed classification of unknown effects

```{r unknown}
ds <- data.frame(
  effect = factor(
    1:3, 
    labels = c(
      "potential positive\neffect", "potential\nnegative\neffect", 
      "unknown\neffect"
    )
  ),  
  mean = c(1, -1, -0.3),
  sd = c(1.5, 1.5, 2)
)
ggplot(ds, aes(x = effect, y = mean, link_sd = sd)) +
  stat_fan(geom = "rect") +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-2, 2), linetype = 3) +
  scale_y_continuous(
    breaks = c(-2, 0, 2), 
    labels = c("lower\nthreshold", "reference", "upper\nthreshold")
  ) +
  theme(
    axis.title = element_blank()
  )
```

## Why don't you use p-values?

```{r p-values}
data.frame(
  mean = c(1.2, 10.5, 0, 0),
  sd = c(0.3, 5, 0.3, 5)
) %>%
  mutate(
    effect = classification(qnorm(0.05, mean, sd), qnorm(0.95, mean, sd), 2),
    p = sprintf("p = %.4f", 1 - 2 * abs(0.5 - pnorm(mean, sd = sd))),
    p_y = -10
  ) %>%
  ggplot(aes(x = effect, y = mean, link_sd = sd, label = p)) +
  stat_fan(geom = "rect") +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-2, 2), linetype = 3) +
  geom_text(aes(y = p_y)) +
  scale_y_continuous(
    breaks = c(-2, 0, 2), 
    labels = c("lower\nthreshold", "reference", "upper\nthreshold")
  ) +
  theme(
    axis.title = element_blank()
  )
```

## Overview

```{r overview}
overview <- data.frame(
  thresholds = c(3, 2, 2, 1, 1, 1, 0, 0, 0, 0),
  mean = c(0, 1, -1, 2, 0, -2, 3, 1, -1, -3),
  sd = c(2, 1, 1, 1, 1, 1, 0.5, 0.5, 0.5, 0.5)
) %>%
  mutate(
    effect = classification(qnorm(0.05, mean, sd), qnorm(0.95, mean, sd), 2),
    coarse = coarse_classification(effect),
    thresholds = sprintf("interval contains %i thresholds", thresholds)
  )
ggplot(overview, aes(x = effect, y = mean, link_sd = sd)) +
  stat_fan(geom = "rect") +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-2, 2), linetype = 3) +
  facet_wrap(~thresholds, scales = "free_x") +
  scale_y_continuous(
    breaks = c(-2, 0, 2), 
    labels = c("lower\nthreshold", "reference", "upper\nthreshold")
  ) +
  theme(
    axis.title = element_blank()
  )
```

## Example

```{r example5}
grid.arrange(
  ggplot(example, aes(x = year, y = exp(log_mean), link_sd = log_sd)) +
    geom_hline(yintercept = 1, linetype = 2) +
    geom_hline(yintercept = c(0.75, 1.33), linetype = 3) +
    stat_fan(link = "log") +
    geom_line() +
    scale_x_continuous(breaks = pretty) +
    scale_y_continuous(
      "Change (2007 = 100%)", 
      breaks = pretty,
      labels = percent
    ) +
    theme(axis.title.x = element_blank()),
  ggplot(example, aes(x = year, y = exp(log_mean), link_sd = log_sd / 10)) +
    geom_hline(yintercept = 1, linetype = 2) +
    geom_hline(yintercept = c(0.75, 1.33), linetype = 3) +
    stat_fan(link = "log") +
    geom_line() +
    scale_x_continuous(breaks = pretty) +
    scale_y_continuous(
      "Change (2007 = 100%)", 
      breaks = pretty,
      labels = percent
    ) +
    theme(axis.title.x = element_blank()),
  nrow = 1
)
```

# How to define reference and thresholds?

## Width of the confidence interval

> "The value for which $p=0.05$, or 1 in 20, is 1.96 or nearly 2; it is convenient to take this point as a limit in judging whether a deviation is to be considered significant or not." [@fisher1925StatisticalMethodsResearch]

- two-sided test with $p = 0.05$ as threshold
    - $p=0.05/2=0.025$ on each side
    - 2.5% and 97.5% quantiles = 95% CI
- smaller CI
    - larger $p$
    - more false positives & fewer false negatives
    
|                 | no effect | effect |
| --------------- | :-------: | :----: |
| not significant | TN        | FN     |
| significant     | FP        | TP     |
    
- Change detection in ecology: fewer false negatives = early warning

## Reference

- Null hypothesis
- A model parameter is zero
    - No effect of treatment X
    - No change in time
    - e.g. Natura2000: population remains stable
- A prespecified target
    - e.g. PARTRIDGE: population increases by 30% over 4 years

## Threshold

- Effect size from power analysis
    - e.g. Natura 2000: detect -25% over 24 year
- Expert judgement
    - What change is large enough to be relevant / important?
    - What change is small enough to be irrelevant / not important?
    - Do consult a statistician

# Tools

## R package 

### effectclass

https://effectclass.netlify.com [@effectclass]

### functions

- `classification()`: classifies intervals based on a reference and thresholds
- `format_ci()`: format an effect size with its confidence interval
- `stat_fan()`: fan plot based on mean and standard error
- `stat_effect() + scale_effect()`: display point with shape based on classification

## `classification(reference = 3, threshold = c(2, 4))` and `format_ci()`

```{r format-ci}
ds <- expand.grid(
  estimate = pi * seq(0, 1.9, by = 0.5),
  se = c(1, 0.1, 0.01)
)
ds$lcl <- qnorm(0.05, ds$estimate, ds$se)
ds$ucl <- qnorm(0.95, ds$estimate, ds$se)
ds$class <- classification(
  ds$lcl, ds$ucl, reference = 3, threshold = c(2, 4)
)
ds <- ds[order(ds$class, ds$estimate, ds$ucl), ]
ds$class <- format(ds$class, type = "markdown")
ds$formatted <- format_ci(ds$estimate, lcl = ds$lcl, ucl = ds$ucl)
kable(ds, align = "lllcc", row.names = FALSE, format = "markdown")
```

## `stat_effect() + scale_effect()`

```{r stat-effect}
expand.grid(
  mean = seq(-3, 3, length = 21),
  sd = c(0.07, 0.1, 0.15, 0.2, 0.3, 0.4, 0.5, 0.7, 1, 1.5, 2, 3)
) %>%
  mutate(
    lcl = qnorm(0.05, mean, sd),
    ucl = qnorm(0.95, mean, sd)
  ) %>%
  ggplot(aes(x = mean, y = sd, colour = mean)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_vline(xintercept = c(-2, 2), linetype = 3) +
  stat_effect(aes(ymin = lcl, ymax = ucl), threshold = 2, size = 3) +
  scale_effect() +
  scale_x_continuous("effect size") +
  scale_y_continuous("uncertainty", breaks = c(0.1, 0.2, 0.5, 1, 2)) +
  coord_trans(y = "log10") +
  scale_colour_gradient2("effect size", guide = "none")
```

# Your feedback

## Questions for the audience

- Is the classification easy to understand and use?
- How do you feel about the fan-plots?
- Should we use a standardise symbology and terminology?

### effectclass

```{r qrcode, fig.width = 1.5, fig.height = 1.5}
par(mar = rep(0, 4), mai = rep(0, 4))
qrcode_gen(
  "https://effectclass.netlify.com", 
  bColor = inbo.hoofd, 
  plotQRcode = FALSE,
  dataOutput = TRUE
) %>%
image(
  asp = 1, axes = FALSE, col = c(NA, inbo.hoofd)
)
```

https://effectclass.netlify.com

## Suggested symbology

```{r symbols, results = "asis", eval = FALSE}
overview %>%
  arrange(effect) %>%
  transmute(
    "unsigned coarse" = effect %>%
      coarse_classification() %>%
      remove_sign() %>%
      format(type = "markdown"),
    coarse = effect %>%
      coarse_classification(),
    detailed = format(effect, type = "markdown"),
    "unsigned detailed" = effect %>%
      remove_sign() %>%
      format(type = "markdown"),
  ) %>%
  kable(escape = FALSE, format = "markdown", align = "c")
```

```{r overview2}
overview %>%
  mutate(effect = reorder(effect, -as.integer(effect))) %>%
  ggplot(aes(x = effect, y = mean, link_sd = sd)) +
    stat_fan(geom = "rect") +
    stat_effect(
      aes(ymin = qnorm(0.05, mean, sd), ymax = qnorm(0.95, mean, sd)),
      threshold = 2, 
      size = 5,
      colour = inbo.steun.donkerroos
    ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_hline(yintercept = c(-2, 2), linetype = 3) +
    scale_y_continuous(
      breaks = c(-2, 0, 2), 
      labels = c("lower\nthreshold", "reference", "upper\nthreshold")
    ) +
    scale_effect(guide = guide_legend(ncol = 1)) +
    theme(
      axis.title = element_blank()
    )
```

## Suggested terminology

Class | trend              | state                     |
:---: | ------------------ | ------------------------- |
`++`  | strong increase    | goal strongly exceeded    |
`+`   | increase           | goal exceed               |
`+~`  | moderate increase  | goal moderately exceeded  |
`~`   | stable             | goal reached              |
`-~`  | moderate decrease  | goal almost achieved      |
`-`   | decrease           | goal not achieved         |
`--`  | strong decrease    | goal clearly not achieved |
`?+`  | potential increase | goal maybe achieved       |
`?-`  | potential decrease | goal maybe not achieved   |
`?`   | unclear            | unknown if goal achieved  |

### effectclass

https://effectclass.netlify.com
