---
title: "Match scientific names with the GBIF backbone"
author: "Dirk Maes, Dimitri Brosens"
date: 2018-06-14
categories: ["r"]
tags: ["API", "webservice", "r", "gbif", "biodiversity"]
output: 
    md_document:
        preserve_yaml: true
        variant: markdown_github
---

## Introduction

This tutorial will explain how you can match a list of 'scientific names' to the GBIF taxonomic backbone

Important is that you have [rgbif](https://github.com/ropensci/rgbif) and [inborutils](https://inbo.github.io/inborutils/) installed and available:

```{r message=FALSE, warning=FALSE}
library(tidyverse) # tidyverse
library(rgbif)     # To Match GBIF
library(inborutils) # wrap GBIF api data
library(knitr)
```

## Read data file containing the scientific names

Read file containing the scientific names you want to check against the [GBIF](https://www.gbif.org/species/search?q=) taxonomic backbone:

```{r}
species_list <- read_csv("https://raw.githubusercontent.com/inbo/inbo-pyutils/master/gbif/gbif_name_match/sample.csv", trim_ws = TRUE, col_types = cols())
```
   
Take a look at the data:

```{r}
knitr::kable(head(species_list))
```

## Request species information

Match the column containing the scientificName with GBIF, using the `gbif_species_name_match` function from the [inborutils](https://inbo.github.io/inborutils/reference/gbif_species_name_match.html) package:

```{r, message=FALSE, warning=FALSE}
species_list_matched <- species_list %>% 
    gbif_species_name_match(name_col = "name") 
```

The `name_col` argument is the column name containing the scientific names. 

Take a look at the updated data:

```{r}
knitr::kable(head(species_list_matched))
```

You can also specify which info (`gbif_terms`) GBIF should return. For example, return a lot of terms: 

```{r, message=FALSE, warning=FALSE}
species_list %>% 
    gbif_species_name_match(name_col = "name",
                            gbif_terms = c('usageKey', 'scientificName', 
                                           'rank', 'order', 'matchType', 
                                           'phylum', 'kingdom', 'genus', 
                                           'class', 'confidence', 'synonym', 
                                           'status', 'family')) %>%
    head() %>%
    kable()
```

or rather just a subset:

```{r, message=FALSE, warning=FALSE}
species_list %>% 
    gbif_species_name_match(name_col = "name",
                            gbif_terms = c('scientificName', 'family',
                                           'order', 'matchType')) %>%
    head() %>%
    kable()
```

