---
title: "Tutorial on how to retrieve data from the INBOVEG database"
author: "Els De Bie, Hans Van Calster & Jo Loos"
date: 2020-02-11
categories: ["databases"]
tags: ["database", "queries"] 
output: 
  md_document:
    preserve_yaml: true
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

# Introduction
The Flemish vegetation database, INBOVEG, is an application developed to provide
a repository of relevés and makes the relevés available for future use.

INBOVEG supports different types of recordings: BioHab recordings (protocol of 
Natura 2000 monitoring) and the classic relevés. The classic relevés can stand 
alone, be an element of a collection or element of a chain where the linkage is 
used to give information about the relative position of recording within a series.
Ample selection and export functions toward analysis tools are provided. It also
provides standardized lists of species, habitats, life forms, scales etc. 
Original observations are preserved and a full history of subsequent 
identifications is saved.


# Aim
In this tutorial we make functions available to query data directly from the 
INBOVEG SQL-server database. This to avoid writing your own queries or 
to copy/paste them from the access-frontend for INBOVEG.

We have provided functions to query

* survey (INBOVEG-projects)

* recordings (vegetation relevés)

* metadata of recordings (header info)

* classification (Natura2000 or local classification like BWK)

* qualifiers (management and site characteristics)
  
# Packages and connection
In order to run the functionalities, some R packags need to be installed. 

The following packages are needed to run this code:

* glue

* DBI

* assertthat

* dplyr

The main functions that we will use in this tutorial all start with `inboveg_*`. These functions are made available by loading the `inborutils` package.


You can install inborutils from github with:
install.packages("devtools")
devtools::install_github("inbo/inborutils")

This tutorial will only work for people with access to the INBO network. As an INBO employee, you should make sure you have reading-rights for CYDONIA, 
otherwise place an ICT-call (ict.helpdesk@inbo.be)


```{r load_libraries, message = FALSE, warning = FALSE}
library(glue)
library(DBI)
library(assertthat)
library(dplyr)
library(inborutils)
```

The following R-code can be used to establish a connection to INBOVEG 
by means of a connection string:

<!--better to use a connection string than dsn. 
dsn requires extra steps and settings in windows odbc manager-->

```{r connection_str}
connection <- dbConnect(odbc::odbc(),
.connection_string = "Driver=SQL Server;Server=inbo-sql07-prd.inbo.be,1433;
Database=D0010_00_Cydonia;Trusted_Connection=Yes;")
```

Or using dbconnection of the inborutils-package with the database 'Cydonia' 
on the inbo-sql07-prd server:

```{r connection_inbo}
con <- connect_inbo_dbase("D0010_00_Cydonia")
```

# Functionality

## Survey information

The function `inboveg_survey` queries the INBOVEG database for survey information 
(metadata about surveys) for one or more survey(s) by the name of the survey.

###Examples

Three examples are given, this can be used as base to continue selecting the wanted data


Get information of a specific survey and collect data.

```{r examples_survey}
survey_info <- inboveg_survey(con, 
                              survey_name = "OudeLanden_1979", 
                              collect = TRUE)
```

```{r}
survey_info
```


Get information of all surveys. This time we will not use `collect = TRUE`, which will return a [lazy query](https://docs.lucee.org/guides/cookbooks/lazy_queries.html):

```{r}
allsurveys <- inboveg_survey(con)
```

```{r}
allsurveys
```


If only a part of the survey name is known?

```{r}
partsurveys <- inboveg_survey(con, 
                              survey_name = "%MILKLIM%",
                              collect = TRUE)
```

```{r}
head(partsurveys, 10)
```


## Recording information

The function  `inboveg_recordings` queries the INBOVEG database for relevé 
information (which species were recorded in which plots and in which vegetation 
layers with which cover) for one or more surveys.

### Examples

Four examples are given, this can be used as base to continue selecting the wanted data.


Get the relevés from one survey and collect the data

```{r examples_recordings}
recording_heischraal2012 <- inboveg_recordings(con, 
                                      survey_name = "MILKLIM_Heischraal2012",
                                      collect = TRUE)
```

```{r}
head(recording_heischraal2012, 10)
```


Get all recordings from MILKLIM surveys (partial matching), don't collect

```{r}
recording_milkim <- inboveg_recordings(con,
                              survey_name = "%MILKLIM%")
```

```{r}
recording_milkim
```


Get recordings from several specific surveys

```{r}
recording_severalsurveys <- inboveg_recordings(con,
                survey_name = c("MILKLIM_Heischraal2012", "NICHE Vlaanderen"),
                multiple = TRUE,
                collect = TRUE)
```

```{r}
head(recording_severalsurveys, 10)
```


Get all relevés of all surveys,  don't collect the data

```{r}
allrecordings <- inboveg_recordings(con)
```

```{r}
allrecordings
```


## Header information

This function `inboveg_header`queries the INBOVEG database for header information 
(metadata for a vegetation-relevé) for one survey by the name of the survey 
and the recorder type. 

###Examples

Two examples are given, this can be used as base to continue selecting the wanted data


Get header information from a specific survey and a specific recording type and collect the data

```{r examples_header}
header_info <- inboveg_header(con, 
                              survey_name = "OudeLanden_1979",
                              rec_type = "Classic",
                              collect = TRUE)
```

```{r}
header_info
```


Get header information of all surveys,  don't collect the data

```{r}
all_header_info <- inboveg_header(con)
```

```{r}
all_header_info
```


## Classification information

The function `inboveg_classification` queries the INBOVEG database for information 
on the field classification (N2000 or BWK-code) of the relevé for one or more 
survey(s) by the name of the survey. 

###Examples

Two  examples are given, this can be used as base to continue selecting the wanted data


Get a specific classification from a survey and collect the data

```{r examples_classification}
classif_info <- inboveg_classification(con, 
                              survey_name = "MILKLIM_Heischraal2012",
                              classif = "4010",
                              collect = TRUE)
```

```{r}
head(classif_info, 10)
```


Get all surveys, all classifications,  don't collect the data

```{r}
allecodes <- inboveg_classification(con)
```

```{r}
allecodes
```



## Qualifiers information

This function `inboveg_qualifiers`queries the INBOVEG database for
qualifier information on recordings for one or more surveys. These qualifiers 
give information on management, location description, ... 

###Examples

Four examples are given, this can be used as base to continue selecting the wanted data.


Get the qualifiers from one survey

```{r examples_qualifiers}
qualifiers_heischraal2012 <- inboveg_qualifiers(con,
                                  survey_name = "MILKLIM_Heischraal2012")
```

```{r}
head(qualifiers_heischraal2012, 10)
```


Get all qualifiers from MILKLIM surveys (partial matching):
```{r}
qualifiers_milkim <- inboveg_qualifiers(con,
                                        survey_name = "%MILKLIM%")
```

```{r}
head(qualifiers_milkim, 10)
```

Get qualifiers from several specific surveys
```{r}
qualifiers_severalsurveys <- inboveg_qualifiers(con, 
                survey_name = c("MILKLIM_Heischraal2012", "NICHE Vlaanderen"),
                multiple = TRUE)
```

```{r}
head(qualifiers_severalsurveys, 10)
```


Get all qualifiers of all surveys

```{r}
allqualifiers <- inboveg_qualifiers(con)
```

```{r}
head(allqualifiers,10)
```




## More complex queries

These functions gives the basis information out of INBOVEG. If more precise information is needed 'dplyr' is the magic word.

### Examples

Still in progress


# Closing the connection 
Close the connection when done

```{r closing_conn}
dbDisconnect(con)
rm(con)
```
