---
title: "Reading large data files in R"
author: "Bart Aelterman, Stijn Van Hoey"
date: "February 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

R is known to have difficulties handling large data files. Here we will explore some tips that make working with such files in R less painfull.

## tl;dr

* If you can comfortably work with the entire file in memory, but reading the file is rather slow, consider using the `data.table` package and [read the file with its `fread` function](#loading-a-large-dataset-use-fread-instead-of-read).
* If your file does not comfortably fit in memory:
    * [Use `sqldf`](#3-limiting-both-the-number-of-rows-and-the-number-of-columns-using-sqldf) if you have to stick to `csv` files.
    * Use a SQLite database and query it using either [SQL queries](#working-with-sqlite-databases) or [`dplyr`](#interacting-with-sqlite-databases-using-dplyr).

## Downloading the example files

While you can directly test this tutorial on your own large data files, we will use bird tracking data from the [LifeWatch bird tracking network](https://doi.org/10.3897/zookeys.555.6173) for the examples. We have made two versions of some tracking data available for download: a `.csv` file (text data) and a `.db` file (sqlite data). Both contain processed log files; for more information on the processing, see the [BirdTrackingEtl package](https://github.com/inbo/bird-tracking-etl).

```{r filenaming}
csv.name <- "2016-04-20-processed-logs-big-file-example.csv"
db.name <- "2016-04-20-processed-logs-big-file-example.db"
```

The evaluation of the next code chunk is ignored by default as the downloading and unzipping of the files results in more than 3 GB of data. If you do want to download the files yourself and test the other chunks, run the code and download the `csv` and `sqlite` examples. Make sure you have the `R.utils` package available (for unzipping the downloaded files). If not, use the command `install.packages("R.utils")` in your R console to download the package.

```{r datadownload, eval=FALSE, warning=FALSE, message=FALSE}
library("R.utils")
# download the CSV file example
csv.url <- paste("https://s3-eu-west-1.amazonaws.com/lw-birdtracking-data/", 
                 csv.name, ".gz", sep = "")
if(!file.exists(csv.name)){
    download.file(csv.url, destfile=paste(csv.name, ".gz", sep = ""))
    gunzip(paste(csv.name, ".gz", sep = ""))
}

# download the sqlite database example
db.url <- paste("https://s3-eu-west-1.amazonaws.com/lw-birdtracking-data/", 
                db.name, ".gz", sep = "")
if(!file.exists(db.name)){
    download.file(db.url, destfile=paste(db.name, ".gz", sep = ""))
    gunzip(paste(db.name, ".gz", sep = ""))
}
```

## Loading a large dataset: use `fread` instead of `read`.

```{r loaddatatable, warning=FALSE, message=FALSE}
library("data.table")
```

If you really need to read an entire csv in memory, by default, R users use the `read.table` method or variations thereof (such as `read.csv`). However, `fread` from the `data.table` package is a lot faster. Let's measure the time to read in the data using these two different methods.


```{r comp_timing, warning=FALSE, message=FALSE, cache=TRUE}
read.table.timing <- system.time(read.table(csv.name, header = TRUE, sep = ","))
data.table.timing <- system.time(allData <- fread(csv.name, showProgress = FALSE))
data <- data.frame(method = c('read.table', 'fread'), 
                  timing = c(read.table.timing[3], data.table.timing[3]))
data
```

`fread` is indeed much faster then the default `read.table`. However, the result of `fread` is a `data.table`, not a `data.frame`. The `data.table` package describes the `data.table` object as a more performant replacement for the `data.frame`. This means that selecting, filtering and aggregating data is much faster on a `data.table` compared to the standard `data.frame` but it requires you to use a slightly different syntax.

You can find the `data.table` package on [CRAN](https://cran.r-project.org/web/packages/data.table/index.html). A good place to learn this package are the package vignettes. The [introduction to data.table](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro-vignette.html) should be enough to get started.

## Data files that don't fit in memory

If you are not able to read in the data file, because it does not fit in memory (or because R becomes too slow when you load the entire dataset), you will need to limit the amount of data that will actually be stored in memory. There are a couple of options which we will investigate:

1. limit the number of lines you are trying to read for some exploratory analysis. Once you are happy with the analysis you want to run on the entire dataset, move to another machine.
1. limit the number of columns you are reading to reduce the memory required to store the data.
1. limit both the number of rows and the number of columns using `sqldf`.
1. stream the data.

### 1. Limit the number of lines you read (`fread`)

Limiting the number of lines you read is easy. Just use the `nrows` and/or `skip` option (available to both `read.table` and `fread`). `skip` can be used to skip a number of rows, but you can also pass a string to this parameter causing `fread` to only start reading lines from the first line matching that string. Let's say we only want to start reading lines after we find a line matching the pattern `801,2014-06-29`. We can do that like this:


```{r limitlines}
sprintf("Number of lines in full data set: %s", nrow(allData))
subSet <- fread(csv.name, skip = "2015-06-12 15:14:39", showProgress = FALSE)
sprintf("Number of lines in data set with skipped lines: %s", nrow(subSet))
```

Skipping rows this way is obviously not giving you the entire dataset, so this strategy is only useful for doing exploratory analysis on a subset of your data. If you want to explore the whole dataset, limiting the number of columns you read can be a more useful strategy.

### 2. Limit the number of columns you read (`fread`)

If you only need 4 columns of the 21 columns present in the file, you can tell `fread` to only select those 4. This can have a major impact on the memory footprint of your data. The option you need for this is: `select`. With this, you can specify a number of columns to keep. The opposite - specifying the columns you want to drop - can be accomplished with the `drop` option.

```{r limitcols}
fourColumns = fread(csv.name, select = c("device_info_serial", "date_time", 
                                         "latitude", "longitude"), 
                    showProgress = FALSE)
sprintf("Size of total data in memory: %s MB", utils::object.size(allData)/1000000)
sprintf("Size of only four columns in memory: %s MB", utils::object.size(fourColumns)/1000000)
```

The difference might not be as large as you would expect. R objects claim more memory then needed to store the data alone, as they keep pointers, and other object attributes. But still, the difference could save you.

### 3. Limiting both the number of rows and the number of columns using `sqldf`

The [sqldf package](https://cran.r-project.org/web/packages/sqldf/sqldf.pdf) allows you to run SQL-like queries on a file, resulting in only a selection of the file being read. It allows you to limit both the number of lines and the number of rows at the same time. In the background, this actually creates a sqlite database on the fly to execute the query. Consider using the package when starting from a csv file, but the actual strategy boils down to making a sqlite database file of your data. See [this section below](#the-database-file-strategy) to learn how to interact with those. 

### 4. Streaming data

**Short:** streaming a file in R is a bad idea. If you are interested why, read the rest of this section.

Streaming a file means reading it line by line and only keeping the lines you need or do stuff with the lines while you read through the file. It turns out that R is really not very efficient in streaming files. The main reason is the memory allocation process that has difficulties with a constantly growing object (which can be a dataframe containing only the selected lines).

In the next code block, we will read parts of our data file once using the `fread `function, and once line by line. You'll see the performance issue with the streaming solution.

```{r streamexample, warning=FALSE, message=FALSE}
library(ggplot2)
allowedDevices = c(753, 801, 852)
minDate = strptime('1/3/2014',format = '%d/%m/%Y')
maxDate = strptime('1/10/2014',format = '%d/%m/%Y')

streamFile <- function(limit) {
    con <- file(csv.name, open = "r")
    selectedRecords <- list()
    i <- 0
    file.streaming.timing <- system.time(
        while (i < limit) {
            oneLine <- readLines(con, n = 1, warn = FALSE)
            vec = (strsplit(oneLine, ","))
            selectedRecords <- c(selectedRecords, vec)
            i <- i + 1
        }
    )
    close(con)
    return(file.streaming.timing[[3]])
}

freadFile <- function(limit) {
    file.fread.timing = system.time(
        d <- fread(csv.name, showProgress = FALSE, nrows = limit)
    )
    return(file.fread.timing[[3]])
}

maxLines <- c(5000, 10000, 15000, 20000, 25000, 30000)
streamingTimes <- sapply(maxLines, streamFile)
freadTimes <- sapply(maxLines, freadFile)
data <- data.frame(n = maxLines, streaming = streamingTimes, 
                   fread = freadTimes)
pdata <- melt(data, id = c("n"))
colnames(pdata) <- c("n", "algorithm", "execTime")
qplot(n, execTime, data = pdata, color = algorithm, 
      xlab = "number of lines read", ylab = "execution time (s)")
```

## The database file strategy

### Working with SQLite databases

SQLite databases are single file databases meaning you can simply download them, store them in a folder or share them with colleages. Similar to a csv. They are however more powerful than csv's because of two important features:

* Support for SQL: this allows you to execute intelligent filters on your data, similar to the `sqldf` package or database environments you are familiar with. That way, you can reduce the amount of data that's stored in memory by filtering out rows or columns.
* Indexes: SQLite databases contain indexes. An index is something like an ordered version of a column. When enabled on a column, you can search through the column much faster. We will demonstrate this below.

We have downloaded a second file `2016-04-20-processed-logs-big-file-example.db` that contains the same data as the `2016-04-20-processed-logs-big-file-example.csv` file, but as a sqlite database. Furthermore, the database file contains indexes which will dramatically drop the time needed to perform search queries. If you do not have a SQLite database containing your data, try creating one as described [here](https://cran.r-project.org/web/packages/dplyr/vignettes/databases.html).

Let's first connect to the database and list the available tables.

```{r sqliteconnect}
library(RSQLite)
db <- dbConnect(SQLite(), dbname = db.name)

# show the tables in this database
dbListTables(db)
```

Let's try to select rows where the device id matches a given value (e.g. 860), and the date time is between two given timestamps. For our analysis, we only need `date_time`, `latitude`, `longitude` and `altitude` so we will only select those.

```{r}
sqlTiming <- system.time(data <-  dbGetQuery(conn = db,
   "SELECT date_time, latitude, longitude, altitude
    FROM processed_logs 
    WHERE device_info_serial = 860 
        AND date_time < '2014-07-01' 
        AND date_time > '2014-03-01'"
))

print(sqlTiming[3])
```

This provides a convenient and fast way to request subsets of data from our large data file. We could do the same analysis for each of the serial numbers, each time only loading that subset of the data. As an example, consider the calculation of the average altitude over the specified period for each of the bird serial identifiers in the list `serial_id_list`. By using a for loop, the calculation is done for each of the birds separately and the amount of data loaded into memory at the same time is lower:
    
```{r forloop, message=FALSE, warning=FALSE}
serial_id_list <- c(853, 860, 783)
print("Average altitude between 2014-03-01 and 2014-07-01:")
for (serialid in serial_id_list) {
    data <-  dbGetQuery(conn = db,
                       sprintf("SELECT date_time, latitude, longitude, altitude
                                FROM processed_logs 
                                WHERE device_info_serial = %d 
                                    AND date_time < '2014-07-01' 
                                    AND date_time > '2014-03-01'", serialid))
    print(sprintf("serialID %d: %f", serialid, mean(data$altitude)))
    }
```

Remark that we use the `sprintf` function to dynamically replace the serial id in the sqlite query we will execute. For each loop, the `%d` is replaced by the value of the serial id of the respective loop. Read the [manual](https://stat.ethz.ch/R-manual/R-devel/library/base/html/sprintf.html) of the `sprintf` function for more information and options.

### Interacting with SQLite databases using `dplyr`

If you're not comfortable with writing queries in SQL, R has a great alternative: `dplyr`. Dplyr can connect to a SQLite database and you can perform the same operations on it that you would do on a dataframe. However, dplyr will translate your commands to SQL, allowing you to take advantage of the indexes in the SQLite database.

```{r dplyr, message=FALSE, warning=FALSE}
library(dplyr)
my_db <- src_sqlite(db.name, create = FALSE)
bird_tracking <- tbl(my_db, "processed_logs")
results <- bird_tracking %>%
    filter(device_info_serial == 860) %>%
    select(date_time, latitude, longitude, altitude) %>%
    filter(date_time < "2014-07-01") %>%
    filter(date_time > "2014-03-01")
head(results)
```

Dplyr provides the ability to perform queries as above without the need to know SQL. If you want to learn more about how to use `dplyr` with a SQLite database, head over to [this vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/databases.html).
