---
title: "How to make spatial joins (point in polygon)?"
description: "Various ways to make a spatial join between GIS point data and polygon data"
author: "Stijn Van Hoey, Hans Van Calster"
date: 2019-06-20
categories: ["r", "gis"]
tags: ["gis", "r", "maps"]
output: 
    md_document:
        preserve_yaml: true
        variant: markdown_github
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE
)
```


```{r}
library(R.utils)
library(rgdal)
library(tidyverse)
library(leaflet)
library(sp)
library(sf)
library(rgbif)
library(DBI)
```

## What we want to do

In this short tutorial, we explore various options to deal with the situation where we have (1) a spatially referenced GIS file with polygons and (2) a spatially referenced set of points that overlaps with the GIS polygons. 

Typically, both data sources contain information (apart from the spatial locations) that needs to be related to each other in some way. In this case study, we want to know for each point in which polygon it is located.

## Get some data to work with

For the point data, we will work with data on the invasive species - Chinese mitten crab (**Eriocheir sinensis**) in  Flanders, Belgium, from the year 2008 (GBIF.org (20th June 2019) GBIF Occurrence Download https://doi.org/10.15468/dl.decefb).

We will use convenience functions from the `rgbif` package to download the data as a zip file and to import the data as a `tibble` in the R environment. 

```{r}
invasive_species <- occ_download_get("0032582-190415153152247",
                                     path = tempdir(),
                                     overwrite = TRUE) %>% 
  occ_download_import() %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))
```

We will use the European reference grid system from the European Environmental Agency as an example of a GIS vector layer (each grid cell is a polygon). The Belgian part of the grid system can be downloaded as a sqlite/spatialite database from the EEA website using the following code:

```{r}
# explicitly set mode = "wb", otherwise zip file will be corrupt
download.file("https://www.eea.europa.eu/data-and-maps/data/eea-reference-grids-2/gis-files/belgium-spatialite/at_download/file", destfile = file.path(tempdir(), "Belgium_spatialite.zip"), mode = "wb")

# this will extract a file Belgium.sqlite to the temporary folder
unzip(zipfile = file.path(tempdir(), "Belgium_spatialite.zip"), exdir = tempdir())  
```


## Point in polygon with the `sf` package

The spatial query can be done with the aid of the [`sf` package](https://r-spatial.github.io/sf/). The package has built-in functions to read spatial data (which uses GDAL as backbone). 

We will project the data to Belgian Lambert 72 (<https://epsg.io/31370>), because the join assumes planar coordinates.

```{r}
be10grid <- read_sf(file.path(tempdir(), "Belgium.sqlite"), 
                    layer = "be_10km")  %>%
  # convert to Belgian Lambert 72
  st_transform(crs = 31370)
```

We now get a `sf` object which is also a `data.frame` and a `tbl`:

```{r}
class(be10grid)
```

Let's have a look at this object:

```{r paged.print = FALSE}
be10grid
```

We can see that the spatial information resides in a `GEOMETRY` list column.

Similarly, the package has built-in functions to convert a `data.frame` containing coordinates to a spatial `sf` object:

```{r}
invasive_spatial <- st_as_sf(invasive_species,
                             coords = c("decimalLongitude",
                                        "decimalLatitude"),
                             crs = 4326) %>%
  # convert to Lambert72
  st_transform(crs = 31370)
```

Resulting in:

```{r paged.print = FALSE}
invasive_spatial
```


Now we are ready to make the spatial overlay. This is done with the aid of `sf::st_join`. The default join type is `st_intersects`. This will result in the same spatial overlay as `sp::over` (see next section). We join the information from the grid to the points through a left join. See [the DE-9IM topological model](https://en.wikipedia.org/wiki/DE-9IM) for explanations about all possible spatial joins. 

Note that with `st_intersects` points on a polygon boundary and points corresponding to a polygon vertex are considered to be inside the polygon. In the case where points are joined with polygons, `st_intersects` and `st_covered_by` will give the same result. The join type `st_within` can be used if we want to join only when points are completely within (excluding polygon boundary).


```{r}
invasive_be10grid_sf <- invasive_spatial %>%
  st_join(be10grid,
          join = st_intersects, 
          left = TRUE) %>%
  select(-eoforigin, -noforigin)  # exclude the EofOrigin NofOrigin columns
```

Looking at the resulting object:

```{r paged.print = FALSE}
invasive_be10grid_sf
```

## Point in polygon with the `sp` package

Instead of `sf` objects (= `data.frames` or `tibbles` with a geometry list-column), the `sp` package works with `Spatial` spatial data classes (which has many derived spatial data classes for points, polygons, ...).

First, we need to convert the `data.frame` with point locations to a `SpatialPointsDataFrame`. We also need to ensure that the coordinate reference system (CRS) for both the point locations and the grid is the same. The data from GBIF are in WGS84 format.

```{r}
crs_wgs84 <- CRS("+init=epsg:4326")
coord <- invasive_species %>%
    select(decimalLongitude, decimalLatitude)
invasive_spatial <- SpatialPointsDataFrame(coord,
                                           data = invasive_species,
                                           proj4string = crs_wgs84)
```

The `sp` package has no native methods to read the Belgium 10 km x 10 km grid, but we can use `rgdal::readOGR` to connect with the sqlite/spatialite database and extract the Belgium 10 km x 10 km grid as a `SpatialPolygonsDataFrame`. Apart from the 10 km x 10 km grid, the database also contains 1 km x 1 km and 100 km x 100 km grids as raster or vector files.

```{r}
be10grid <- readOGR(dsn = file.path(tempdir(), "Belgium.sqlite"), 
                  layer = "be_10km")
```


We transform the 10 km x 10 km grid to the same CRS system:

```{r}
be10grid <- spTransform(be10grid, crs_wgs84)
```

Now we are ready to spatially join (overlay) the `SpatialPointsDataFrame` wth the 10 km x 10 km grid. This can be done using `sp::over`. The first two arguments of the function give, respectively, the geometry (locations) of the queries, and the layer from which the geometries or attributes are queried. See `?sp::over`. In this case, when x = "SpatialPoints" and  y = "SpatialPolygonsDataFrame", it
returns a `data.frame` of the second argument with row entries corresponding to the first argument.

```{r}
invasive_be10grid <- over(x = invasive_spatial, y = be10grid)
invasive_species_be10grid <- bind_cols(invasive_species,
                                       invasive_be10grid)
```

To see what the result looks like, we can select the most relevant variables and print it (first ten rows).

```{r}
invasive_species_be10grid %>%
  select(species, starts_with("decimal"),
         eventDate, cellcode) %>%
  head(10)
```


## What have we done?

To wrap this up, we make a map that shows what we have done. We will use the results obtained with the `sf` package. 

First, we need to transform `invasive_be10grid_sf` back to WGS84 (the background maps in leaflet are in WGS84) (`be10grid` is already in WGS84 format).

```{r}
invasive_be10grid_sf <- invasive_be10grid_sf %>%
  st_transform(crs = 4326)
```


Zooming in on the point markers and hovering over a marker will show the reference grid identifier for the grid cell as it is joined to spatial points object `invasive_be10grid`. Clicking in a grid cell will bring up a popup showing the reference grid identifier for the grid cell as it is named in `be10grid`.

```{r}
leaflet(be10grid) %>%
  addTiles() %>%
  addPolygons(popup = ~cellcode) %>%
  addMarkers(data = invasive_be10grid_sf,
             label = ~cellcode)
```







